---
title: "p1-01 · Characterise lysine-rich domains"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Characterise lysine-rich domains across the human proteome — maximum
K-score vs protein length, subcellular localisation, functional-class enrichment,
and K-scores in histones.

**Inputs:** `data/processed_data_from_PNAS2022/all_protein_feature_per_position.csv`, the cd-code annotation table under `data/public_data/`, and the UniProt reference proteome;
annotation packages `org.Hs.eg.db` and `subcellularvis` (loaded where used).

**Outputs:** figures (rendered on knit).

**Upstream:** none. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Import data

Loads the per-position protein feature table, maps UniProt IDs to gene symbols, and prepares the K-score data used throughout.

```{r import_data, warning=FALSE}

## Define paths to directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p1.results.dir <- file.path(results.dir, "p1_characterise_domains")
create.dirs(c(results.dir, p1.results.dir))

## Load data into environment 
protein.feature.dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/all_protein_feature_per_position.csv"))

## Load library  
library("org.Hs.eg.db")

## Convert gene id and names
# Map UniProt IDs to gene symbols and ENSEMBL IDs
gene.id.dt <- select(
    org.Hs.eg.db,
    keys = protein.feature.dt[, unique(uniprot_id)],
    columns = c("SYMBOL","ENSEMBL"),
    keytype = "UNIPROT"
) %>%
    data.table

# Rename columns 
setnames(
    gene.id.dt,
    old = c("UNIPROT", "SYMBOL","ENSEMBL"),
    new = c("uniprot_id", "gene_name", "gene_id")
)

## Clean and organize gene mapping table
# Sort gene_id in ascending order, remove duplicate UniProt id and remove rows missing gene_id 
gene.id.dt <- gene.id.dt[order(gene_id)][!duplicated(uniprot_id)][!is.na(gene_id)]

# Rename data file and save to directory 
fwrite(gene.id.dt, file.path(p1.results.dir,"uniprot-ensembl.csv"))
```

# Maximum K score and protein length

Computes each protein's maximum K score (lysine richness) and its length.

```{r Maximum_K_score_Protein_Length}
#-----------------------------------
# Maximum K score and Protein Length
#-----------------------------------

## Compute maximum K_ratio and protein length for each protein
max.k.score.dt <- protein.feature.dt[ 
  , list(
    max_k_ratio = max(K_ratio), 
    protein_len = max(position)
    ), 
  by = list(Accession, uniprot_id)] 

## Merge gene annotations with protein summary data (max K_ratio, protein length)
# Merge tables using the UniProt ID as the key
max.k.score.dt <- merge(
    gene.id.dt,
    max.k.score.dt,
    all.y = TRUE,
    by = "uniprot_id"
)

# Create dataset with max k score 
fwrite(max.k.score.dt, file.path(p1.results.dir, "max-k-score-per-protein.csv")) 

print("The ratio of proteins with the max K ratio > 0.3")
nrow(max.k.score.dt[max_k_ratio > 0.3]) / nrow(max.k.score.dt)


# Median K score of all region in human proteome#
protein.feature.dt[, median(K_ratio)]

```

# K score distributions

Visualises the distribution of K scores and the relationship between maximum K score and protein length.

```{r Data_Visualization_Max_K_Score_&_Protein_Length}

#---------------------------------------------------------
# Data Visualization - K score, Maximum K score and Protein Length
#---------------------------------------------------------

## K_ratio distribution 
# Count how many proteins have each K_ratio value
protein.feature.dt[, table(K_ratio)] 

# Create a new column `K_ratio_group` based on the K_ratio
protein.feature.dt[, K_ratio_group := case_when(
  K_ratio >= 0.9 ~ 0.9, #If K_ratio is 0.9 or higher, assign value 0.9
  TRUE ~ K_ratio #If K_ratio is lower than 0.9, then assign original K_ratio value
)]

# Plot - The number of proteins per K score
ggplot(
    protein.feature.dt,
    aes(
        x = K_ratio_group
    )
) +
    geom_bar() +
    scale_x_continuous(breaks=seq(0, 1, 0.1))


# Plot - The number of proteins per max K score
ggplot(
    max.k.score.dt,
    aes(
        x = max_k_ratio
    )
) +
    geom_bar() +
    scale_x_continuous(breaks=seq(0, 1, 0.1)) + 
  labs(x = "max_k_ratio", y = "number of protein_count")

#Median - 24112025
median(max.k.score.dt[, max_k_ratio])
# 0.3

# Plot - Max K ratio vs protein length
ggplot(
  max.k.score.dt,
  aes(
    y = protein_len,
    x = factor(max_k_ratio)
  )
) + geom_boxplot(fill = "steelblue", outlier.shape = NA) +
  labs(x = "max_K_ratio", y = "protein_length") +
  coord_cartesian(ylim = c(0, 2000))

# Plot - K_score of disordered regions of proteins 
# This plot indicates K score depending on the level of protein disorderliness

ggplot(
  protein.feature.dt,
  aes(
    x = factor(K_ratio_group),
    y = IUPRED2
  )
) + 
  geom_boxplot(fill = "steelblue", outlier.shape = NA)


# Plot - Protein charge versus K_ratio_group
ggplot(
  protein.feature.dt,
  aes(
    x = factor(K_ratio_group),
    y = windowCharge
  )
) + 
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(fill = "steelblue", outlier.shape = NA)

```

# Subcellular localisation

Annotates proteins with subcellular localisation to relate lysine richness to cellular compartment.

```{r subcellular_localisation_analysis}

#--------------------------------------
# Sub-cellular localisation analysis
#--------------------------------------

# Packaged installed 20th Oct 2025
# install.packages(file.path("/fast/AG_Sugimoto/home/users/pallavi/Software_packages/R/subcellularvis"), repos = NULL, type = "source")

## Load library
library("subcellularvis")

## Run sub-cellular compartment analysis with enrichment score according Gene Ontology
# The 'extractCompartment' function runs sub-cellular compartment analysis, retrieves 
extractCompartment <- function(gene.set, class.name){
  comp.out <- compartmentData(  # Run compartment enrichment analysis using the compartmentData() function
    gene.set,
    id_type = "UNIPROT",
    aspect = c("Whole cell"),
    organism = c("Human"),
    annotationSource =c("Gene Ontology") # Alternatively, Human Protein Atlas can also be used
  )
  
  # Convert enrichment results from comp.out into a data.table 
  comp.out.dt <- data.table(comp.out$enrichment)
  
  # Add metadata columns 
  comp.out.dt[, `:=`(
    Genes = NA,
    n_unmapped = length(str_split(comp.out$unmapped, ",")[[1]]), # Counting the number of unmapped IDs
    n_mapped = comp.out$nMapped, # adding mapped IDs
    class_name = class.name 
  )]
  
  #  Updated data table 
  return(comp.out.dt)
}

# Add column containing values from 0 - 0.9 only
max.k.score.dt[, max_k_ratio_group := case_when(
  max_k_ratio >= 0.9 ~ 0.9, # if ≥ 0.9 → set value to 0.9
  TRUE ~ max_k_ratio  #if < 0.9, then retain original max k ratio value
)]

# Apply the 'extractCompartment' function to each subset of protein grouped by
# max_K_ratio
k.score.compartment.dt <- mapply(
  extractCompartment,
  gene.set = lapply(
    seq(0, 9, by = 1) / 10, # create a sequence of 0 to 0.9 
    function(x){ # For each x, extract the 'uniport_id' belonging to that max_k_ratio_group
      max.k.score.dt[max_k_ratio_group == x, uniprot_id]}
    ),
  class.name = as.character(seq(0, 0.9, by = 0.1)), # convert and assign class names as strings
  SIMPLIFY = FALSE # Return list instead of matrix or array
) %>% rbindlist # combine all lists into a data table
  
# Calculate the negative log10 of FDR and create a new column called 'mlog10FDR'
k.score.compartment.dt[, mlog10FDR := -log10(FDR)]

# Convert data table into wide format 
# rows: 'Compartment'
# column: 'class_name'
# values: 'mlog10FDR'
d.k.score.compartment.dt <- dcast(
  k.score.compartment.dt,
  Compartment ~ factor(class_name),
  value.var = "mlog10FDR"
)

# convert data table into matrix 
d.k.score.compartment.mat <- as.matrix(d.k.score.compartment.dt[, 2:ncol(d.k.score.compartment.dt)])

# Assign Compartment names as rownames of the matrix 
rownames(d.k.score.compartment.mat) <- d.k.score.compartment.dt[, Compartment]

```



# Functional class enrichment

Tests which functional protein classes are enriched among lysine-rich proteins.

```{r functional_classes}

library("mgcv")
library("subcellularvis")

comp.out <- compartmentData(
    max.k.score.dt[, uniprot_id],
    id_type = "UNIPROT",
    aspect = c("Whole cell"),
    organism = c("Human"),
    annotationSource =c("Gene Ontology") #c("Human Protein Atlas")
  )

localisation_coef = function(i){
  localised_genes <- comp.out$enrichment[i, "Genes"] %>% {strsplit(., split = ",", fixed = TRUE)[[1]]}
  max.k.score.dt[, loc := uniprot_id %in% localised_genes]
  len.gam <- gam(max_k_ratio ~ s(protein_len) + loc, data = max.k.score.dt, method = "REML")
  
  dt <- data.table(
    functional_type = "localisation", 
    subtype = comp.out$enrichment[, "Compartment"][i],
    coef = summary(len.gam)$p.coeff["locTRUE"],
    se = summary(len.gam)$se["locTRUE"],
    p_value = summary(len.gam)$p.pv["locTRUE"],
    n = length(localised_genes)
  )
  
  return(dt)
}

localisation_dt <- lapply(1:length(comp.out$enrichment[, "Compartment"]), localisation_coef) %>%
  rbindlist

# Molecular condensates
cd_code <- fread(file.path(data.dir, "public_data/20241113_cd-code.csv"))

condensates_coef <- function(i){  
  condensate_genes <- cd_code[i, Proteins] %>% {strsplit(., split = "\t", fixed = TRUE)[[1]]}
  max.k.score.dt[, loc := uniprot_id %in% condensate_genes]
  len.gam <- gam(max_k_ratio ~ s(protein_len) + loc, data = max.k.score.dt, method = "REML")
  
  dt <- data.table(
    functional_type = "condensates", 
    subtype = cd_code[i, Name],
    coef = summary(len.gam)$p.coeff["locTRUE"],
    se = summary(len.gam)$se["locTRUE"],
    p_value = summary(len.gam)$p.pv["locTRUE"],
    n = length(condensate_genes)
  )
  
  return(dt)
}

condensate_dt <- lapply(1:nrow(cd_code), condensates_coef) %>%
  rbindlist

condensate_dt <- condensate_dt[order(p_value)][1:5]

# histone
max.k.score.dt[, histone_protein := grepl("^H[1-4]", gene_name)]
len.gam <- gam(max_k_ratio ~ s(protein_len) + histone_protein, data = max.k.score.dt, method = "REML")
  
histone_dt <- data.table(
  functional_type = "localisation", 
  subtype = "Histone",
  coef = summary(len.gam)$p.coeff["histone_proteinTRUE"],
  se = summary(len.gam)$se["histone_proteinTRUE"],
  p_value = summary(len.gam)$p.pv["histone_proteinTRUE"],
  n = nrow(max.k.score.dt[histone_protein == TRUE])
)

# Merge all three data
functional_class_summary <- rbindlist(list(
  localisation_dt, histone_dt, condensate_dt
))

functional_class_summary <- functional_class_summary[order(-functional_type, -coef)]
functional_class_summary[, `:=`(
  functional_type = factor(functional_type, levels = c("localisation", "condensates")),
  subtype = factor(subtype, levels = subtype)
)]

functional_class_summary

ggplot(
  data = functional_class_summary,
  aes(
    x = subtype,
    y = coef
  )) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point() +
  geom_errorbar(aes(
    ymin = coef - se, ymax = coef + se, width = 0.3
  )) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  facet_grid(~ functional_type, scales = "free", space = "free")

ggplot(
  data = max.k.score.dt,
  aes(
    x = factor(max_k_ratio),
    y = protein_len,
    fill = histone_protein
  )
) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(preserve = "single")) +
  scale_fill_bright() +
  coord_cartesian(ylim = c(0, 2500))
```

# K score in histones

Examines histones — well-defined lysine-rich proteins — as a reference point.

The definition of histone tail was obtained from "Histone post-translational modifications — cause and consequence of genome function (https://doi.org/10.1038/s41576-022-00468-7)".

```{r histone_k_score_per_position}

#H2A
ggplot(
  protein.feature.dt[grepl("Q6FI13", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0,0.5)) +
  annotate("rect", xmin = 2, xmax = 26, ymin = 0, ymax = 1, fill = "red", alpha = 0.2) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("Q6FI13", Accession)][1, Accession])

#H2B
ggplot(
  protein.feature.dt[grepl("P33778", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0,0.5)) +
  annotate("rect", xmin = 2, xmax = 33, ymin = 0, ymax = 1, fill = "red", alpha = 0.2) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("P33778", Accession)][1, Accession])

#H3
ggplot(
  protein.feature.dt[grepl("Q16695", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0,0.5)) +
  annotate("rect", xmin = 2, xmax = 44, ymin = 0, ymax = 1, fill = "red", alpha = 0.2) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("Q16695", Accession)][1, Accession])

#H4
ggplot(
  protein.feature.dt[grepl("P62805", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0,0.5)) +
  annotate("rect", xmin = 2, xmax = 24, ymin = 0, ymax = 1, fill = "red", alpha = 0.2) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("P62805", Accession)][1, Accession])

print("Median K score of histone tails")
rbindlist(list(
  protein.feature.dt[grepl("Q6FI13", Accession)][2:26],
  protein.feature.dt[grepl("P33778", Accession)][2:33],
  protein.feature.dt[grepl("Q16695", Accession)][2:44],
  protein.feature.dt[grepl("P62805", Accession)][2:24]
))[, median(K_ratio_score)]

```

# K score in disordered regions

Relates lysine richness to protein disorder.

```{r kscore_disorderedness_other_proteins}

library("patchwork")

dkc1.1 <- ggplot(
  protein.feature.dt[grepl("O60832", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0, 1)) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("O60832", Accession)][1, Accession]) +
  geom_point(x = 475, y = 0, color = "red") +
  geom_point(data = data.table(x = c(501:505), y = rep(0, 5)), aes(x = x, y = y), color = "blue")

dkc1.2 <- ggplot(
  protein.feature.dt[grepl("O60832", Accession)],
  aes(
    x = position,
    y = IUPRED2
  )
) +
  coord_cartesian(ylim = c(0, 1)) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("O60832", Accession)][1, Accession])

dkc1.1 / dkc1.2


ARL6IP4.1 <- ggplot(
  protein.feature.dt[grepl("Q66PJ3", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  coord_cartesian(ylim = c(0, 1)) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("Q66PJ3", Accession)][1, Accession]) +
  geom_point(data = data.table(x = c(290, 292, 294), y = rep(0, 3)), aes(x = x, y = y), color = "red") +
  geom_point(data = data.table(x = c(290, 292, 294, 303, 304, 306, 307), y = rep(0, 7)), aes(x = x, y = y), color = "blue")

ARL6IP4.2 <- ggplot(
  protein.feature.dt[grepl("Q66PJ3", Accession)],
  aes(
    x = position,
    y = IUPRED2
  )
) +
  coord_cartesian(ylim = c(0, 1)) +
  geom_area(fill = "gray40") +
  ggtitle(protein.feature.dt[grepl("Q66PJ3", Accession)][1, Accession])

ARL6IP4.1 / ARL6IP4.2


```

# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
