---
title: "1-1. Proteins with lysine-rich domains"
author: "Pallavi Kesavan and Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---


# Environment setup

```{r env_set}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

## Define project directory 
project.dir <- file.path("/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains")

## Use 'renv::snapshot', if there are updates in the current dependencies 
# renv::snapshot(file.path(project.dir, "R"))

## Restore packages
#renv::restore(file.path(project.dir, "R"))

## Check package versions of current loaded library and lockfile 
#renv::status

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files(file.path(project.dir, "R/functions"), pattern="*.R", full.names = TRUE), source)

# Load PTM stiochiometry functions
library("ptm.stoichiometry")
```


# 1.1 Import data

```{r import_data, warning=FALSE}

## Define paths to directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p1_K_rich_domains <- file.path(results.dir, "p1_K_rich_domains")
dir.create(p1_K_rich_domains, showWarnings = FALSE, recursive = TRUE)

## Load data into environment 
protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))

## Load library  
library("org.Hs.eg.db")

## Convert gene id and names
# Map UniProt IDs to gene symbols and ENSEMBL IDs
gene.id.dt <- select(
    org.Hs.eg.db,
    keys = protein.feature.dt[, unique(uniprot_id)],
    columns = c("SYMBOL","ENSEMBL"),
    keytype = "UNIPROT"
) %>%
    data.table

# Rename columns 
setnames(
    gene.id.dt,
    old = c("UNIPROT", "SYMBOL","ENSEMBL"),
    new = c("uniprot_id", "gene_name", "gene_id")
)

## Clean and organize gene mapping table
# Sort gene_id in ascending order, remove duplicate UniProt id and remove rows missing gene_id 
gene.id.dt <- gene.id.dt[order(gene_id)][!duplicated(uniprot_id)][!is.na(gene_id)]

# Rename data file and save to directory 
fwrite(gene.id.dt, file.path(results.dir, "p1_K_rich_domains","uniprot-ensembl.csv"))
```

# 1.2 Maximum K score and Protein Length

```{r Maximum_K_score_Protein_Length}
#-----------------------------------
# Maximum K score and Protein Length
#-----------------------------------

## Compute maximum K_ratio and protein length for each protein
max.k.score.dt <- protein.feature.dt[ 
  , list(
    max_k_ratio = max(K_ratio), 
    protein_len = max(position)
    ), 
  by = list(Accession, uniprot_id)] 

## Merge gene annotations with protein summary data (max K_ratio, protein length)
# Merge tables using the UniProt ID as the key
max.k.score.dt <- merge(
    gene.id.dt,
    max.k.score.dt,
    all.y = TRUE,
    by = "uniprot_id"
)

# Create dataset with max k score 
fwrite(max.k.score.dt, file.path(results.dir, "p1_K_rich_domains", "max-k-score-per-protein.csv")) 

# Import human protein reference data from specified file path 
ref_protein_data <- import_reference_fasta(file.path
("/fast/AG_Sugimoto/reference/uniprot/human", 
  "UP000005640_9606.fasta")) 

# Median K score of all region in human proteome
all_protein_K_ratio <- sapply(ref_protein_data$protein_seq, function(seq){
  if (is.na(seq) || nchar(seq) == 0) # checks for missing values 
    return(NA_real_)
  k_count <- str_count(seq, "K") # counts number of K in each sequence
  len <- nchar(seq) # counts the number of amino acid in sequence
  (k_count/len)*10 # compute K per 10 residues
})

# median K score
median_k_ratio <- median(all_protein_K_ratio, na.rm = TRUE)
print(median_k_ratio)

```

# Data Visualization - K score, Maximum K score and Protein Length

```{r Data_Visualization_Max_K_Score_&_Protein_Length}

#---------------------------------------------------------
# Data Visualization - K score, Maximum K score and Protein Length
#---------------------------------------------------------

## K_ratio distribution 
# Count how many proteins have each K_ratio value
protein.feature.dt[, table(K_ratio)] 

# Create a new column `K_ratio_group` based on the K_ratio
protein.feature.dt[, K_ratio_group := case_when(
  K_ratio >= 0.9 ~ 0.9, #If K_ratio is 0.9 or higher, assign value 0.9
  TRUE ~ K_ratio #If K_ratio is lower than 0.9, then assign original K_ratio value
)]

# Plot - The number of proteins per K score
ggplot(
    protein.feature.dt,
    aes(
        x = K_ratio_group
    )
) +
    geom_bar() +
    scale_x_continuous(breaks=seq(0, 1, 0.1))


# Plot - The number of proteins per max K score
ggplot(
    max.k.score.dt,
    aes(
        x = max_k_ratio
    )
) +
    geom_bar() +
    scale_x_continuous(breaks=seq(0, 1, 0.1)) + 
  labs(x = "max_k_ratio", y = "number of protein_count")

#Median - 24112025
median(max.k.score.dt[, max_k_ratio])
# 0.3

# Plot - Max K ratio vs protein length
ggplot(
  max.k.score.dt,
  aes(
    y = protein_len,
    x = factor(max_k_ratio)
  )
) + geom_boxplot(fill = "steelblue", outlier.shape = NA) +
  labs(x = "max_K_ratio", y = "protein_length") +
  coord_cartesian(ylim = c(0, 2000))

# Plot - K_score of disordered regions of proteins 
# This plot indicates K score depending on the level of protein disorderliness

ggplot(
  protein.feature.dt,
  aes(
    x = factor(K_ratio_group),
    y = IUPRED2
  )
) + 
  geom_boxplot(fill = "steelblue", outlier.shape = NA)


# Plot - Protein charge versus K_ratio_group
ggplot(
  protein.feature.dt,
  aes(
    x = factor(K_ratio_group),
    y = windowCharge
  )
) + 
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(fill = "steelblue", outlier.shape = NA)

```

# 1.3 RNA-Binding proteins

```{r RNA_binding_proteins, warning=FALSE}

#---------------------------
# RNA-Binding Proteins (RBP)
#---------------------------

# Load data into environment 
rbp.dt <- readxl::read_excel(file.path(data.dir, "20241113_RBPbase_Hs_DescriptiveID.xlsx")) %>%
  data.table()

# Remove line breaks and split into two columns
setnames(rbp.dt, old = colnames(rbp.dt), new = str_split_fixed(colnames(rbp.dt), "\n", n = 2)[, 1])
rbp.dt <- janitor::clean_names(rbp.dt, case = "none") #Rename duplicate column names
setnames(rbp.dt, old = "UnitProtSwissProtID_Hs", new = "uniprot_id") 


# Dataset containing only UniProt id and Hs_ columns
rbp.dt <- rbp.dt[
  , c("uniprot_id", grep("^Hs_", colnames(rbp.dt), value = TRUE)), with = FALSE
  ]

# Remove NA and duplicated values
rbp.dt <- rbp.dt[!is.na(uniprot_id) & !duplicated(uniprot_id)]

# Reshape dataset into long format for easy data management and visualization 
m.rbp.dt <- melt(
    rbp.dt,
    id.vars = c("uniprot_id"),
    variable.name = "dataset",
    value.name = "RBP"
)

# Dataset containing count of identified RBP and total RBP 
rbp.count.dt <- m.rbp.dt[!is.na(uniprot_id),
  list(
    identified_RBP = sum(RBP == "YES"),
    total = sum(RBP == "YES") + sum(RBP == "no") 
    ), by = uniprot_id
  ]

rbp.count.dt[, table(identified_RBP)]

#-----------------------------------
## Comparing Max K score versus RBP
#-----------------------------------

# Create dataset containing max K score and RBP count
rbp.count.dt <- merge(
  max.k.score.dt, rbp.count.dt, by = "uniprot_id"
)

data.set.names <- c(
  "Others", 
  "RBP (1 dataset)", "RBP (2 dataset)", "RBP (> 2 dataset)"
  )

# Assign values to RBP dataset columns based on RBP count 
rbp.count.dt[, RBP := case_when(
  identified_RBP > 2 ~ "RBP (> 2 dataset)",
  identified_RBP > 1 ~ "RBP (2 dataset)",
  identified_RBP > 0 ~ "RBP (1 dataset)", 
  TRUE ~ "Others") %>% factor(levels = data.set.names)]
rbp.count.dt[, table(RBP)]

```

# Data Visualization - RNA-Binding Proteins (Max K score versus RBP)

```{r Data_Visualization_RNA_Binding_Proteins}

# Plot - Max K score versus RBP
 ggplot(
  data = rbp.count.dt,
  aes(
    x = RBP,
    y = max_k_ratio
    )
  ) +
  geom_boxplot(fill = "steelblue", outlier.shape = NA) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(aspect.ratio = 3)

```


# 1.4 Subcellular localisation

```{r subcellular_localisation_analysis}

#--------------------------------------
# Sub-cellular localisation analysis
#--------------------------------------

# Packaged installed 20th Oct 2025
# install.packages(file.path("/fast/AG_Sugimoto/home/users/pallavi/Software_packages/R/subcellularvis"), repos = NULL, type = "source")

## Load library
library("subcellularvis")

## Run sub-cellular compartment analysis with enrichment score according Gene Ontology
# The 'extractCompartment' function runs sub-cellular compartment analysis, retrieves 
extractCompartment <- function(gene.set, class.name){
  comp.out <- compartmentData(  # Run compartment enrichment analysis using the compartmentData() function
    gene.set,
    id_type = "UNIPROT",
    aspect = c("Whole cell"),
    organism = c("Human"),
    annotationSource =c("Gene Ontology") # Alternatively, Human Protein Atlas can also be used
  )
  
  # Convert enrichment results from comp.out into a data.table 
  comp.out.dt <- data.table(comp.out$enrichment)
  
  # Add metadata columns 
  comp.out.dt[, `:=`(
    Genes = NA,
    n_unmapped = length(str_split(comp.out$unmapped, ",")[[1]]), # Counting the number of unmapped IDs
    n_mapped = comp.out$nMapped, # adding mapped IDs
    class_name = class.name 
  )]
  
  #  Updated data table 
  return(comp.out.dt)
}

# Add column containing values from 0 - 0.9 only
max.k.score.dt[, max_k_ratio_group := case_when(
  max_k_ratio >= 0.9 ~ 0.9, # if ≥ 0.9 → set value to 0.9
  TRUE ~ max_k_ratio  #if < 0.9, then retain original max k ratio value
)]

# Apply the 'extractCompartment' function to each subset of protein grouped by
# max_K_ratio
k.score.compartment.dt <- mapply(
  extractCompartment,
  gene.set = lapply(
    seq(0, 9, by = 1) / 10, # create a sequence of 0 to 0.9 
    function(x){ # For each x, extract the 'uniport_id' belonging to that max_k_ratio_group
      max.k.score.dt[max_k_ratio_group == x, uniprot_id]}
    ),
  class.name = as.character(seq(0, 0.9, by = 0.1)), # convert and assign class names as strings
  SIMPLIFY = FALSE # Return list instead of matrix or array
) %>% rbindlist # combine all lists into a data table
  
# Calculate the negative log10 of FDR and create a new column called 'mlog10FDR'
k.score.compartment.dt[, mlog10FDR := -log10(FDR)]

# Convert data table into wide format 
# rows: 'Compartment'
# column: 'class_name'
# values: 'mlog10FDR'
d.k.score.compartment.dt <- dcast(
  k.score.compartment.dt,
  Compartment ~ factor(class_name),
  value.var = "mlog10FDR"
)

# convert data table into matrix 
d.k.score.compartment.mat <- as.matrix(d.k.score.compartment.dt[, 2:ncol(d.k.score.compartment.dt)])

# Assign Compartment names as rownames of the matrix 
rownames(d.k.score.compartment.mat) <- d.k.score.compartment.dt[, Compartment]

```


# Data Visualization - Subcellular Localisation

```{r Data_Visualization_Subcellular_Localisation}

# Load libraries -  'RColorBrewer' and 'pheatmap'
library(RColorBrewer)
library("pheatmap")

mat_breaks <- seq(min(d.k.score.compartment.mat), max(d.k.score.compartment.mat), length.out = 40)

# Plot - heatmap of subcellular localisation
pheatmap(
  d.k.score.compartment.mat,
  cluster_cols = FALSE,
  color = colorRampPalette(c("white", "steelblue"))(length(mat_breaks)),
  breaks = mat_breaks
)

# Plot - max_k_score versus −log10(FDR) of nuclear localisation enrichment
ggplot(
  data = k.score.compartment.dt[Compartment == "Nucleus"],
  aes(
    x = factor(class_name),
    y = mlog10FDR
  )
) +
  geom_bar(stat = "identity") +
  geom_hline(yintercept = -log10(0.05), color = "gray60") +
  ylab("-log10(FDR) of nuclear localisation enrichment") +
  xlab("Maximum K score of protein")

```

# 1.5 Enrichment of lysine-rich domains by functional protein classes

```{r functional_classes}

library("mgcv")
library("subcellularvis")

comp.out <- compartmentData(
    max.k.score.dt[, uniprot_id],
    id_type = "UNIPROT",
    aspect = c("Whole cell"),
    organism = c("Human"),
    annotationSource =c("Gene Ontology") #c("Human Protein Atlas")
  )

localisation_coef = function(i){
  localised_genes <- comp.out$enrichment[i, "Genes"] %>% {strsplit(., split = ",", fixed = TRUE)[[1]]}
  max.k.score.dt[, loc := uniprot_id %in% localised_genes]
  len.gam <- gam(max_k_ratio ~ s(protein_len) + loc, data = max.k.score.dt, method = "REML")
  
  dt <- data.table(
    functional_type = "localisation", 
    subtype = comp.out$enrichment[, "Compartment"][i],
    coef = summary(len.gam)$p.coeff["locTRUE"],
    se = summary(len.gam)$se["locTRUE"],
    p_value = summary(len.gam)$p.pv["locTRUE"],
    n = length(localised_genes)
  )
  
  return(dt)
}

localisation_dt <- lapply(1:length(comp.out$enrichment[, "Compartment"]), localisation_coef) %>%
  rbindlist

# Molecular condensates
cd_code <- fread(file.path(data.dir, "20241113_cd-code.csv"))

condensates_coef <- function(i){  
  condensate_genes <- cd_code[i, Proteins] %>% {strsplit(., split = "\t", fixed = TRUE)[[1]]}
  max.k.score.dt[, loc := uniprot_id %in% condensate_genes]
  len.gam <- gam(max_k_ratio ~ s(protein_len) + loc, data = max.k.score.dt, method = "REML")
  
  dt <- data.table(
    functional_type = "condensates", 
    subtype = cd_code[i, Name],
    coef = summary(len.gam)$p.coeff["locTRUE"],
    se = summary(len.gam)$se["locTRUE"],
    p_value = summary(len.gam)$p.pv["locTRUE"],
    n = length(condensate_genes)
  )
  
  return(dt)
}

condensate_dt <- lapply(1:nrow(cd_code), condensates_coef) %>%
  rbindlist

condensate_dt <- condensate_dt[order(p_value)][1:5]

# histone
max.k.score.dt[, histone_protein := grepl("^H[1-4]", gene_name)]
len.gam <- gam(max_k_ratio ~ s(protein_len) + histone_protein, data = max.k.score.dt, method = "REML")
  
histone_dt <- data.table(
  functional_type = "localisation", 
  subtype = "Histone",
  coef = summary(len.gam)$p.coeff["histone_proteinTRUE"],
  se = summary(len.gam)$se["histone_proteinTRUE"],
  p_value = summary(len.gam)$p.pv["histone_proteinTRUE"],
  n = nrow(max.k.score.dt[histone_protein == TRUE])
)

# Merge all three data
functional_class_summary <- rbindlist(list(
  localisation_dt, histone_dt, condensate_dt
))

functional_class_summary <- functional_class_summary[order(-functional_type, -coef)]
functional_class_summary[, `:=`(
  functional_type = factor(functional_type, levels = c("localisation", "condensates")),
  subtype = factor(subtype, levels = subtype)
)]

functional_class_summary

ggplot(
  data = functional_class_summary,
  aes(
    x = subtype,
    y = coef
  )) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_point() +
  geom_errorbar(aes(
    ymin = coef - se, ymax = coef + se, width = 0.3
  )) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  facet_grid(~ functional_type, scales = "free", space = "free")

ggplot(
  data = max.k.score.dt,
  aes(
    x = factor(max_k_ratio),
    y = protein_len,
    fill = histone_protein
  )
) +
  geom_boxplot(outlier.shape = NA, position = position_dodge(preserve = "single")) +
  scale_fill_bright() +
  coord_cartesian(ylim = c(0, 2500))
```


```{r histone_versus_max_K_score}
library("RColorBrewer")

#install.packages("ggpubr")
library("ggpubr")

# Filter the K scores for histones 1-4
max.k.score.dt[, histone_protein := grepl("^H[1-4]", gene_name)]
histone_protein_dt <- max.k.score.dt[histone_protein == "TRUE"]

median_k_ratio <- histone_protein_dt[, median(max_k_ratio, na.rm = TRUE)]
print(median_k_ratio)

range_max_k <- histone_protein_dt[, range(max_k_ratio, na.rm = TRUE)]
print(range_max_k)

H2A_dt <- histone_protein_dt[grepl("^H2A", gene_name)]
H2B_dt <- histone_protein_dt[grepl("^H2B", gene_name)]
H3_dt <- histone_protein_dt[grepl("^H3", gene_name)]
H4_dt <- histone_protein_dt[grepl("^H4", gene_name)]


# Plot - Histones 
P2_H2A <- ggplot(
  data = H2A_dt,
  mapping = aes(
    x = gene_name,
    y = max_k_ratio,
  )
) +
  geom_col(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8)) +
  ggtitle("") + 
  xlab("") + 
  ylab("")


P3_H2B <- ggplot(
  data = H2B_dt,
  mapping = aes(
    x = gene_name,
    y = max_k_ratio,
  )
) +
  geom_col(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8)) +
  ggtitle("") + 
  xlab("") + 
  ylab("")

P4_H3 <- ggplot(
  data = H3_dt,
  mapping = aes(
    x = gene_name,
    y = max_k_ratio,
  )
) +
  geom_col(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8)) +
  ggtitle("") + 
  xlab("") + 
  ylab("")

P5_H4 <- ggplot(
  data = H4_dt,
  mapping = aes(
    x = gene_name,
    y = max_k_ratio,
  )
) +
  geom_col(fill = "steelblue") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme(axis.text.x = element_text(size = 8),
        axis.text.y = element_text(size = 8)) +
  ggtitle("") + 
  xlab("") + 
  ylab("")


combined <- ggarrange(
  P2_H2A, P3_H2B, P4_H3, P5_H4,
  ncol = 2, nrow = 2,
  labels = c("H2A", "H2B", "H3", "H4"),
  label.x = 0.0,          
  label.y = 1,  
  hjust = -0.5, 
  vjust = 1,
  heights = 1,
  font.label = list(size = 8, face = "bold")
)

annotate_figure(
  combined,
  left   = text_grob("Max K Score", rot = 90, size = 10),
  bottom = text_grob("Histone", size = 10)
)


```
```{r k_score_per_position}
#H3
ggplot(
  protein.feature.dt[grepl("P62805", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  geom_area(color = "black", fill = "steelblue", alpha = 0.6) 
# H2AX 
ggplot(
  protein.feature.dt[grepl("P16104", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  geom_area(color = "black", fill = "steelblue", alpha = 0.6)

#H2B
ggplot(
  protein.feature.dt[grepl("P33778", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  geom_area(color = "black", fill = "steelblue", alpha = 0.6) 

#H4

ggplot(
  protein.feature.dt[grepl("P62805", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  geom_area(color = "black", fill = "steelblue", alpha = 0.6) 

#H2A

ggplot(
  protein.feature.dt[grepl("Q6FI13", Accession)],
  aes(
    x = position,
    y = K_ratio_score
  )
) +
  geom_area(color = "black", fill = "steelblue", alpha = 0.6) 



```



# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
