---
title: "p2-09 · Export stoichiometry tables"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Export processed BRD stoichiometry (MS_SS, MS_KR_1, PNAS) to a
formatted Excel workbook (supplementary data).

**Inputs:** stoichiometry tables from p2-01; reference proteome.

**Outputs:** Excel workbook under `results/`.

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")
library("openxlsx")

```

# Export tables to Excel

Assembles the per-site stoichiometry tables (PNAS data-A/B/C, MS_KR_1, MS_SS) and writes them to a formatted Excel workbook for the supplementary data.

```{r PNAS_DI_load_data_excel}



#---------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset A)
#---------------------------------------------------------------------

 MQ_DI_dtA_stoic_dt <- read_stoic_data(
  prefix = "DI_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtA_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset B1 and B2)
#--------------------------------------------------------------------------

# Load sample run info data (MQ DI with no waterloss)
MQ_DI_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_run_info[, sample_id := 1:.N]

MQ_DI_dtB_stoic_dt <- lapply(
  MQ_DI_run_info[data != c("data-A", "data-C"), prefix],
  read_stoic_data,
  pre_prefix = "DI_", 
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI")
) %>% rbindlist

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtB_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#---------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset C)
#---------------------------------------------------------------------

 MQ_DI_dtC_stoic_dt <- read_stoic_data(
  prefix = "DI_data-C_trp_m7_v7_mCC_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtC_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]


#------------------------------------------------------
# Load PNAS stoichiometry data manually curated sites
#------------------------------------------------------

# Load PNAS stoichiometry data
pnas2022_stoic_dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022_stoic_dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022_stoic_dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022_stoic_dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

## Add columns to pnas2022.stoic.dt metadata
# new columns: accession_position, ptm, sample_name 
pnas2022_stoic_dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos), #this column combines data from protein_accession and aa_pos
    ptm = fifelse(curated_oxK_site == TRUE, "[Oxidation (K)]", " ")
    )
  ]

#Change col names to make it same across datasets
setnames(pnas2022_stoic_dt, c("stoichiometry", "protein_accession", "gene_name", "aa_pos", "curated_oxK_site"),  c("Stoichiometry_PNAS2022", "Protein_accession", "Gene_name", "Amino_acid_pos", "PNAS2022_curated_oxK_site"))

#-----------------------------
# Load MS_KR_1 stoichiometry data (Dataset D)
#-----------------------------

## Read stoichiometry data for MS_KR1 data
 MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data (Dataset E)
#-----------------------------

## Read stoichiometry data for MS_SS data 
 MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

```


```{r PNAS_MQ_DI_data_A}

# Subset the data to contain only Lysine
MQ_DI_dtA_stoic_dt <- MQ_DI_dtA_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtA_stoic_dt <-  MQ_DI_dtA_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtA_stoic_dt,
  append(setdiff(names(MQ_DI_dtA_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtA_stoic_dt,
  append(setdiff(names(MQ_DI_dtA_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MQ_DI_dtA_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset A and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtA_stoic_dt <- MQ_DI_dtA_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtA_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtA_stoic_dt.xlsx"))

```


```{r PNAS_MQ_DI_data_B}

# Subset the data to contain only Lysine
MQ_DI_dtB_stoic_dt <- MQ_DI_dtB_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtB_stoic_dt <-  MQ_DI_dtB_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtB_stoic_dt,
  append(setdiff(names(MQ_DI_dtB_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtB_stoic_dt,
  append(setdiff(names(MQ_DI_dtB_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MQ_DI_dtB_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Total_number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset B1 and B2 and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtB_stoic_dt <- MQ_DI_dtB_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtB_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtB_stoic_dt.xlsx"))

```


```{r PNAS_MQ_DI_data_C}

# Subset the data to contain only Lysine
MQ_DI_dtC_stoic_dt <- MQ_DI_dtC_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtC_stoic_dt <-  MQ_DI_dtC_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtC_stoic_dt,
  append(setdiff(names(MQ_DI_dtC_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtC_stoic_dt,
  append(setdiff(names(MQ_DI_dtC_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

# changing column names - publication ready
setnames(MQ_DI_dtC_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset C and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtC_stoic_dt <- MQ_DI_dtC_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]
 
 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtC_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtC_stoic_dt.xlsx"))

```


```{r MS_KR_excel}

# Subset the data to contain only Lysine
MS_KR1_stoic_K_dt <- MS_KR1_stoic_dt[aa == "K"]

# removing unused data columns from data table
MS_KR1_stoic_K_dt <-  MS_KR1_stoic_K_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition") := NULL]

# changing stoichiometry column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MS_KR1_stoic_K_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))


# Merge MS_KR_1 and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MS_KR1_stoic_K_dt <- MS_KR1_stoic_K_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to Excel table
write.xlsx(MS_KR1_stoic_K_dt, file.path(results.dir, "p2_MS_SS_KR", "MS_KR1_stoic_K.xlsx"))

```


```{r MS_SS_excel}

# Subset the data to contain only Lysine
MS_SS_stoic_K_dt <- MS_SS_stoic_dt[aa == "K"]

# removing unsed data columns from data table
MS_SS_stoic_K_dt <-  MS_SS_stoic_K_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition") := NULL]

# Remove BRD2/3 samples reporting BRD4 stoichiometry and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_stoic_K_dt <- MS_SS_stoic_K_dt[
  # Drop cross-protein false positives: BRD4 signal reported in BRD2/3 samples
  # (and vice versa), e.g. from shared / homologous peptides.
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# changing stoichiometry column position 
setcolorder(
  MS_SS_stoic_K_dt,
  append(setdiff(names(MS_SS_stoic_K_dt), "stoichiometry"), "stoichiometry", after = 6)
)

# changing sample_group column position 
setcolorder(
  MS_SS_stoic_K_dt,
  append(setdiff(names(MS_SS_stoic_K_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MS_SS_stoic_K_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))


# Merge MS_SS and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MS_SS_stoic_K_dt <- MS_SS_stoic_K_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to Excel table
write.xlsx(MS_SS_stoic_K_dt, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_stoic_K.xlsx"))

```


# Session information

```{r session_Info}

sessioninfo::session_info()

```
