---
title: "p2-02 · Optimise database search"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Determine optimal MaxQuant database-search settings for
lysine-derivatised samples — effect of miscleavages and modifications on MS/MS
counts, coverage of lysine-rich regions, and runtime.

**Inputs:** MaxQuant evidence / runtime / mqpar under `data/MQ_Std/PNAS2022/`; protein feature data.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

## Script-specific packages and the results subdirectory used throughout
library("RColorBrewer")
library("patchwork")
p2_results_dir <- file.path(results.dir, "p2-analysis-setting")

```

# Import data

Loads the sample run-info table, the protein feature data, and the reference proteome used in the comparisons below.

```{r load_inputs}

# Sample run info (one row per run); data.dir is defined in _setup.R
all_sample_run_info <- read_excel(
  file.path(data.dir, "analysis_setting", "PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_Std"
) %>% data.table

# Protein feature data, harmonised to the shared column names
protein_feature_dt <- fread(
  file.path(data.dir, "processed_data_from_PNAS2022/all_protein_feature_per_position.csv")
)
setnames(
  protein_feature_dt,
  old = c("uniprot_id", "position"),
  new = c("protein_accession", "aa_pos")
)

```

# Helper functions

Readers and the coverage-plot helper used by the analysis chunks below.

```{r helper_functions}

# Read a processed evidence table (MULTI-MSMS + MULTI-SECPEP search). Returns an
# empty data.table if the file is missing so missing runs are skipped silently.
read_evidence <- function(prefix, dir_path) {
  input_file <- file.path(
    dir_path, paste0("including_SECPEP_", prefix, "all_processed_evidence_data.csv")
  )
  if (file.exists(input_file)) {
    dt <- fread(input_file)
    dt[, condition := gsub("_$", "", prefix)]
  } else {
    dt <- data.table()
  }
  dt
}

# Read a per-position table from the MULTI-MSMS + MULTI-SECPEP search
read_per_pos_secpep <- function(prefix, dir_path) {
  dt <- fread(file.path(dir_path, paste0("including_SECPEP_", prefix, "all_per_pos_data.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  dt
}

# Read a MaxQuant runtime table, tidy its column/job names, and relabel the
# search-stage jobs to publication-friendly names.
read_runtime_data <- function(prefix, dir_path) {
  mq_setting <- gsub("_$", "", sub("^data-[^_]+_", "", prefix))
  dt <- fread(file.path(dir_path, mq_setting, paste0(prefix, "runningTimes.txt")))
  dt[, condition := gsub("_$", "", prefix)]

  dt <- janitor::clean_names(dt)
  dt[, job := janitor::clean_names(setNames(nm = job)) %>% names %>% as.character]

  dt[, job := dplyr::case_when(
    job == "ms_ms_first_search"     ~ "First_search",
    job == "ms_ms_main_search"      ~ "Main_search",
    job == "second_peptide_search"  ~ "Second_peptide_search",
    TRUE ~ job
  )]
  dt
}

# Plot per-position PSM coverage of an accession across the m2/m5/m7 settings
plot_coverage <- function(stoichiometry_data, accession, plot_range) {
  coverage_dt <- lapply(
    c("data-A_trp_m2_v2_def", "data-A_trp_m5_v5_def", "data-A_trp_m7_v7_def"),
    function(setting) {
      preprocess_stoichiometry_data_for_plotting(
        stoichiometry_data = stoichiometry_data[
          sample_name == "JQ1_HeLaWT_derivatised" & condition == setting
        ],
        accession      = accession,
        plot_range     = plot_range,
        all.protein.bs = all.protein.bs
      )$coverage_data %>%
        {.[, condition := setting]}
    }
  ) %>% rbindlist

  ggplot(
    coverage_dt,
    aes(x = aa_pos, y = psm_coverage, color = condition)
  ) +
    geom_line() +
    scale_x_continuous(
      breaks = plot_range[1]:plot_range[2],
      labels = all.protein.bs[
        str_split_fixed(names(all.protein.bs), "\\|", n = 3)[, 2] == accession
      ] %>%
        {strsplit(as.character(.), "")[[1]]} %>%
        {.[plot_range[1]:plot_range[2]]}   # amino-acid residues as axis labels
    ) +
    scale_color_bright() +
    theme(axis.ticks.x = element_blank(), aspect.ratio = 0.3)
}

```

# MS/MS counts by setting

Compares the number of MS/MS identifications obtained under each MaxQuant search setting.

```{r plot_msms_count}

# Benchmark the search settings on data-A and data-D to compare BRD datasets with or without derivatisation
evidence_dt <- lapply(
  all_sample_run_info[data %in% c("data-A", "data-D") & !is.na(prefix), prefix],
  read_evidence,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS_SECPEP")
) %>% rbindlist

data_count_dt <- evidence_dt[, list(
  total_peptide_count = .N,                 # number of peptides
  total_msms_count    = sum(psm_mapped),    # mapped PSMs
  total_intensity     = sum(peak_intensity)
), by = list(file_name, sample_name, condition, type)]

data_count_dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting       = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2]
)]

# Counts relative to the m2_v2 / MULTI-MSMS baseline of each file
data_col <- c("total_peptide_count", "total_msms_count", "total_intensity")

m2_v2_dt <- data_count_dt[
  MQ_setting == "m2_v2" & type == "MULTI-MSMS"
][, c("file_name", data_col), with = FALSE]
setnames(m2_v2_dt, old = data_col, new = paste0("m2_v2_", data_col))

rel_count_dt <- merge(data_count_dt, m2_v2_dt, by = "file_name")

rel_count_dt[, `:=`(
  rel_total_peptide_count = total_peptide_count / m2_v2_total_peptide_count,
  rel_msms_count          = total_msms_count / m2_v2_total_msms_count,
  rel_intensity           = total_intensity / m2_v2_total_intensity
)]

rel_count_dt[type == "MULTI-MSMS", .N, by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_dt[, .N, by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_stat_dt <- rel_count_dt[, list(
  mean_rel_total_peptide_count = mean(rel_total_peptide_count),
  sd_rel_total_peptide_count   = sd(rel_total_peptide_count),
  mean_rel_msms_count          = mean(rel_msms_count),
  sd_rel_msms_count            = sd(rel_msms_count),
  mean_rel_intensity           = mean(rel_intensity),
  sd_rel_intensity             = sd(rel_intensity)
), by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_stat_dt[, `:=`(
  protease_setting = factor(protease_setting, levels = c("trp", "argC")),
  type             = factor(type, levels = c("MULTI-MSMS", "MULTI-SECPEP"))
)]

ggplot(
  rel_count_stat_dt[type == "MULTI-MSMS"],
  aes(x = MQ_setting, y = mean_rel_msms_count, fill = type)
) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(
    ymin = mean_rel_msms_count - sd_rel_msms_count,
    ymax = mean_rel_msms_count + sd_rel_msms_count
  ), width = 0.5, position = position_dodge(width = 0.9)) +
  facet_grid(~ propionylation + protease_setting, scales = "free", space = "free") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Relative MS/MS count") +
  scale_fill_manual(values = c("MULTI-MSMS" = "#4477AA")) #+
# scale_color_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE"))

```

The results indicate that the setting with 7 miscleavage and 7 variable modifications identify the highest MS/MS count.


# Propionylation position QC

Quality control: checks the positional distribution of propionylated lysines, confirming efficient lysine derivatisation.

```{r plot_propionylation_qc}

# QC uses data-A searched with m7_v7 (7 miscleavages).
all_per_pos_secpep_dt <- lapply(
  all_sample_run_info[data %in% c("data-A") & grepl("m7_v7", prefix) & !is.na(prefix), prefix],
  read_per_pos_secpep,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS_SECPEP")
) %>% rbindlist

all_per_pos_secpep_dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting        = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting  = str_split_fixed(condition, "_", n = 3)[, 2],
  is_peptide_C_term = peptide_end == aa_pos,
  is_propionylated  = ptm %in% c("[Propionylation]", "[Oxidised Propionylation]")
)]

all_per_k_pos_secpep_dt <- all_per_pos_secpep_dt[aa == "K"]

propionylated_k_count_dt <- all_per_k_pos_secpep_dt[
  , list(total_MS_MS_count = sum(psm_mapped)),
  by = list(is_propionylated, is_peptide_C_term, type)
]

propionylated_k_count_dt

ggplot(
  propionylated_k_count_dt,
  aes(x = is_peptide_C_term, y = total_MS_MS_count, fill = is_propionylated)
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_fill_manual(values = c("TRUE" = "#4477AA", "FALSE" = "#BBBBBB")) +
  facet_grid(~ type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


Based on the analyses here, we decided to focus on using MULTI-MSMS data.


# Coverage by K score

Examines how the search settings change coverage as a function of lysine richness (K score).

```{r plot_coverage_by_kscore}

kscore_stoic_dt <- lapply(
  all_sample_run_info[
    data != "data-D" &
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) &
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS")
) %>% rbindlist

kscore_stoic_dt[, MQ_setting := str_extract(condition, "m\\d+_v\\d+")]

ms_ms_count_dt <- kscore_stoic_dt[
  , list(total_ms_ms_count = sum(sum_psm_mapped)),
  by = list(protein_accession, aa_pos, MQ_setting)
]

ms_ms_count_dt <- merge(
  ms_ms_count_dt,
  protein_feature_dt[, .(protein_accession, aa_pos, K_ratio, K_ratio_score, IUPRED2)],
  by = c("protein_accession", "aa_pos")
)

# One column of MS/MS counts per setting (wide form)
ms_ms_count_wide_dt <- dcast(
  copy(ms_ms_count_dt)[, IUPRED2 := NULL],
  protein_accession + aa_pos + K_ratio + K_ratio_score ~ MQ_setting,
  value.var = "total_ms_ms_count", fill = 0
)

ms_ms_count_wide_dt[, table(K_ratio)]

p1 <- ggplot(
  ms_ms_count_wide_dt[order(K_ratio)],
  aes(x = m2_v2, y = m5_v5)
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p2 <- ggplot(
  ms_ms_count_wide_dt[order(K_ratio)],
  aes(x = m5_v5, y = m7_v7)
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p1 + p2

p3 <- ggplot(
  ms_ms_count_wide_dt,
  aes(x = factor(K_ratio), y = m5_v5 - m2_v2)
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25)) +
  theme(aspect.ratio = 1)

p4 <- ggplot(
  ms_ms_count_wide_dt,
  aes(x = factor(K_ratio), y = m7_v7 - m5_v5)
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25))

p3 + p4

```


# Runtime by setting

Compares MaxQuant search runtime across the settings.

```{r plot_runtime}

# Prefixes for the data-A runtime comparison (reused for the condition factor)
runtime_prefixes <- all_sample_run_info[
  data == "data-A" &
    !grepl("including_SECPEP", prefix) &
    grepl("m[2-8]_v[2-8]_(def|mCC)", prefix) &
    !is.na(prefix), prefix
]

all_runtime_dt <- lapply(
  runtime_prefixes,
  read_runtime_data,
  dir_path = file.path(data.dir, "MQ_Std/PNAS2022/data-A")
) %>% rbindlist

job_names <- all_runtime_dt[, unique(job)]

all_runtime_dt[, `:=`(
  job       = factor(job, levels = rev(job_names)),
  condition = factor(condition, levels = gsub("_$", "", runtime_prefixes))
)]

all_runtime_dt[, protease_setting :=
  str_split_fixed(condition, "_", n = 5)[, 2] %>% factor(levels = c("trp", "argC"))
]

runtime_plot_colors <- setNames(rep("gray80", length(job_names)), nm = job_names)
runtime_plot_colors[c("First_search", "Main_search", "Second_peptide_search")] <-
  c("royalblue1", "royalblue2", "royalblue3")

ggplot(
  all_runtime_dt,
  aes(x = condition, y = running_time_min, fill = job),
  color = NA
) +
  geom_bar(stat = "identity") +
  scale_fill_manual(
    breaks = c("First_search", "Main_search", "Second_peptide_search"),
    values = runtime_plot_colors
  ) +
  facet_grid(~ protease_setting, scales = "free_x", space = "free") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Run time [min]")

```

# BRD2/3/4 coverage

Coverage of the lysine-rich regions of BRD2, BRD3 and BRD4 under each setting.

```{r plot_brd_coverage}

coverage_stoic_dt <- lapply(
  all_sample_run_info[
    data == "data-A" &
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) &
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS")
) %>% rbindlist

# BRD4
plot_coverage(coverage_stoic_dt, accession = "O60885", plot_range = c(531, 581))

# BRD3
plot_coverage(coverage_stoic_dt, accession = "Q15059", plot_range = c(483, 533))

# BRD2
plot_coverage(coverage_stoic_dt, accession = "P25440", plot_range = c(540, 590))

```


# Notes

The following sites were identified by non-unique peptides (therefore total number used here is 153 instead of 150 in the paper).

           Accession position curated_oxK_site    screen
1:  Q14331|FRG1_HUMAN       27  JMJD6_substrate FLAGJMJD6
2:  Q14331|FRG1_HUMAN       29  JMJD6_substrate FLAGJMJD6
3:  Q14331|FRG1_HUMAN       30  JMJD6_substrate FLAGJMJD6
4: Q9UQ35|SRRM2_HUMAN      241  JMJD6_substrate FLAGJMJD6
5: Q9UQ35|SRRM2_HUMAN      243  JMJD6_substrate FLAGJMJD6
6: Q9UQ35|SRRM2_HUMAN      244  JMJD6_substrate FLAGJMJD6

In the end, 91% of hydroxylated sites were covered by the data, and 80% (120 / 150) of reported hydroxylation sites were identified by this workflow.



# Session information

```{r session_info}

sessioninfo::session_info()

```