---
title: "p2-04 · Identify diagnostic ions"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Identify immonium / diagnostic ions that mark hydroxylysine,
using FragPipe diagnostic-ion output and a WT vs JMJD6-KO contrast.

**Inputs:** FragPipe `diagnosticIons.tsv` + `psm.tsv` (`data/FP_diagnostic_ion_search/`); data-A sample info; reference proteome.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** p2-05.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")

```

# Import data

Loads the FragPipe diagnostic-ion and PSM tables, the data-A sample info, and the reference proteome.

```{r import_basic_data}

# Define path to data directory
data.dir <- file.path(project.dir, "data")

# Load diagnostic ion data
diagnostic_ion_data <- file.path(
  data.dir, "FP_diagnostic_ion_search/fragpipe_dataset-A/dataset01.diagnosticIons.tsv" 
) %>% 
  fread %>%
  clean_names

# Load Fragpipe psm data 
fragpipe_psm <- file.path(
  data.dir, "FP_diagnostic_ion_search/fragpipe_dataset-A/psm.tsv"
) %>%
  fread %>%
  clean_names

# Load MQ standard data
MQ_Std_data <- file.path(
  data.dir,
  "MQ_Std/PNAS2022" 
)

# Load Data A sample info from data directory 
dataA_sample_info <- fread(file.path(
  MQ_Std_data, "sample_info/MS_dataset_overview_PXD031221_data-A.csv"
))

# Import human protein reference data

```

# Identify diagnostic ions

Identifies which immonium / diagnostic-ion masses mark hydroxylysine, by contrasting their intensity between JMJD6 WT and KO cells at BRD4 sites.

```{r identify_diagnostic_ions}

# Merge 'diagnostic_ion_data' and 'fragpipe_psm' data by column 'spectrum'
diagnostic_ion_data <- merge(
  fragpipe_psm[, .(spectrum, protein_id, gene, protein_start, protein_end)], 
  diagnostic_ion_data, 
  by = "spectrum"
)

# Create new column 'file_name' from column 'spectrum'  
diagnostic_ion_data[, `:=`(
  file_name = str_split_fixed(spectrum, "\\.", n = 2)[, 1] # split 'spectrum' by the first '.', take the part before it 
)] 

# Merge 'dataA_sample_info' and 'diagnostic_ion_data' by column 'file_name'
diagnostic_ion_data <- merge(
  dataA_sample_info,
  diagnostic_ion_data,
  by = "file_name"
)

# Retrieve hydroxylation info for BRD4
brd4_hydroxylation_site_di <- diagnostic_ion_data[
  # Focus on the BRD4 lysine-rich region (around residues 535-555).
  # NOTE: confirm the residue range.
  gene == "BRD4" & (555 >= protein_start & 535 <= protein_end)
]

# Reshape data table from wide to long format 
m.brd4_hydroxylation_site_di <- melt(
  brd4_hydroxylation_site_di,
  measure.vars = grep("intensity$",colnames(brd4_hydroxylation_site_di), value = TRUE), # get column names whose names end with "intensity" 
  value.name = "intensity" # new column - "intensity" 
)

# Add columns 'monoisotopic_mass', 'diagnostic_ion' and 'genotype'
m.brd4_hydroxylation_site_di[, `:=`(
  monoisotopic_mass = variable %>% # take values from column 'variable'
    str_replace_all("ox_", "") %>% # remove the prefix "ox_"
    str_replace_all("_intensity", "") %>% # remove the suffix "_intensity"
    str_replace_all("_", ".") %>% # replace remaining underscores with dots
    factor(levels = c(
      "101.1079", #Lysine	K			immonium ion	C5 H13 N2 +
      "100.0762", #Hydroxylation (K)	K	O	15.9949	diagnostic ion	C5 H10 N O+
      "82.0657", #Hydroxylation (K)	K	O	15.9949	diagnostic ion (water loss)	C5 H8 N+
      "117.1028", #Hydroxylation (K)	K	O	15.9949	immonium ion	C5 H13 O N2 + calculated
      "117.0658", #Hydroxylation (K)	K	O	15.9949	immonium ion	C5 H13 O N2 + (from citation)
      "138.0919", #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937	diagnostic ion (water loss)	C8 H12 N1 O+
      "145.0977", #Hydroxylation (K)	K	O	15.9949	(intact, water loss)	C6H13N2O2+
      "156.1025", #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937	diagnostic ion	C8 H14 N1 O2+
      "173.1290" #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937	immonium ion	C8 H17 O2 N2 +
    )),
  diagnostic_ion = intensity > 0, 
  genotype = factor(genotype, levels = c("WT", "JMJD6KO"))
)]


# Load libraries
library("ggbeeswarm")
library("khroma")

# Plot - Quasirandom plot of monoisotopic mass versus intensity in WT and JMJD6KO genotype
ggplot(
  data = m.brd4_hydroxylation_site_di,
  aes(
    x = monoisotopic_mass,
    y = intensity,
    color = genotype
  )
) +
  geom_quasirandom(dodge.width=0.5, size = 0.2) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_color_bright()

# Plot - Violin plot of monoisotopic mass versus intensity in WT and JMJD6KO genotype
ggplot(
  data = m.brd4_hydroxylation_site_di,
  aes(
    x = monoisotopic_mass,
    y = intensity,
    fill = genotype
  )
) +
  geom_violin(scale='width') +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_fill_bright()


```

```{r plot_with_only_diagnostic_ions, fig.width = 6, fig.height = 4}

selected.ms <- c("100.0762", "156.1025")

# Count the number of PSMs for each selected monoisotopic mass 
psm.count.by.di <- m.brd4_hydroxylation_site_di[monoisotopic_mass %in% c(selected.ms)][, .N, by = list(
  monoisotopic_mass, genotype, diagnostic_ion
)]

print("Ratio of PSM with diagnostic ion: 100.0762")
psm.count.by.di[monoisotopic_mass == "100.0762" & genotype == "WT" & diagnostic_ion == TRUE, N] /
  psm.count.by.di[monoisotopic_mass == "100.0762" & genotype == "WT", sum(N)]

print("Ratio of PSM with diagnostic ion: 156.1025")
psm.count.by.di[monoisotopic_mass == "156.1025" & genotype == "WT" & diagnostic_ion == TRUE, N] /
  psm.count.by.di[monoisotopic_mass == "156.1025" & genotype == "WT", sum(N)]

# Perform Fisher's test for each selected monoisotopic mass 
for(mi.ms in selected.ms){
  print(mi.ms)
  
  psm.count.by.di[monoisotopic_mass == mi.ms] %>% # subsets psm counts for current mi.ms 
    dcast(genotype ~ diagnostic_ion, value.var = "N") %>% # convert long to wide format data table
    setnafill(cols = c("FALSE", "TRUE"), fill = 0)  %>% # Fill NA values as 0 
    {as.matrix(.[, c("FALSE", "TRUE"), with = FALSE])} %>% # convert into data matrix 
    fisher.test %>% print # perform Fisher's test and print results 
}

# Plot - Violin plot of selected monoisotopic mass versus intensity in WT and JMJD6KO genotype
ggplot(
  data = m.brd4_hydroxylation_site_di[monoisotopic_mass %in% c(selected.ms)],
  aes(
    x = monoisotopic_mass,
    y = intensity,
    fill = genotype
  )
) +
  geom_violin(scale='width') +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_fill_bright()

m.brd4_hydroxylation_site_di[monoisotopic_mass %in% c(selected.ms)][, .N, by = list(monoisotopic_mass, genotype)]

# Plot - Bar chart of proportion of PSMs with diagnostic ion 
ggplot(
  data = m.brd4_hydroxylation_site_di[monoisotopic_mass %in% c(selected.ms)],
  aes(
    x = genotype,
    fill = diagnostic_ion
  )
) +
  geom_bar(position = "fill") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  facet_grid(~ monoisotopic_mass) +
  scale_fill_manual(values = c("TRUE" = "#A50026", "FALSE" = "#DDDDDD")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ylab("Proportion of PSMs with diagnostic ions")

```

# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
