---
title: "p2-05 · Plot diagnostic-ion analysis"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Evaluate how diagnostic ions improve hydroxylation analysis —
precision, WT / KO overlap, and comparison with PNAS 2022.

**Inputs:** stoichiometry tables from p2-01; PNAS2022 reference data.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01, p2-04. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")

```

# Import data

Loads the diagnostic-ion stoichiometry tables and protein feature data.

```{r import_basic_data, warning=FALSE}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")


#----------------
# MQ_DI
#----------------

# Load sample run info data (MQ DI with no waterloss)
MQ_DI_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_run_info[, sample_id := 1:.N]

```

# Helper functions

Defines the helper functions used in this script (e.g. reading stoichiometry tables).

# Diagnostic ions and precision

Assesses how requiring a diagnostic ion changes the precision of hydroxylation calls.

```{r precision}

# Read stoichiometry data for MQ_DI (change from the older version: a potential diagnostic ions with water loss is no longer considered)
MQ_DI_stoic_dt <- lapply(
  MQ_DI_run_info[, prefix],
  read_stoic_data,
  pre_prefix = "DI_", 
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI")
) %>% rbindlist

# fwrite(MQ_DI_stoic_dt[ptm != ""], file = file.path(results.dir, "p2-analysis-setting", "all_hydroxylysine.csv"))

MQ_DI_stoic_dt <- MQ_DI_stoic_dt[aa == "K"]
MQ_DI_stoic_dt[, genotype := str_split_fixed(sample_name, "_", n = 3)[, 2] %>% factor(levels = c("HeLaWT", "HeLaJMJD6KO"))]

# Extract the sites with diagnostic ions
DI_sites <- MQ_DI_stoic_dt[diagnostic_peak == "+"][!duplicated(paste(protein_accession, aa_pos)), .(protein_accession, aa_pos)]

MQ_DI_stoic_dt[, DI_site :=
  paste(protein_accession, aa_pos) %in% DI_sites[, paste(protein_accession, aa_pos)] 
]

# Load protein feature data 
protein.feature.dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/all_protein_feature_per_position.csv"))
protein.feature.dt <- protein.feature.dt[residue == "K"]
setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos")) # change column names 

protein.feature.dt[, `:=`( #create a new column 'met_within_2'
  met_within_2 = case_when(
    nchar(Window) == 11 & grepl("M", substr(Window, start = 4, stop = 8)) ~ "Yes", # if M is in the middle, Yes
    nchar(Window) == 11 ~ "No", # if M is not in the sequence, then No 
    TRUE ~ "edge" # if M is in the sequence but not in the middle, then edge
  )
)]

# Merge data tables by protein_accession and aa_pos (MQ_DI)
wt_vs_KO_MQ_DI_stoic_dt <- copy(MQ_DI_stoic_dt)

wt_vs_KO_MQ_DI_stoic_dt <- merge(
  wt_vs_KO_MQ_DI_stoic_dt,
  protein.feature.dt,
  by = c("protein_accession", "aa_pos")
)

# Only analyse the sites with a coverage with 2 PSMs in both JMJD6 WT and KO data
wt_data_coverage <- wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaWT"][
  , list(total_wt_sum_psm_mapped = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos)]
KO_data_coverage <- wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaJMJD6KO"][
  , list(total_ko_sum_psm_mapped = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos)]

all_data_coverage <- merge(
  wt_data_coverage,
  KO_data_coverage,
  by = c("protein_accession", "aa_pos")
)

wt_vs_KO_MQ_DI_stoic_dt <- wt_vs_KO_MQ_DI_stoic_dt[
  paste(protein_accession, aa_pos) %in% 
    # Keep only sites covered (>= 1 PSM) in both WT and KO, so presence/absence is comparable.
    all_data_coverage[total_wt_sum_psm_mapped > 0 & total_ko_sum_psm_mapped > 0][, paste(protein_accession, aa_pos)]
]

# The # of sites analysed
wt_vs_KO_MQ_DI_stoic_dt[!duplicated(paste(genotype, protein_accession, aa_pos))][, .N, by = genotype]

koh_wt_vs_KO_MQ_DI_stoic_dt <- wt_vs_KO_MQ_DI_stoic_dt[ptm == "[Oxidation (K)]"]
koh_wt_vs_KO_MQ_DI_stoic_dt <- koh_wt_vs_KO_MQ_DI_stoic_dt[order(-DI_site)][!duplicated(paste(genotype, protein_accession, aa_pos))]

g1 <- ggplot(
  data = koh_wt_vs_KO_MQ_DI_stoic_dt[met_within_2 != "edge"],
  aes(
    x = genotype,
    fill = met_within_2
  )
) +
  geom_bar() +  
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g2 <- ggplot(
  data = koh_wt_vs_KO_MQ_DI_stoic_dt[met_within_2 != "edge" & DI_site == TRUE],
  aes(
    x = genotype,
    fill = met_within_2
  )
) +
  geom_bar() +  
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

# Number of sites for plotting g2
koh_wt_vs_KO_MQ_DI_stoic_dt[met_within_2 != "edge" & DI_site == TRUE][
  , .N, by = list(met_within_2, genotype)
]

library("patchwork")

g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = "bottom") 

```

# WT vs KO site overlap

Compares hydroxylysine sites detected in JMJD6 WT vs KO cells (overlap and methionine proximity).

```{r overlap_wt_ko}

koh_per_site <- dcast(
  koh_wt_vs_KO_MQ_DI_stoic_dt,
  protein_accession + aa_pos + DI_site + met_within_2 ~ genotype,
  fun.aggregate = length 
)

koh_per_site[, `:=`(Hyl_found_in = case_when(
  HeLaWT == 1 & HeLaJMJD6KO == 1 ~ "both",
  HeLaWT == 1 ~ "WT",
  HeLaJMJD6KO == 1 ~ "JMJD6KO",
  TRUE ~ "not_found"
) %>% factor(levels = c("WT", "JMJD6KO", "both", "not_found")))]

koh_per_site <- koh_per_site[met_within_2 != "edge"]

koh_per_site_count <- koh_per_site[, .N, by = list(Hyl_found_in, met_within_2, DI_site)]

g1 <- ggplot(koh_per_site_count, aes(x = Hyl_found_in, y = N, fill = met_within_2)) +
  geom_bar(stat = "identity") +  
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g2 <- ggplot(koh_per_site_count[DI_site == TRUE], aes(x = Hyl_found_in, y = N, fill = met_within_2)) +
  geom_bar(stat = "identity") +  
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")
  
merge(
  koh_wt_vs_KO_MQ_DI_stoic_dt[!duplicated(paste(protein_accession, gene_name)), .(protein_accession, gene_name)],
  koh_per_site[DI_site == TRUE & Hyl_found_in == "both"]
)

# Statistical significance of methionine enrichment
# All sites, found in both
all.K.m.count <- rbind(
  copy(protein.feature.dt[met_within_2 != "edge", .N, by = met_within_2])[, data_type := "all_K"],
  copy(koh_per_site_count[, list(N = sum(N)), by = met_within_2])[, data_type := "koh"]
) %>%
  dcast(
    met_within_2 ~ data_type,
    value.var = "N"
  )

fisher.test(all.K.m.count[, .(all_K, koh)], alternative = "two.sided")

library("eulerr")
library("RColorBrewer")

vennlist <-  list(
  WT_DI_site = koh_wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaWT", paste(protein_accession, aa_pos)],
  KO_DI_site = koh_wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaJMJD6KO", paste(protein_accession, aa_pos)]
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(3, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated Sites Overlap (All)")

vennlist <-  list(
  WT_DI_site = koh_wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaWT" & DI_site == TRUE, paste(protein_accession, aa_pos)],
  KO_DI_site = koh_wt_vs_KO_MQ_DI_stoic_dt[genotype == "HeLaJMJD6KO" & DI_site == TRUE, paste(protein_accession, aa_pos)]
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(3, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated Sites Overlap (with DI)")

```

# Comparison with PNAS 2022

Benchmarks the diagnostic-ion-based sites against the manually curated sites from the PNAS 2022 study.

```{r pnas2022_comparison_MQ_DI}

# Define file path to load 'long_K_stoichiometry_data'  
pnas2022.stoic.dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/long_K_stoichiometry_data.csv"))

#change the column names
setnames(pnas2022.stoic.dt, old = c("uniprot_id", "position", "residue"), new = c("protein_accession", "aa_pos", "aa")) 

# Create new column 'accession_position'
pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos) # paste protein_accession and aa_pos together '_'
)]

# Check how many hydroxylation sites that were reported by PNAS2022 are identified by the new workflow
MQ_DI_stoic_dt[, `:=`( 
  curated_oxK_site = 
    paste0(protein_accession, "_", aa_pos) %in%
    pnas2022.stoic.dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)] 
)]

MQ_DI_d_hydroxyK_dt <- MQ_DI_stoic_dt[ptm == "[Oxidation (K)]" & genotype == "HeLaWT"][order(genotype, DI_site)][!duplicated(paste(protein_accession, aa_pos))]

# Convert the data table into table and sum the number of diagnostic peak and curated oxK sites
hyl_precision <- MQ_DI_d_hydroxyK_dt[, table(DI_site, curated_oxK_site) %>% addmargins] %>% data.table
hyl_precision <- hyl_precision[DI_site %in% c("Sum", "TRUE") & curated_oxK_site != "Sum"]

hyl_precision

ggplot(
  hyl_precision,
  aes(
    x = DI_site,
    y = N,
    fill = curated_oxK_site
  )
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("TRUE" = "coral2", "FALSE" = "#BBBBBB")) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(aspect.ratio = 3.5)

dcast(
  hyl_precision,
  DI_site ~ curated_oxK_site,
  value.var = "N"
) %>%
  {fisher.test(.[, c("FALSE", "TRUE"), with = FALSE], alternative = "two.sided")}

```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
