---
title: "p2-03 · Compare miscleavage settings"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Compare PSM coverage of lysine-rich domains across miscleavage
settings (m2 / m5 / m7).

**Inputs:** stoichiometry tables from p2-01 (`results/p2-analysis-setting/`); sample matrix.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# PSM coverage by miscleavage

Compares PSM coverage of lysine-rich domains between the m2, m5 and m7 miscleavage search settings.

```{r comparison_PNAS2022_PTM_PSM}

pnas2022.stoic.dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/long_K_stoichiometry_data.csv"))

setnames(
  pnas2022.stoic.dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name")
)

pnas2022.stoic.dt[, `:=`(
  sum_psm_mapped = total_n_feature_K + total_n_feature_oxK
)]

pnas2022.stoic.dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)],
  pnas2022.stoic.dt,
  by = "protein_accession"
)

pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos),
  ptm = "[Oxidation (K)]",
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>%
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)}
)]


# Load library 
library(ggplot2)

```

```{r comparison_MQ_Std_PSM_PTM}

# Load sample run info data (MQ standard)
MQ_Std_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_Std" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_Std_run_info[, sample_id := 1:.N]

## Read stoichiometry data for MQ_Std
MQ_Std_stoic_dt <- lapply(
  MQ_Std_run_info[, prefix],
  read_stoic_data,
  pre_prefix = "", 
  post_fix = "",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS")
) %>% rbindlist



```


```{r MQ_Std_PSM_miscleavages}

## MQ_DI PSM coverage according to miscleavages (m2, m5, m7)

# m2
MQ_PSM_WT <- plot_psm_coverage(
  MQ_Std_stoic_dt[sample_name == "JQ1_HeLaWT_derivatised" & condition == "data-A_trp_m2_v2_def"], accession = "O60885", plot_range = c(531, 581), all.protein.bs, sample_colors = NA) 
m2.g <- MQ_PSM_WT$coverage_plot_g + coord_cartesian(ylim = c(0, 350)) +
    labs(title = "BRD4 PSM miscleavages m2")

# m5
MQ_PSM_WT <- plot_psm_coverage(
  MQ_Std_stoic_dt[sample_name == "JQ1_HeLaWT_derivatised" & condition == "data-A_trp_m5_v5_def"], accession = "O60885", plot_range = c(531, 581), all.protein.bs, sample_colors = NA)
m5.g <- MQ_PSM_WT$coverage_plot_g + coord_cartesian(ylim = c(0, 350))

# m7
MQ_PSM_WT <- plot_psm_coverage(
  MQ_Std_stoic_dt[sample_name == "JQ1_HeLaWT_derivatised" & condition == "data-A_trp_m7_v7_def"], accession = "O60885", plot_range = c(531, 581), all.protein.bs, sample_colors = NA)
m7.g <- MQ_PSM_WT$coverage_plot_g + coord_cartesian(ylim = c(0, 350))

library("patchwork")
m2.g / m5.g / m7.g

plot_ptm_stoichiometry(MQ_Std_stoic_dt[sample_name == "JQ1_HeLaWT_derivatised" & condition == "data-A_trp_m7_v7_def"], accession = "O60885", plot_range = c(531, 581), all.protein.bs)


```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```

