---
title: "p2-06 · Analyse MS_KR_1 stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Analyse MS_KR_1 lysine-hydroxylation stoichiometry — oxygen
sensitivity (hypoxia vs normoxia) across BRD2/3/4, including JMJD6 re-expression.
Stoichiometry itself is computed in p2-01; this script reads those tables.

**Inputs:** stoichiometry tables from `results/p2-analysis-setting/MS_KR_1/`
(written by p2-01); UniProt reference proteome (`reference_fasta`).

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Per-site stoichiometry across conditions

Reads the MS_KR_1 stoichiometry table and plots site-specific hydroxylation along
BRD4 (O60885), BRD3 (Q15059) and BRD2 (P25440) across the four conditions —
parental WT, induced cells without doxycycline, and JMJD6 re-expression (+Dox)
under normoxia and 0.1% O2. The `diagnostic_peak` flag marks sites confirmed by the
hydroxylysine immonium ion.

```{r plot_brd_stoichiometry}

# Read stoichiometry data for MS_KR_1
ms_kr1_stoic_dt <- read_stoic_data(
  prefix     = "MS_KR_1_",
  pre_prefix = "",
  post_fix   = "_DI",
  dir_path   = file.path(results.dir, "p2-analysis-setting", "MS_KR_1")
)

# Diagnostic-ion flag and an ordered sample factor (all four conditions)
ms_kr1_stoic_dt[, is_diagnostic_peak := diagnostic_peak == "+"]

kr1_conditions <- c(
  "HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA",
  "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"
)
ms_kr1_stoic_dt[, sample_name := factor(sample_name, levels = kr1_conditions)]

# Per-site stoichiometry for one accession across all four conditions
plot_brd_stoic <- function(accession, plot_range) {
  plot_ptm_stoichiometry(
    ms_kr1_stoic_dt[sample_name %in% kr1_conditions],
    accession = accession, plot_range = plot_range, all.protein.bs, sample_colors = NA
  )
}

plot_brd_stoic("O60885", c(531, 581))  # BRD4
plot_brd_stoic("Q15059", c(483, 533))  # BRD3
plot_brd_stoic("P25440", c(540, 590))  # BRD2

```

# Normoxia vs hypoxia (JMJD6 re-expression)

Restricts the data to the two JMJD6 re-expression conditions (+Dox: normoxia vs
0.1% O2), contrasts per-site stoichiometry between them, and tests whether the
normoxia − hypoxia difference differs between BRD2/3/4 (ANOVA and boxplot).

```{r contrast_normoxia_hypoxia}

# Keep only the JMJD6 re-expression (+Dox) conditions; the parental WT and the
# un-induced (no-Dox) samples are dropped because this comparison isolates the
# oxygen effect on JMJD6-dependent hydroxylation, which requires JMJD6 expression.
ms_kr1_stoic_dt <- droplevels(
  ms_kr1_stoic_dt[!sample_name %in% c("HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA")]
)
ms_kr1_stoic_dt[, sample_name :=
  factor(sample_name, levels = c("HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"))
]

# Restrict to BRD2/3/4 and label the oxygen condition
ms_kr1_stoic_dt <- ms_kr1_stoic_dt[gene_name %in% c("BRD2", "BRD3", "BRD4")]
ms_kr1_stoic_dt[, oxygen_levels := factor(
  fcase(
    grepl("^HeLaiJMJD6_Dox_N_NA", sample_name),       "HeLaNormoxia_JMJD6KO_reexp",
    grepl("^HeLaiJMJD6_Dox_01O224h_NA", sample_name), "HeLaHypoxia",
    default = NA_character_
  ),
  levels = c("HeLaNormoxia_JMJD6KO_reexp", "HeLaHypoxia")
)]

# Derive the oxygen status (Normoxia / Hypoxia) used to group the contrast.
# The per-sample zero-fill keys on the finer 'oxygen_levels' label.
ms_kr1_stoic_dt[, oxygen_stat := str_extract(oxygen_levels, "(?<=HeLa)(Normoxia|Hypoxia)")]

# Zero-fill non-hydroxylated K and reshape to one column per oxygen status.
# (Shared helper contrast_hydroxylation() lives in R/functions/2-useful_functions.R.)
ms_kr1_hydroxy_dt <- contrast_hydroxylation(
  ms_kr1_stoic_dt,
  group_col  = "oxygen_stat",
  sample_col = "oxygen_levels"
)

# Per-site diagnostic-peak status (most confident row per site)
di_info_dt <- ms_kr1_stoic_dt[
  , .(protein_accession, aa_pos, diagnostic_peak)
][
  order(protein_accession, aa_pos, -diagnostic_peak)
][
  !duplicated(paste(protein_accession, aa_pos))
]

ms_kr1_hydroxy_dt <- merge(ms_kr1_hydroxy_dt, di_info_dt, by = c("protein_accession", "aa_pos"))
ms_kr1_hydroxy_dt[, is_diagnostic_peak := diagnostic_peak == "+"]

# Keep only sites whose hydroxylation is confirmed by the diagnostic (immonium) ion
ms_kr1_hydroxy_dt <- ms_kr1_hydroxy_dt[is_diagnostic_peak == TRUE]

# Normoxia - hypoxia stoichiometry difference, per site
ms_kr1_hydroxy_dt[, Stoic_diff := Normoxia - Hypoxia]
ms_kr1_hydroxy_dt[, gene_name := factor(gene_name)]

```

```{r anova_stoic_diff}

# Analyse only sites with appreciable normoxic hydroxylation (stoichiometry >= 0.1).
# little signal for a normoxia-hypoxia difference.
ms_kr1_hydroxy_dt <- subset(ms_kr1_hydroxy_dt, Normoxia >= 0.1)

ms_kr1_hydroxy_dt[, table(gene_name)]

# ANOVA: does the normoxia - hypoxia difference vary between BRD2/3/4?
anova_ms_kr1 <- aov(Stoic_diff ~ gene_name, data = ms_kr1_hydroxy_dt)
summary(anova_ms_kr1)

```


```{r plot_stoic_diff}

# Stoichiometry difference (normoxia - hypoxia) between BRD2, 3 and 4
ggplot(
  ms_kr1_hydroxy_dt[is_diagnostic_peak == TRUE],
  aes(x = gene_name, y = Stoic_diff)
) +
  stat_boxplot(geom = "errorbar", width = 0.25) +
  geom_boxplot() +
  geom_point() +
  labs(title = "Stoichiometry difference (Normoxia 0.1 between BRD2, 3 and 4") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) +
  theme(legend.position = "none")

```

# Session information

```{r session_info}

sessioninfo::session_info()

```
