---
title: "p2-06 · Analyse MS_KR_1 stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Analyse MS_KR_1 lysine-hydroxylation stoichiometry — oxygen
sensitivity (hypoxia vs normoxia) across BRD2/3/4, including JMJD6 re-expression.
Stoichiometry itself is computed in p2-01; this script reads those tables.

**Inputs:** stoichiometry tables from `results/p2-analysis-setting/MS_KR_1/`
(written by p2-01); UniProt reference proteome (`reference_fasta`).

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")

```

# Per-site stoichiometry across conditions

Reads the MS_KR_1 stoichiometry table and plots site-specific hydroxylation along
BRD4 (O60885), BRD3 (Q15059) and BRD2 (P25440) across the four conditions —
parental WT, induced cells without doxycycline, and JMJD6 re-expression (+Dox)
under normoxia and 0.1% O2. The `diagnostic_peak` flag marks sites confirmed by the
hydroxylysine immonium ion.

```{r hydroxylation_hypoxia_vs_normoxia}



## Read stoichiometry data for MS_KR1 data
 MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1"))


#Filter data table with presence of diagnostic_peak and WT stoichiometry values
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

# categorize data according to sample names 
MS_KR1_stoic_dt[, `:=`(
  sample_name =  
    factor(sample_name, levels = c("HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"))
)]

# BRD4 
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "O60885", plot_range = c(531, 581), all.protein.bs, sample_colors = NA,
)

# BRD3
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "Q15059", plot_range = c(483, 533), all.protein.bs, sample_colors = NA
)

# BRD2
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "P25440", plot_range = c(540, 590), all.protein.bs, sample_colors = NA
)
```

# Normoxia vs hypoxia (JMJD6 re-expression)

Restricts the data to the two JMJD6 re-expression conditions (+Dox: normoxia vs
0.1% O2), contrasts per-site stoichiometry between them, and tests whether the
normoxia − hypoxia difference differs between BRD2/3/4 (ANOVA and boxplot).

```{r JMJD6_reexpression_vs_hypoxia}


# Keep only the JMJD6 re-expression (+Dox) conditions; the parental WT and the
# un-induced (no-Dox) samples are dropped because this comparison isolates the
# oxygen effect on JMJD6-dependent hydroxylation, which requires JMJD6 expression.
MS_KR1_stoic_dt <-  droplevels(MS_KR1_stoic_dt[!sample_name %in% c("HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA")])


# categorize data according to sample names 
MS_KR1_stoic_dt[, `:=`(
  sample_name =  
    factor(sample_name, levels = c("HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"))
)]


# Filter MS_KR_1 data to retain only BRD2, 3, 4 data
MS_KR1_stoic_dt <-  MS_KR1_stoic_dt[gene_name %in% c("BRD2", "BRD3", "BRD4")]

MS_KR1_stoic_dt[, `:=`(
  oxygen_levels =
    factor(
      fcase(
        grepl("^HeLaiJMJD6_Dox_N_NA", sample_name),
        "HeLaNormoxia_JMJD6KO_reexp",
        grepl("^HeLaiJMJD6_Dox_01O224h_NA", sample_name),
        "HeLaHypoxia",
        default = NA_character_
      ),
      levels = c("HeLaNormoxia_JMJD6KO_reexp", "HeLaHypoxia")
    )
)]

# Define function - 'contrast_hydroxylation_by_oxygen_stat'
contrast_hydroxylation_by_oxygen_stat <- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    oxygen_stat = str_extract(oxygen_levels, "(?<=HeLa)(Normoxia|Hypoxia)"), # new column - oxygen status, extract Normoxia or Hypoxia from oxygen_levels column
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(oxygen_levels, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Keep sites with sparser coverage out: stoichiometry estimates are unreliable
  # unless a position has > 2 PSMs of coverage.
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, oxygen_stat)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ oxygen_stat, # values in oxygen_stat become into separate columns (Normoxia & Hypoxia)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

# Use contrast function on 'MS_KR1_stoic_dt' to set NA to 0, only if the sample contains an [oxidation K] 
MS_KR_1_hydroxy_dt <-  contrast_hydroxylation_by_oxygen_stat(MS_KR1_stoic_dt)

# Generate data table with WT and JMJD6KO with diagnostic peaks
DI_info <- MS_KR1_stoic_dt[
  , .(protein_accession, aa_pos, diagnostic_peak) # filter these columns and create a separate data table 
][
  order(protein_accession, aa_pos, -diagnostic_peak) # order the data based on the highest diagnostic peak 
][                                              
  !duplicated(paste(protein_accession, aa_pos)) # Keep only the first row for each unique protein and amino acid position,
  # removing duplicates while preserving the row with the highest diagnostic_peak (from previous sorting)
]

# merge the modified data table 'DI_info' with the metadata 'MS_KR_1_hydroxy_dt'
MS_KR_1_hydroxy_dt <- merge(MS_KR_1_hydroxy_dt, DI_info, by = c("protein_accession", "aa_pos"))

#Filter data table with presence of diagnostic_peak and WT stoichiometry values
MS_KR_1_hydroxy_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

# Keep only sites whose hydroxylation is confirmed by the diagnostic (immonium) ion.
MS_KR_1_hydroxy_dt <- MS_KR_1_hydroxy_dt[is_diagnostic_peak == TRUE]

# Calculate the difference in stoichiometric values of JMJD6 re expression and Hypoxia
MS_KR_1_hydroxy_dt <- MS_KR_1_hydroxy_dt[, `:=`(
  Stoic_diff = (Normoxia - Hypoxia)
)]

# Factor the gene names
MS_KR_1_hydroxy_dt[, `:=`(
  gene_name =  
    factor(gene_name)
)]

```

```{r ANOVA_ttest}

# Analyse only sites with appreciable normoxic hydroxylation (stoichiometry >= 0.1).
# NOTE: please confirm the rationale for the 0.1 cut-off — sites near zero carry
# little signal for a normoxia-hypoxia difference.
MS_KR_1_hydroxy_dt <- subset(MS_KR_1_hydroxy_dt, Normoxia >= 0.1)

MS_KR_1_hydroxy_dt[, table(gene_name)]

# ANOVA
anova_MS_KR_1 <- aov(Stoic_diff ~ gene_name, data = MS_KR_1_hydroxy_dt)

summary(anova_MS_KR_1)

```


```{r boxplots_stoic_diff_ratio}

# Plot - Stoic difference between BRD2, 3 and 4
ggplot(
  data = MS_KR_1_hydroxy_dt[is_diagnostic_peak == TRUE],
  aes(
    x = gene_name,
    y = Stoic_diff
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point() +
  labs(title = "Stoichiometry difference (Normoxia 0.1 between BRD2, 3 and 4") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(legend.position="none")


```

# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
