---
title: "p2-07 · Plot site-specific stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Visualise site-specific hydroxylation stoichiometry in BRD2/3/4
across O2% and dox induction (via `plot_ptm_stoichiometry`).

**Inputs:** MS_KR_1 / MS_SS / PNAS stoichiometry tables from p2-01; reference proteome.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")

```

# Functions and data preprocessing

Loads and preprocesses the MS_KR_1 / MS_SS / PNAS stoichiometry tables (via load_stoichiometry_datasets()).

```{r load_PNAS_MS_KR_1_MS_SS_data}

# Load and preprocess the MS_KR_1, MS_SS and PNAS stoichiometry tables
# (see load_stoichiometry_datasets() in R/functions/2-useful_functions.R).
.stoic_data       <- load_stoichiometry_datasets(results.dir, data.dir)
MS_KR1_stoic_dt   <- .stoic_data$MS_KR1_stoic_dt
MS_SS_stoic_dt    <- .stoic_data$MS_SS_stoic_dt
pnas2022_stoic_dt <- .stoic_data$pnas2022_stoic_dt
pnas2022_dt       <- .stoic_data$pnas2022_dt

```


```{r data_preprocessing}

# Remove BRD2/3 samples reporting BRD4 stoichiometry and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_stoic_dt <- MS_SS_stoic_dt[
  # Drop cross-protein false positives: BRD4 signal reported in BRD2/3 samples
  # (and vice versa), e.g. from shared / homologous peptides.
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_PNAS_dt <- rbindlist(list(MS_SS_stoic_dt,
                         MS_KR1_stoic_dt,
                         pnas2022_stoic_dt),
                         use.names = TRUE)

MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Column "sample group" created to simplify sample names
MS_SS_KR_PNAS_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

# Normoxia time course

Plots site-specific stoichiometry under normoxia across the JMJD6 re-expression time course for BRD2/3/4.

```{r stoichiometry_BRD2_at_21pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented 
# in the PNAS2022 paper. 

# subset data to specific BRD2 protein
MS_SS_KR_PNAS_BRD2_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            # Select the appropriate minus-Dox (un-induced) control and exclude the JQ1/PNAS
                                            # derivatised samples from these plots.
                                            # NOTE: confirm the choice of minus-Dox control.
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
                                            gene_name == "BRD2"]


# reorder sample names
MS_SS_KR_PNAS_BRD2_dt <-  MS_SS_KR_PNAS_BRD2_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD3_at_21pc}

# subset data to specific BRD3 protein
MS_SS_KR_PNAS_BRD3_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD3"]



MS_SS_KR_PNAS_BRD3_dt <-  MS_SS_KR_PNAS_BRD3_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD4_at_21pc}

# subset data to specific BRD4 protein
MS_SS_KR_PNAS_BRD4_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD4", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD4"]


# reorder sample names
MS_SS_KR_PNAS_BRD4_dt <-  MS_SS_KR_PNAS_BRD4_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD4",
             "8h_21pc_BRD4",
             "18h_21pc_BRD4",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```

# Normoxia vs hypoxia

Plots site-specific stoichiometry comparing normoxia and hypoxia for BRD2/3/4.

```{r stoichiometry_BRD2_at_hypoxia_pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented in the PNAS2022 paper. 

# HeLaWT_NA_N_NA - WT_Inf_21pc_KR
# JQ1_HeLaWT_derivatised - WT_Inf_21pc_PNAS
# JQ1_HeLaJMJD6KO_derivatised - iJ6_Inf_21pc_PNAS
# HeLaiJMJD6_Dox_N_NA - iJ6_24h_21pc_KR
# 18h_4pc_BRD23 - iJ6_18h_4pc_SS
# 18h_1pc_BRD23 - iJ6_18h_1pc_SS
# HeLaiJMJD6_Dox_01O224h_NA - iJ6_24h_01pc_KR
# minusDox_BRD23 - iJ6_0h_21pc_SS
# HeLaiJMJD6_noDox_N_NA - iJ6_0h_21pc_KR

#------
# BRD2
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD2_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD2"]

# reorder sample names
MS_SS_KR_PNAS_BRD2_O2pc_dt <-  MS_SS_KR_PNAS_BRD2_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_O2pc_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD3_at_hypoxia_pc}

#------
# BRD3
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD3_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD3"]


# reorder sample names
MS_SS_KR_PNAS_BRD3_O2pc_dt <-  MS_SS_KR_PNAS_BRD3_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]


# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_O2pc_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD4_at_hypoxia_pc}

#------
# BRD4
#------

# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD4_O2pc_dt <- MS_SS_KR_PNAS_dt[  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD4", "18h_21pc_BRD4","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD4"]


#iJ6_0h_21pc_SS - minusDox_BRD23
# HeLaiJMJD6_noDox_N_NA

# reorder sample names
MS_SS_KR_PNAS_BRD4_O2pc_dt <-  MS_SS_KR_PNAS_BRD4_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD4",
             "18h_1pc_BRD4",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_O2pc_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```

# Session information

```{r session_Info}

sessioninfo::session_info()

```
