---
title: "p2-04 · Identify diagnostic ions"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Identify immonium / diagnostic ions that mark hydroxylysine,
using FragPipe diagnostic-ion output and a WT vs JMJD6-KO contrast.

**Inputs:** FragPipe `diagnosticIons.tsv` + `psm.tsv` (`data/FP_diagnostic_ion_search/`); data-A sample info; reference proteome.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** p2-05.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

## Script-specific packages
library("janitor")
library("ggbeeswarm")

```

# Import data

Loads the FragPipe diagnostic-ion and PSM tables, and the data-A sample info.

```{r load_inputs}

# FragPipe diagnostic-ion table (data-A); data.dir is defined in _setup.R
diagnostic_ion_dt <- file.path(
  data.dir, "FP_diagnostic_ion_search/fragpipe_dataset-A/dataset01.diagnosticIons.tsv"
) %>%
  fread %>%
  clean_names

# FragPipe PSM table (data-A)
fragpipe_psm_dt <- file.path(
  data.dir, "FP_diagnostic_ion_search/fragpipe_dataset-A/psm.tsv"
) %>%
  fread %>%
  clean_names

# data-A sample info (MaxQuant standard search output root)
mq_std_data_dir <- file.path(data.dir, "MQ_Std/PNAS2022")
data_a_sample_info_dt <- fread(file.path(
  mq_std_data_dir, "data-A", "MS_dataset_overview_PXD031221_data-A.csv"
))

```

# Identify diagnostic ions

Identifies which immonium / diagnostic-ion masses mark hydroxylysine, by contrasting their intensity between JMJD6 WT and KO cells at BRD4 sites.

```{r prep_di_data}

# Attach protein / position info from the PSM table, keyed by spectrum
diagnostic_ion_dt <- merge(
  fragpipe_psm_dt[, .(spectrum, protein_id, gene, protein_start, protein_end)],
  diagnostic_ion_dt,
  by = "spectrum"
)

# Raw file name = part of 'spectrum' before the first '.'
diagnostic_ion_dt[, file_name := str_split_fixed(spectrum, "\\.", n = 2)[, 1]]

# Attach the data-A sample info, keyed by raw file name
diagnostic_ion_dt <- merge(
  data_a_sample_info_dt,
  diagnostic_ion_dt,
  by = "file_name"
)

# Restrict to the BRD4 lysine-rich region (around residues 535-555).
brd4_di_dt <- diagnostic_ion_dt[
  gene == "BRD4" & (555 >= protein_start & 535 <= protein_end)
]

# Reshape the per-ion intensity columns to long format
brd4_di_long_dt <- melt(
  brd4_di_dt,
  measure.vars = grep("intensity$", colnames(brd4_di_dt), value = TRUE),
  value.name   = "intensity"
)

# Derive the monoisotopic mass (from the column name), a diagnostic-ion flag,
# and an ordered genotype factor
brd4_di_long_dt[, `:=`(
  monoisotopic_mass = variable %>%
    str_replace_all("ox_", "") %>%          # drop "ox_" prefix
    str_replace_all("_intensity", "") %>%   # drop "_intensity" suffix
    str_replace_all("_", ".") %>%           # remaining "_" -> "."
    factor(levels = c(
      "101.1079", #Lysine	K			immonium ion	C5 H13 N2 +
      "100.0762", #Hydroxylation (K)	K	O	15.9949	cyclic immonium ion	C5 H10 N O+
      "82.0657", #Hydroxylation (K)	K	O	15.9949	diagnostic ion (water loss)	C5 H8 N+
      "117.1028", #Hydroxylation (K)	K	O	15.9949 immonium ion	C5 H13 O N2 + calculated
      "117.0658", #Hydroxylation (K)	K	O	15.9949	immonium ion	C5 H13 O N2 + (from citation)
      "138.0919", #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937	diagnostic ion (water loss)	C8 H12 N1 O+
      "145.0977", #Hydroxylation (K)	K	O	15.9949	(intact, water loss)	C6H13N2O2+
      "156.1025", #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937	cyclic immonium ion	C8 H14 N1 O2+
      "173.1290" #Hydroxylation-Propionylation (K)	K	C3 H4 O2	72.02112937 immonium ion	C8 H17 O2 N2 +
    )),
  diagnostic_ion = intensity > 0,
  genotype = factor(genotype, levels = c("WT", "JMJD6KO"))
)]

```



```{r plot_selected_di, fig.width = 4, fig.height = 5}

selected_ms <- c("100.0762", "156.1025")

# PSM counts for each selected monoisotopic mass
psm_count_by_di_dt <- brd4_di_long_dt[monoisotopic_mass %in% selected_ms][
  , .N, by = list(monoisotopic_mass, genotype, diagnostic_ion)
]

print("Ratio of PSM with diagnostic ion: 100.0762")
psm_count_by_di_dt[monoisotopic_mass == "100.0762" & genotype == "WT" & diagnostic_ion == TRUE, N] /
  psm_count_by_di_dt[monoisotopic_mass == "100.0762" & genotype == "WT", sum(N)]

print("Ratio of PSM with diagnostic ion: 156.1025")
psm_count_by_di_dt[monoisotopic_mass == "156.1025" & genotype == "WT" & diagnostic_ion == TRUE, N] /
  psm_count_by_di_dt[monoisotopic_mass == "156.1025" & genotype == "WT", sum(N)]

# Fisher's test of diagnostic-ion presence by genotype, per selected mass
for (ms in selected_ms) {
  print(ms)

  psm_count_by_di_dt[monoisotopic_mass == ms] %>%      # counts for this mass
    dcast(genotype ~ diagnostic_ion, value.var = "N") %>%   # long -> wide
    setnafill(cols = c("FALSE", "TRUE"), fill = 0) %>%      # NA -> 0
    {as.matrix(.[, c("FALSE", "TRUE"), with = FALSE])} %>%  # to matrix
    fisher.test %>% print
}

brd4_di_long_dt[monoisotopic_mass %in% selected_ms][, .N, by = list(monoisotopic_mass, genotype)]

# Proportion of PSMs with a diagnostic ion, by genotype
ggplot(
  brd4_di_long_dt[monoisotopic_mass %in% selected_ms],
  aes(x = genotype, fill = diagnostic_ion)
) +
  geom_bar(position = "fill") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  facet_grid(~ monoisotopic_mass) +
  scale_fill_manual(values = c("TRUE" = "#A50026", "FALSE" = "#DDDDDD")) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  coord_cartesian(ylim = c(0, 0.25)) +
  ylab("Proportion of PSMs with diagnostic ions")

```

# Session information

```{r session_info}

sessioninfo::session_info()

```
