---
title: "p2-01 · Compute PTM stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Compute site-specific lysine-hydroxylation (PTM) stoichiometry from
MaxQuant outputs for every dataset in the study. This is the single computation
step of Part 2; all downstream scripts read the tables written here.

**Inputs:**

- MaxQuant `evidence.txt` per dataset under `data/MQ_Std/` and `data/MQ_DI/`
- sample matrix `data/analysis_setting/PXD031221_sample_matrix.xlsx` and per-dataset `sample_info`
- PTM mapping `data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv`
- UniProt reference proteome (`reference_fasta`; see Setup)

**Outputs:** stoichiometry tables under `results/p2-analysis-setting/`
(`MQ_Std_MSMS_SECPEP/`, `MQ_Std_MSMS/`, `MQ_DI/`, `MS_KR_1/`, `MS_SS/`).

**Upstream:** none. **Downstream:** p2-02 … p2-09 read these tables.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths
## (project.dir, data.dir, results.dir, reference_fasta).
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Helper functions and shared settings

Defines the two stoichiometry wrappers used throughout this script and the common
PTM-mapping file. Every dataset section below routes through one of these wrappers:

- `process_stoichiometry()` — standard MaxQuant search (no diagnostic ions); loops
  over a sample matrix and calls `calculate_stoichiometry()`.
- `process_stoichiometry2()` — diagnostic-ion search; processes a single dataset
  (locating its per-PTM site files) and calls `calculate_stoichiometry2()`.

```{r helper_functions}

# Common PTM mapping file, shared by all datasets
ptm_mapping_file <- file.path(
  project.dir,
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
)

# Automate stoichiometry calculations across standard-search MaxQuant datasets
# (e.g. MULTI-MSMS + MULTI-SECPEP and MULTI-MSMS only). `sample_dt` is the sample
# matrix (one row per run); `data_dir` is the MaxQuant output root for the dataset.
process_stoichiometry <- function(sample_dt,
                                  data_dir,
                                  out_dir,
                                  ref_protein_dt,
                                  ptm_mapping_file,
                                  selected_type = NA,
                                  prefix_label = "") {

  # Create output directory (and parents) if it does not already exist
  create.dir(out_dir)

  # For each run in the sample matrix, calculate the stoichiometry
  for (i in seq_len(nrow(sample_dt))) {
    prefix     <- sample_dt$prefix[i]                          # run prefix
    sample     <- sample_dt$data[i]                            # dataset folder
    mq_setting <- gsub("_$", "", sub("^data-[^_]+_", "", prefix))  # search setting

    # Path to this run's MaxQuant evidence file
    mq_evidence_file <- file.path(
      data_dir, sample, mq_setting, paste0(prefix, "evidence.txt")
    )

    message("Processing: ", basename(mq_evidence_file))

    # Proceed only if the evidence file exists, otherwise skip with a message
    if (file.exists(mq_evidence_file)) {

      # Corresponding sample information (metadata) file
      sample_info_file <- file.path(
        data_dir, sample,
        paste0("MS_dataset_overview_PXD031221_", sample, ".csv")
      )

      # Run main stoichiometry calculation
      calculate_stoichiometry(
        mq_evidence_data = mq_evidence_file,
        sample_info_file = sample_info_file,
        ref_protein_dt   = ref_protein_dt,
        output_prefix    = file.path(out_dir, paste0(prefix_label, prefix)),
        ptm_mapping_file = ptm_mapping_file,
        K_only           = FALSE,
        selected_type    = selected_type
      )

    } else {
      message("File does not exist: ", mq_evidence_file)
    }
    gc()  # free memory between runs
  }
}

# Automate diagnostic-ion stoichiometry calculations for a single dataset.
# Counterpart to process_stoichiometry(), but wraps calculate_stoichiometry2()
# (which additionally needs the per-PTM site files in `ptm_dir`).
process_stoichiometry2 <- function(mq_evidence_file,
                                   sample_info_file,
                                   ptm_dir,
                                   output_prefix,
                                   ref_protein_dt,
                                   ptm_mapping_file,
                                   selected_type = "MULTI-MSMS",
                                   parse_protein_accession_function = NA) {

  message("Processing: ", basename(mq_evidence_file))

  # Proceed only if the evidence file exists, otherwise skip with a message
  if (file.exists(mq_evidence_file)) {

    # Locate all PTM site files within the ptm folder
    ptm_files <- list.files(ptm_dir, full.names = TRUE)

    # Generate PTM names from file names: add brackets, drop "Sites.txt"
    ptm_names <- paste0(
      "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
    )

    # Run the stoichiometry calculation
    calculate_stoichiometry2(
      mq_evidence_data = mq_evidence_file,
      sample_info_file = sample_info_file,
      ref_protein_dt   = ref_protein_dt,
      ptm_files        = ptm_files,
      ptm_names        = ptm_names,
      output_prefix    = output_prefix,
      ptm_mapping_file = ptm_mapping_file,
      K_only           = FALSE,
      selected_type    = selected_type,
      parse_protein_accession_function = parse_protein_accession_function
    )
  } else {
    message("File does not exist: ", mq_evidence_file)
  }
  gc()  # free memory
}

```

# Stoichiometry without diagnostic ions

Computes stoichiometry for the PNAS2022 standard MaxQuant search — once including secondary-peptide identifications (MULTI-MSMS + MULTI-SECPEP) and once with MULTI-MSMS only.

```{r load_mq_std}

# MaxQuant output root for the standard (no diagnostic ion) PNAS2022 search
mq_std_data_dir <- file.path(project.dir, "data/MQ_Std/PNAS2022")

# Sample matrix (one row per run)
mq_std_sample_dt <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_Std"
) %>% data.table

# Sequential sample id (1:N)
mq_std_sample_dt[, sample_id := 1:.N]

```

```{r run_mq_std}

## Stoichiometry analysis using MULTI-MSMS with MULTI-SECPEP (secondary peptide)
## and, separately, MULTI-MSMS only.

## --- MULTI-MSMS + MULTI-SECPEP ---------------------------------------------

# Restrict the standard search to data-A and data-D for the MULTI-MSMS +
# MULTI-SECPEP comparison.
# Data-A and data-D — BRD proteins with or without derivatisation — are analysed to assess the effect of derivatisation.
mq_std_secpep_sample_dt <- mq_std_sample_dt[data %in% c("data-A", "data-D")]

mq_std_msms_secpep_dir <- file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS_SECPEP")

process_stoichiometry(
  sample_dt        = mq_std_secpep_sample_dt,
  data_dir         = mq_std_data_dir,
  out_dir          = mq_std_msms_secpep_dir,
  ref_protein_dt   = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file,
  selected_type    = NA,
  prefix_label     = "including_SECPEP_"
)

## --- MULTI-MSMS only -------------------------------------------------------

mq_std_msms_dir <- file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS")

process_stoichiometry(
  sample_dt        = mq_std_sample_dt,
  data_dir         = mq_std_data_dir,
  out_dir          = mq_std_msms_dir,
  ref_protein_dt   = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file,
  selected_type    = c("MULTI-MSMS"),
  prefix_label     = ""
)

```

# Stoichiometry with diagnostic ions — PNAS

Computes stoichiometry for the diagnostic-ion (no water-loss) MaxQuant search of the PNAS2022 datasets (A, B1, B2, C).

```{r load_mq_di}

# MaxQuant output root for the diagnostic-ion PNAS2022 search
mq_di_data_dir <- file.path(project.dir, "data/MQ_DI/PNAS2022")

# Output directory for results
mq_di_dir <- file.path(results.dir, "p2-analysis-setting", "MQ_DI")
create.dir(mq_di_dir)

# Sample matrix (one row per run)
mq_di_sample_dt <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI"
) %>% data.table

# Sequential sample id (1:N)
mq_di_sample_dt[, sample_id := 1:.N]

```

```{r run_mq_di}

# For each run in the sample matrix, calculate the stoichiometry
for (i in seq_len(nrow(mq_di_sample_dt))) {

  prefix     <- mq_di_sample_dt$prefix[i]
  mq_setting <- gsub("_$", "", sub("^data-[^_]+_", "", prefix))  # search setting

  # Per-run directory holding the evidence file and the ptm/ folder
  run_dir <- file.path(mq_di_data_dir, mq_di_sample_dt[i, data], mq_setting)

  process_stoichiometry2(
    mq_evidence_file = file.path(run_dir, paste0(prefix, "evidence.txt")),
    sample_info_file = file.path(
      mq_di_data_dir,
      mq_di_sample_dt[i, data],
      paste0("MS_dataset_overview_PXD031221_", mq_di_sample_dt[i, data], ".csv")
    ),
    ptm_dir          = file.path(run_dir, "ptm"),
    output_prefix    = file.path(mq_di_dir, paste0("DI_", prefix)),
    ref_protein_dt   = ref_protein_dt,
    ptm_mapping_file = ptm_mapping_file
  )
}

```

# Stoichiometry with diagnostic ions — MS_KR_1

Computes stoichiometry for the MS_KR_1 dataset (dataset D: hypoxia vs normoxia, with JMJD6 re-expression).

```{r load_ms_kr1}

# MaxQuant output directory for the MS_KR_1 dataset
ms_kr1_data_dir <- file.path(project.dir, "data/MQ_DI/MS_KR_1")

# Output directory for results
ms_kr1_dir <- file.path(results.dir, "p2-analysis-setting", "MS_KR_1")
create.dir(ms_kr1_dir)

```

```{r run_ms_kr1}

process_stoichiometry2(
  mq_evidence_file = file.path(ms_kr1_data_dir, "MS_KR_1_evidence.txt"),
  sample_info_file = file.path(ms_kr1_data_dir, "sample_info.csv"),
  ptm_dir          = file.path(ms_kr1_data_dir, "ptm"),
  output_prefix    = file.path(ms_kr1_dir, "MS_KR_1_"),
  ref_protein_dt   = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file
)

```

# Stoichiometry with diagnostic ions — MS_SS

Computes stoichiometry for the MS_SS dataset (dataset E: O2 and doxycycline titration).

```{r load_ms_ss}

# MaxQuant output directory for the MS_SS dataset
ms_ss_data_dir <- file.path(project.dir, "data/MQ_DI/MS_SS")

# Output directory for results
ms_ss_dir <- file.path(results.dir, "p2-analysis-setting", "MS_SS")
create.dir(ms_ss_dir)

```

```{r run_ms_ss}

process_stoichiometry2(
  mq_evidence_file = file.path(ms_ss_data_dir, "MS_SS_evidence.txt"),
  sample_info_file = file.path(ms_ss_data_dir, "sample_info.csv"),
  ptm_dir          = file.path(ms_ss_data_dir, "ptm"),
  output_prefix    = file.path(ms_ss_dir, "MS_SS_"),
  ref_protein_dt   = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file
)

```

# Session information

```{r session_info}

sessioninfo::session_info()

```
