---
title: "p2-03 · Compare miscleavage settings"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Compare PSM coverage of the lysine-rich domain of BRD4 across miscleavage
settings (m2 / m5 / m7).

**Inputs:** stoichiometry tables from p2-01 (`results/p2-analysis-setting/`); sample matrix.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("khroma")

```

# PSM coverage by miscleavage

Compares PSM coverage of lysine-rich domains between the m2, m5 and m7 miscleavage search settings.

```{r load_mq_std}

# Sample run info (one row per run)
mq_std_sample_dt <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_Std"
) %>% data.table

# Sequential sample id (1:N)
mq_std_sample_dt[, sample_id := 1:.N]

# Stoichiometry tables for the MQ_Std (MULTI-MSMS) search
mq_std_stoic_dt <- lapply(
  mq_std_sample_dt[, prefix],
  read_stoic_data,
  pre_prefix = "",
  post_fix   = "",
  dir_path   = file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS")
) %>% rbindlist

```


```{r plot_psm_by_miscleavage, fig.height = 3}

## BRD4 PSM coverage across miscleavage settings (m2, m5, m7), WT derivatised,
## overlaid on a single axis (one coloured line per setting).

brd4_settings <- c("data-A_trp_m2_v2_def", "data-A_trp_m5_v5_def", "data-A_trp_m7_v7_def")

# Per-position PSM coverage profile for one search setting
brd4_coverage_data <- function(setting) {
  preprocess_stoichiometry_data_for_plotting(
    stoichiometry_data = mq_std_stoic_dt[sample_name == "JQ1_HeLaWT_derivatised" & condition == setting],
    accession = "O60885", plot_range = c(531, 581), all.protein.bs = all.protein.bs
  )$coverage_data[, miscleavage := setting]
}

brd4_coverage_dt <- rbindlist(lapply(brd4_settings, brd4_coverage_data))
brd4_coverage_dt[, miscleavage := factor(miscleavage, levels = brd4_settings, labels = c("m2", "m5", "m7"))]

ggplot(brd4_coverage_dt, aes(x = aa_pos, y = psm_coverage, color = miscleavage)) +
  geom_line() +
  coord_cartesian(ylim = c(0, 350)) +
  labs(title = "BRD4 PSM coverage by miscleavage",
       x = "Amino-acid position", y = "PSM coverage", color = "Miscleavage") +
  scale_color_bright()

```


# Session information

```{r session_info}

sessioninfo::session_info()

```
