---
title: "p2-05 · Plot diagnostic-ion analysis"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Evaluate how diagnostic ions improve hydroxylation analysis —
precision, WT / KO overlap, and comparison with PNAS 2022.

**Inputs:** stoichiometry tables from p2-01; PNAS2022 reference data.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01, p2-04. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

## Script-specific packages
library("patchwork")
library("eulerr")
library("RColorBrewer")

```

# Helper functions

```{r helper_functions}

# Area-proportional Euler diagram of WT vs KO site overlap
plot_overlap_venn <- function(wt_ids, ko_ids, main) {
  venn_list <- list(WT_DI_site = wt_ids, KO_DI_site = ko_ids)
  plot(
    euler(venn_list),
    fills      = list(fill = brewer.pal(3, "Set2"), alpha = 0.4),
    legend     = list(side = "right"),
    quantities = TRUE,
    main       = main
  )
}

```

# Import data

Loads the diagnostic-ion sample matrix (data paths come from `_setup.R`).

```{r load_inputs, warning=FALSE}

# Sample run info (MQ diagnostic-ion search)
mq_di_sample_dt <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI"
) %>% data.table

# Sequential sample id (1:N)
mq_di_sample_dt[, sample_id := 1:.N]

```

# Diagnostic ions and precision

Assesses how requiring a diagnostic ion changes the precision of hydroxylation calls.

```{r precision}

# Read stoichiometry data for MQ_DI (change from the older version: a potential
# diagnostic ion with water loss is no longer considered)
mq_di_stoic_dt <- lapply(
  mq_di_sample_dt[, prefix],
  read_stoic_data,
  pre_prefix = "DI_",
  post_fix   = "_DI",
  dir_path   = file.path(results.dir, "p2-analysis-setting", "MQ_DI")
) %>% rbindlist

# fwrite(mq_di_stoic_dt[ptm != ""], file = file.path(results.dir, "p2-analysis-setting", "all_hydroxylysine.csv"))

mq_di_stoic_dt <- mq_di_stoic_dt[aa == "K"]
mq_di_stoic_dt[, genotype :=
  str_split_fixed(sample_name, "_", n = 3)[, 2] %>% factor(levels = c("HeLaWT", "HeLaJMJD6KO"))
]

# Sites with a diagnostic ion, and a per-row flag for them
di_sites_dt <- mq_di_stoic_dt[diagnostic_peak == "+"][
  !duplicated(paste(protein_accession, aa_pos)), .(protein_accession, aa_pos)
]
mq_di_stoic_dt[, DI_site :=
  paste(protein_accession, aa_pos) %in% di_sites_dt[, paste(protein_accession, aa_pos)]
]

# Protein feature data (lysines only), with a methionine-proximity annotation
protein_feature_dt <- fread(
  file.path(data.dir, "processed_data_from_PNAS2022/all_protein_feature_per_position.csv")
)
protein_feature_dt <- protein_feature_dt[residue == "K"]
setnames(protein_feature_dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

protein_feature_dt[, met_within_2 := case_when(
  nchar(Window) == 11 & grepl("M", substr(Window, start = 4, stop = 8)) ~ "Yes",  # M in the middle
  nchar(Window) == 11 ~ "No",                                                     # no M in window
  TRUE ~ "edge"                                                                   # M present but off-centre
)]

# Attach feature annotation to the stoichiometry data
wt_vs_ko_stoic_dt <- merge(
  copy(mq_di_stoic_dt),
  protein_feature_dt,
  by = c("protein_accession", "aa_pos")
)

# Keep only sites covered (>= 1 PSM) in both WT and KO so presence/absence is comparable
wt_coverage_dt <- wt_vs_ko_stoic_dt[genotype == "HeLaWT"][
  , list(total_wt_sum_psm_mapped = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos)]
ko_coverage_dt <- wt_vs_ko_stoic_dt[genotype == "HeLaJMJD6KO"][
  , list(total_ko_sum_psm_mapped = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos)]

all_coverage_dt <- merge(wt_coverage_dt, ko_coverage_dt, by = c("protein_accession", "aa_pos"))

wt_vs_ko_stoic_dt <- wt_vs_ko_stoic_dt[
  paste(protein_accession, aa_pos) %in%
    all_coverage_dt[total_wt_sum_psm_mapped > 0 & total_ko_sum_psm_mapped > 0][
      , paste(protein_accession, aa_pos)]
]

# Number of sites analysed, per genotype
wt_vs_ko_stoic_dt[!duplicated(paste(genotype, protein_accession, aa_pos))][, .N, by = genotype]

# Hydroxylated K rows only, one (DI-preferred) row per genotype/site
koh_wt_vs_ko_dt <- wt_vs_ko_stoic_dt[ptm == "[Oxidation (K)]"]
koh_wt_vs_ko_dt <- koh_wt_vs_ko_dt[order(-DI_site)][
  !duplicated(paste(genotype, protein_accession, aa_pos))
]

g1 <- ggplot(
  koh_wt_vs_ko_dt[met_within_2 != "edge"],
  aes(x = genotype, fill = met_within_2)
) +
  geom_bar() +
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g2 <- ggplot(
  koh_wt_vs_ko_dt[met_within_2 != "edge" & DI_site == TRUE],
  aes(x = genotype, fill = met_within_2)
) +
  geom_bar() +
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

# Number of sites for plotting g2
koh_wt_vs_ko_dt[met_within_2 != "edge" & DI_site == TRUE][, .N, by = list(met_within_2, genotype)]

g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```

# WT vs KO site overlap

Compares hydroxylysine sites detected in JMJD6 WT vs KO cells (overlap and methionine proximity).

```{r overlap_wt_ko}

koh_per_site_dt <- dcast(
  koh_wt_vs_ko_dt,
  protein_accession + aa_pos + DI_site + met_within_2 ~ genotype,
  fun.aggregate = length
)

koh_per_site_dt[, Hyl_found_in := case_when(
  HeLaWT == 1 & HeLaJMJD6KO == 1 ~ "both",
  HeLaWT == 1 ~ "WT",
  HeLaJMJD6KO == 1 ~ "JMJD6KO",
  TRUE ~ "not_found"
) %>% factor(levels = c("WT", "JMJD6KO", "both", "not_found"))]

koh_per_site_dt <- koh_per_site_dt[met_within_2 != "edge"]

koh_per_site_count_dt <- koh_per_site_dt[, .N, by = list(Hyl_found_in, met_within_2, DI_site)]

g1 <- ggplot(koh_per_site_count_dt, aes(x = Hyl_found_in, y = N, fill = met_within_2)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g2 <- ggplot(koh_per_site_count_dt[DI_site == TRUE], aes(x = Hyl_found_in, y = N, fill = met_within_2)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  theme(aspect.ratio = 3) +
  scale_x_discrete(guide = guide_axis(angle = 90))

g1 + g2 + plot_layout(guides = "collect") & theme(legend.position = "bottom")

merge(
  koh_wt_vs_ko_dt[!duplicated(paste(protein_accession, gene_name)), .(protein_accession, gene_name)],
  koh_per_site_dt[DI_site == TRUE & Hyl_found_in == "both"]
)

# Statistical significance of methionine enrichment (all sites vs hydroxylated)
all_k_met_count_dt <- rbind(
  copy(protein_feature_dt[met_within_2 != "edge", .N, by = met_within_2])[, data_type := "all_K"],
  copy(koh_per_site_count_dt[, list(N = sum(N)), by = met_within_2])[, data_type := "koh"]
) %>%
  dcast(met_within_2 ~ data_type, value.var = "N")

fisher.test(all_k_met_count_dt[, .(all_K, koh)], alternative = "two.sided")

# WT vs KO site overlap (all sites)
plot_overlap_venn(
  wt_ids = koh_wt_vs_ko_dt[genotype == "HeLaWT", paste(protein_accession, aa_pos)],
  ko_ids = koh_wt_vs_ko_dt[genotype == "HeLaJMJD6KO", paste(protein_accession, aa_pos)],
  main   = "Hydroxylated Sites Overlap (All)"
)

# WT vs KO site overlap (diagnostic-ion sites only)
plot_overlap_venn(
  wt_ids = koh_wt_vs_ko_dt[genotype == "HeLaWT" & DI_site == TRUE, paste(protein_accession, aa_pos)],
  ko_ids = koh_wt_vs_ko_dt[genotype == "HeLaJMJD6KO" & DI_site == TRUE, paste(protein_accession, aa_pos)],
  main   = "Hydroxylated Sites Overlap (with DI)"
)

```

# Comparison with PNAS 2022

Benchmarks the diagnostic-ion-based sites against the manually curated sites from the PNAS 2022 study.

```{r compare_pnas2022}

# PNAS2022 curated stoichiometry data
pnas2022_stoic_dt <- fread(
  file.path(data.dir, "processed_data_from_PNAS2022/long_K_stoichiometry_data.csv")
)
setnames(pnas2022_stoic_dt, old = c("uniprot_id", "position", "residue"), new = c("protein_accession", "aa_pos", "aa"))
pnas2022_stoic_dt[, accession_position := paste0(protein_accession, "_", aa_pos)]

# Flag MQ_DI sites that PNAS2022 curated as hydroxylated
mq_di_stoic_dt[, curated_oxK_site :=
  paste0(protein_accession, "_", aa_pos) %in%
  pnas2022_stoic_dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)]
]

# One hydroxylated WT row per site (DI-preferred)
mq_di_hydroxyk_dt <- mq_di_stoic_dt[ptm == "[Oxidation (K)]" & genotype == "HeLaWT"][
  order(genotype, DI_site)
][!duplicated(paste(protein_accession, aa_pos))]

# Cross-tabulate DI sites against PNAS2022 curated sites
hyl_precision_dt <- mq_di_hydroxyk_dt[, table(DI_site, curated_oxK_site) %>% addmargins] %>% data.table
hyl_precision_dt <- hyl_precision_dt[DI_site %in% c("Sum", "TRUE") & curated_oxK_site != "Sum"]

hyl_precision_dt

ggplot(
  hyl_precision_dt,
  aes(x = DI_site, y = N, fill = curated_oxK_site)
) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("TRUE" = "coral2", "FALSE" = "#BBBBBB")) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme(aspect.ratio = 3.5)

dcast(hyl_precision_dt, DI_site ~ curated_oxK_site, value.var = "N") %>%
  {fisher.test(.[, c("FALSE", "TRUE"), with = FALSE], alternative = "two.sided")}

```


# Session information

```{r session_info}

sessioninfo::session_info()

```
