---
title: "p2-06 · Analyse MS_KR_1 stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Analyse MS_KR_1 lysine-hydroxylation stoichiometry — oxygen
sensitivity (hypoxia vs normoxia) across BRD2/3/4, including JMJD6 re-expression.
Stoichiometry itself is computed in p2-01; this script reads those tables.

**Inputs:** stoichiometry tables from `results/p2-analysis-setting/MS_KR_1/`
(written by p2-01); UniProt reference proteome (`reference_fasta`).

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Per-site stoichiometry across conditions

Reads the MS_KR_1 stoichiometry table and plots site-specific hydroxylation along
BRD4 (O60885), BRD3 (Q15059) and BRD2 (P25440) across the four conditions —
parental WT, induced cells without doxycycline, and JMJD6 re-expression (+Dox)
under normoxia and 0.1% O2. The `diagnostic_peak` flag marks sites confirmed by the
hydroxylysine immonium ion.

```{r plot_brd_stoichiometry}

# Read stoichiometry data for MS_KR_1
ms_kr1_stoic_dt <- read_stoic_data(
  prefix     = "MS_KR_1_",
  pre_prefix = "",
  post_fix   = "_DI",
  dir_path   = file.path(results.dir, "p2-analysis-setting", "MS_KR_1")
)

# Diagnostic-ion flag and an ordered sample factor (all four conditions)
ms_kr1_stoic_dt[, is_diagnostic_peak := diagnostic_peak == "+"]

kr1_conditions <- c(
  "HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA",
  "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"
)
ms_kr1_stoic_dt[, sample_name := factor(sample_name, levels = kr1_conditions)]

# Per-site stoichiometry for one accession across all four conditions
plot_brd_stoic <- function(accession, plot_range) {
  plot_ptm_stoichiometry(
    ms_kr1_stoic_dt[sample_name %in% kr1_conditions],
    accession = accession, plot_range = plot_range, all.protein.bs, sample_colors = NA
  )
}

plot_brd_stoic("O60885", c(531, 581))  # BRD4
plot_brd_stoic("Q15059", c(483, 533))  # BRD3
plot_brd_stoic("P25440", c(540, 590))  # BRD2

```

# Session information

```{r session_info}

sessioninfo::session_info()

```
