---
title: "p2-09 · Export stoichiometry tables"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Export processed BRD stoichiometry (MS_SS, MS_KR_1, PNAS) to a
formatted Excel workbook (supplementary data).

**Inputs:** stoichiometry tables from p2-01; reference proteome.

**Outputs:** Excel workbook under `results/`.

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))
library("janitor")
library("openxlsx")

```

# Export tables to Excel

Assembles the per-site stoichiometry tables (PNAS data-A/B/C, MS_KR_1, MS_SS) and writes them to a formatted Excel workbook for the supplementary data.

```{r load_data}

# A diagnostic-peak flag shared by all loaded tables
add_di_flag <- function(dt) dt[, is_diagnostic_peak := diagnostic_peak == "+"][]

mq_di_dir <- file.path(results.dir, "p2-analysis-setting", "MQ_DI")

#--- PNAS dataset A (MaxQuant with diagnostic ions) ---
mq_di_a_stoic_dt <- read_stoic_data(
  prefix = "DI_data-A_trp_m7_v7_def_", pre_prefix = "", post_fix = "_DI",
  dir_path = mq_di_dir
) %>% add_di_flag

#--- PNAS datasets B1 and B2 ---
mq_di_sample_dt <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI"
) %>% data.table
mq_di_sample_dt[, sample_id := 1:.N]

mq_di_b_stoic_dt <- lapply(
  mq_di_sample_dt[data != c("data-A", "data-C"), prefix],
  read_stoic_data,
  pre_prefix = "DI_", post_fix = "_DI", dir_path = mq_di_dir
) %>% rbindlist %>% add_di_flag

#--- PNAS dataset C ---
mq_di_c_stoic_dt <- read_stoic_data(
  prefix = "DI_data-C_trp_m7_v7_mCC_", pre_prefix = "", post_fix = "_DI",
  dir_path = mq_di_dir
) %>% add_di_flag

#--- PNAS manually curated sites ---
pnas2022_stoic_dt <- fread(file.path(data.dir, "processed_data_from_PNAS2022/long_K_stoichiometry_data.csv"))
setnames(
  pnas2022_stoic_dt,
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"),
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)
pnas2022_stoic_dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)],
  pnas2022_stoic_dt,
  by = "protein_accession"
)
pnas2022_stoic_dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos),
  ptm = fifelse(curated_oxK_site == TRUE, "[Oxidation (K)]", " ")
)]
# Harmonise column names with the exported tables
setnames(
  pnas2022_stoic_dt,
  c("stoichiometry", "protein_accession", "gene_name", "aa_pos", "curated_oxK_site"),
  c("Stoichiometry_PNAS2022", "Protein_accession", "Gene_name", "Amino_acid_pos", "PNAS2022_curated_oxK_site")
)

#--- MS_KR_1 (dataset D) ---
ms_kr1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_", pre_prefix = "", post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1")
) %>% add_di_flag

#--- MS_SS (dataset E) ---
ms_ss_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_", pre_prefix = "", post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS")
) %>% add_di_flag

```

```{r export_helper}

# Output directory for the supplementary workbooks (created if missing)
export_out_dir <- file.path(results.dir, "p2_MS_SS_KR")
create.dir(export_out_dir)

# Working columns dropped from every stoichiometry table before export
export_drop_cols <- c(
  "sum_intensity_per_position", "sum_psm_mapped_per_position",
  "sum_the_number_of_peptide", "localization_prob", "score_for_localization",
  "best_localization_ms_ms_id", "best_localization_raw_file", "max_score",
  "diagnostic_peak", "condition"
)

# Publication-ready column renaming, applied identically to every sheet
export_old_names <- c(
  "sample_name", "protein_accession", "gene_name", "aa", "aa_pos", "ptm",
  "stoichiometry", "sum_peak_intensity", "sum_psm_mapped",
  "the_number_of_peptide", "is_diagnostic_peak"
)
export_new_names <- c(
  "Sample_Name", "Protein_accession", "Gene_name", "Amino_acid", "Amino_acid_pos",
  "PTM_type", "Stoichiometry", "Total_peak_intensity", "Total_PSM_mapped",
  "Number_of_peptide", "Diagnostic_peak_presence"
)

## Format a per-site stoichiometry table for the supplementary workbook and write
## it to <results.dir>/p2_MS_SS_KR/<file_name>. Subsets to lysines, drops working
## columns, reorders + renames to publication headers, flags PNAS2022 curated
## sites, and writes the xlsx. `extra_filter` optionally transforms the table
## (original column names still in place) before reordering — used for the MS_SS
## cross-protein clean-up.
export_stoic_table <- function(stoic_dt, file_name, curated_dt, extra_filter = NULL) {
  dt <- stoic_dt[aa == "K"]                       # subset to lysines (copy)
  dt[, (export_drop_cols) := NULL]                # drop working columns

  if (!is.null(extra_filter)) dt <- extra_filter(dt)

  # Move stoichiometry / diagnostic-peak flag to their publication positions
  setcolorder(dt, append(setdiff(names(dt), "stoichiometry"), "stoichiometry", after = 6))
  setcolorder(dt, append(setdiff(names(dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13))

  setnames(dt, export_old_names, export_new_names)  # publication-ready headers

  # Flag sites manually curated as oxK in PNAS2022
  dt[, PNAS2022_curated_oxK_site :=
       paste0(Protein_accession, "_", Amino_acid_pos) %in%
       curated_dt[PNAS2022_curated_oxK_site == TRUE,
                  paste0(Protein_accession, "_", Amino_acid_pos)]]

  write.xlsx(dt, file.path(export_out_dir, file_name))
  invisible(dt)
}

```


```{r export_di_a}

# PNAS dataset A (diagnostic ions)
export_stoic_table(mq_di_a_stoic_dt, "MQ_DI_dtA_stoic_dt.xlsx", pnas2022_stoic_dt)

```


```{r export_di_b}

# PNAS datasets B1 and B2 (diagnostic ions)
export_stoic_table(mq_di_b_stoic_dt, "MQ_DI_dtB_stoic_dt.xlsx", pnas2022_stoic_dt)

```


```{r export_di_c}

# PNAS dataset C (diagnostic ions)
export_stoic_table(mq_di_c_stoic_dt, "MQ_DI_dtC_stoic_dt.xlsx", pnas2022_stoic_dt)

```


```{r export_ms_kr1}

# MS_KR_1 (dataset D)
export_stoic_table(ms_kr1_stoic_dt, "MS_KR1_stoic_K.xlsx", pnas2022_stoic_dt)

```


```{r export_ms_ss}

# MS_SS (dataset E). Drops cross-protein false positives (BRD4 signal reported in
# BRD2/3 samples and vice versa, e.g. from shared / homologous peptides) before
# formatting.
export_stoic_table(
  ms_ss_stoic_dt, "MS_SS_stoic_K.xlsx", pnas2022_stoic_dt,
  extra_filter = function(dt) dt[
    !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
        (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
  ]
)

```


# Session information

```{r session_info}

sessioninfo::session_info()

```
