---
title: "p2-01 · Compute PTM stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Compute site-specific lysine-hydroxylation (PTM) stoichiometry from
MaxQuant outputs for every dataset in the study. This is the single computation
step of Part 2; all downstream scripts read the tables written here.

**Inputs:**

- MaxQuant `evidence.txt` per dataset under `data/MQ_Std/` and `data/MQ_DI/`
- sample matrix `data/analysis_setting/PXD031221_sample_matrix.xlsx` and per-dataset `sample_info`
- PTM mapping `data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv`
- UniProt reference proteome (`reference_fasta`; see Setup)

**Outputs:** stoichiometry tables under `results/p2-analysis-setting/`
(`MQ_Std_MSMS_SECPEP/`, `MQ_Std_MSMS/`, `MQ_DI/`, `MS_KR_1/`, `MS_SS/`).

**Upstream:** none. **Downstream:** p2-02 … p2-09 read these tables.

# Setup

```{r setup, message = FALSE, warning = FALSE}

## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths
## (project.dir, data.dir, results.dir, reference_fasta).
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Stoichiometry without diagnostic ions

Computes stoichiometry for the PNAS2022 standard MaxQuant search — once including secondary-peptide identifications (MULTI-MSMS + MULTI-SECPEP) and once with MULTI-MSMS only.

```{r data_directory_wo_di}

# Load data into environment
pnas2022_data <- file.path(
  project.dir,
  "data/MQ_Std/PNAS2022" 
)

# Load sample info data 
MQ_Std_data <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_Std"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each row (1:N), 
#where N is the total number of rows in the data.table
MQ_Std_data[, sample_id := 1:.N]

```

```{r calculate_stoichiometry_wo_di}

## Stoichiometry analysis using with MULTI-MSMS with 
# MULTI-SECPEP (secondary peptide) and MULTI-MSMS only

# Automate stoichiometry calculations across MaxQuant datasets (MULTI-MSMS + 
# MULTI-SECPEP and MULTI-MSMS only)
process_stoichiometry <- function(MQ_Std_data,
                                  output_dir,
                                  ref_protein_dt,
                                  ptm_mapping_file,
                                  selected_type = NA,
                                  prefix_label = "") {
  
  # Create output directory if it does not already exist 
  dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)
  
  # For each sample_id in the MQ_Std data, calculate the stoichiometry
  for (i in seq_len(nrow(MQ_Std_data))) {
    prefix    <- MQ_Std_data$prefix[i] # Extract the prefix column
    MQ_Std_sample <- MQ_Std_data$data[i] # Extract the data column 
    mq_setting <- gsub("_$", "", sub("^data-[^_]+_", "", prefix)) # search setting
    
    # Define file path to MaxQuant evidence files
    mq_evidence_data <- file.path(pnas2022_data, 
                                  MQ_Std_sample, mq_setting, 
                                  paste0(prefix, "evidence.txt"))
    
    # Print which sample id is being processed
    message("Processing: ", basename(mq_evidence_data)) 
    
    
    # Checks whether the sample ID has the corresponding evidence.txt file. If yes, then proceed
    if (file.exists(mq_evidence_data)) {
      
      # Define corresponding sample information (metadata) file
      sample_info_file <- file.path(
        pnas2022_data, MQ_Std_sample,
        paste0("MS_dataset_overview_PXD031221_", MQ_Std_sample, ".csv")
      )
      
      # Define output prefix for results
      output_prefix <- file.path(output_dir, paste0(prefix_label, prefix))
      
      # Run main stoichiometry calculation
      calculate_stoichiometry(
        mq_evidence_data = mq_evidence_data,
        sample_info_file = sample_info_file,
        ref_protein_dt   = ref_protein_dt,
        output_prefix    = output_prefix,
        ptm_mapping_file = ptm_mapping_file,
        K_only           = FALSE,
        selected_type    = selected_type
      )
      
    } else {
      # If evidence.txt does not exist, skip the function and print below message
      message("File does not exist: ", mq_evidence_data)
    }
    # Garbage collection - to free up memory 
    gc()
  }
}


# Define file path common PTM mapping file
ptm_mapping_file <- file.path(
  project.dir,
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
)

## MSMS + SECPEP

# Restrict the standard search to data-A and data-D for the MULTI-MSMS +
# MULTI-SECPEP comparison.
# NOTE: confirm why data-A and data-D specifically (vs the other MQ_Std datasets).
samples_to_process <- MQ_Std_data[data %in% c("data-A", "data-D")]

# Define output directory for MSMS + SECPEP results
MQ_Std_MSMS_SECPEP_dir <- file.path(results.dir, "p2-analysis-setting","MQ_Std_MSMS_SECPEP")

# Run stoichiometry processing for MSMS + SECPEP
process_stoichiometry(
  MQ_Std_data = samples_to_process,
  output_dir       = MQ_Std_MSMS_SECPEP_dir,
  ref_protein_dt = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file,
  selected_type    = NA,
  prefix_label     = "including_SECPEP_"
)

##MSMS only

# Define output directory for MSMS-only results
MQ_Std_MSMS_dir <- file.path(results.dir,"p2-analysis-setting", "MQ_Std_MSMS")

# Run stoichiometry processing for MSMS-only
process_stoichiometry(
  MQ_Std_data = MQ_Std_data,
  output_dir       = MQ_Std_MSMS_dir,
  ref_protein_dt = ref_protein_dt,
  ptm_mapping_file = ptm_mapping_file,
  selected_type    = c("MULTI-MSMS"),
  prefix_label     = ""
)
```

# Stoichiometry with diagnostic ions — PNAS

Computes stoichiometry for the diagnostic-ion (no water-loss) MaxQuant search of the PNAS2022 datasets (A, B1, B2, C).

```{r data_directory_MQ_DI}

# Load data into environment
MQ_DI_data <- file.path(
  project.dir,
  "data/MQ_DI/PNAS2022" 
)

# Create file path for results
MQ_DI_dir <- file.path(results.dir, "p2-analysis-setting", "MQ_DI")
create.dir(MQ_DI_dir)


```


```{r sample_info_MQ_DI}

# Load sample run info data 
MQ_DI_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_DI" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_run_info[, sample_id := 1:.N]

```


```{r calculate_stoichiometry_MQ_DI}

# For each sample ID calculate the stoichiometry
for (i in seq_len(nrow(MQ_DI_run_info))){ # Takes each row from the MQ_DI_run_info
  
  prefix <- MQ_DI_run_info$prefix[i]
  mq_setting <- gsub("_$", "", sub("^data-[^_]+_", "", prefix)) # search setting
  
  # Retrieves the evidence.txt file for each sample ID 
  mq_evidence_data = file.path(MQ_DI_data,
                               MQ_DI_run_info[i, data], mq_setting,
                               paste0(prefix, "evidence.txt"))
  
  print(paste0("Processing: ", basename(mq_evidence_data)))
  
  # Checks whether the sample ID has the corresponding evidence.txt file. If yes, then proceed
  if (file.exists(mq_evidence_data)) { 
    
    # Locate and fetch all PTM site files within the ptm folder
    ptm_files <- list.files(
      file.path(MQ_DI_data, MQ_DI_run_info[i, data], mq_setting, "ptm"),
      full.names = TRUE)
    
    # Generate PTM names based on file names, add brackets and remove "Sites.txt"
    ptm_names <- paste0(
      "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
    )
    
    # Run the stoichiometry calculation 
    stoic.dt <- calculate_stoichiometry2(
      mq_evidence_data = mq_evidence_data,
      sample_info_file = file.path( # Corresponding sample info data 
        MQ_DI_data,
        MQ_DI_run_info[i, data],
        paste0(
          "MS_dataset_overview_PXD031221_",
          MQ_DI_run_info[i, data],
          ".csv"
        )
      ),
      ref_protein_dt = ref_protein_dt,
      ptm_files = ptm_files,
      ptm_names = ptm_names,
      output_prefix = file.path(MQ_DI_dir,  # Define output prefix for results
                                paste0("DI_", MQ_DI_run_info[i, prefix])),
      ptm_mapping_file = ptm_mapping_file,
      K_only = FALSE,
      selected_type = c("MULTI-MSMS"),
    )
    # If evidence.txt does not exist, skip the function and print below message
  } else {
    print(paste0("File does not exist: ", mq_evidence_data))
  }
  
  
  gc()
  
}

```



# Stoichiometry with diagnostic ions — MS_KR_1

Computes stoichiometry for the MS_KR_1 dataset (dataset D: hypoxia vs normoxia, with JMJD6 re-expression).

```{r data_directory_MS_KR_1}

# MaxQuant output directory for the MS_KR_1 dataset
MS_KR1_data <- file.path(
  project.dir,
  "data/MQ_DI/MS_KR_1"
)

# Output directory for results
MS_KR_1_dir <- file.path(results.dir, "p2-analysis-setting", "MS_KR_1")
create.dir(MS_KR_1_dir)

```

```{r calculate_stoichiometry_MS_KR_1}

# Define file path to MaxQuant evidence file
mq_evidence_data <- file.path(MS_KR1_data, "MS_KR_1_evidence.txt")

message("Processing: ", basename(mq_evidence_data))

if (file.exists(mq_evidence_data)) {

  # Locate all PTM site files within the ptm folder
  ptm_files <- list.files(file.path(MS_KR1_data, "ptm"), full.names = TRUE)

  # Generate PTM names from file names: add brackets, drop "Sites.txt"
  ptm_names <- paste0(
    "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
  )

  # Run the stoichiometry calculation
  stoic.dt <- calculate_stoichiometry2(
    mq_evidence_data = mq_evidence_data,
    sample_info_file = file.path(MS_KR1_data, "sample_info.csv"),
    ref_protein_dt   = ref_protein_dt,
    ptm_files        = ptm_files,
    ptm_names        = ptm_names,
    output_prefix    = file.path(MS_KR_1_dir, "MS_KR_1_"),
    ptm_mapping_file = ptm_mapping_file,
    K_only           = FALSE,
    selected_type    = "MULTI-MSMS",
    parse_protein_accession_function = NA
  )
} else {
  message("File does not exist: ", mq_evidence_data)
}

gc()

```

# Stoichiometry with diagnostic ions — MS_SS

Computes stoichiometry for the MS_SS dataset (dataset E: O2 and doxycycline titration).

```{r data_directory_MS_SS}

# MaxQuant output directory for the MS_SS dataset
MS_SS_data <- file.path(
  project.dir,
  "data/MQ_DI/MS_SS"
)

# Output directory for results
MS_SS_dir <- file.path(results.dir, "p2-analysis-setting", "MS_SS")
create.dir(MS_SS_dir)

```

```{r calculate_stoichiometry_MS_SS}

# Define file path to MaxQuant evidence file
mq_evidence_data <- file.path(MS_SS_data, "MS_SS_evidence.txt")

message("Processing: ", basename(mq_evidence_data))

if (file.exists(mq_evidence_data)) {

  # Locate all PTM site files within the ptm folder
  ptm_files <- list.files(file.path(MS_SS_data, "ptm"), full.names = TRUE)

  # Generate PTM names from file names: add brackets, drop "Sites.txt"
  ptm_names <- paste0(
    "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
  )

  # Run the stoichiometry calculation
  stoic.dt <- calculate_stoichiometry2(
    mq_evidence_data = mq_evidence_data,
    sample_info_file = file.path(MS_SS_data, "sample_info.csv"),
    ref_protein_dt   = ref_protein_dt,
    ptm_files        = ptm_files,
    ptm_names        = ptm_names,
    output_prefix    = file.path(MS_SS_dir, "MS_SS_"),
    ptm_mapping_file = ptm_mapping_file,
    K_only           = FALSE,
    selected_type    = "MULTI-MSMS",
    parse_protein_accession_function = NA
  )
} else {
  message("File does not exist: ", mq_evidence_data)
}

gc()

```

# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
