---
title: "p2-08 · Analyse hydroxylation kinetics"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
   github_document:
     toc: yes
     toc_depth: 3
---

# Overview

**Purpose:** Analyse and visualise hydroxylation across conditions — hypoxia
vs normoxia, kinetics (t50), cross-dataset XIC-vs-stoichiometry, neighbouring-site
interaction, and O2 / dox sensitivity.

**Inputs:** MS_KR_1 / MS_SS / PNAS stoichiometry tables from p2-01; XIC data (`data/xic_MS_SS.csv`); reference proteome.

**Outputs:** figures (rendered on knit).

**Upstream:** p2-01. **Downstream:** none.

# Setup

```{r setup, message = FALSE, warning = FALSE}

# setwd("~/mdc_fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/")
# Sys.setenv(USE_BUNDLED_LIBUV = "1")
# renv::restore(
#   lockfile = "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R/renv.lock",
#   prompt = FALSE
# )


## Resolve the repository root (via the .here sentinel) and load the shared
## setup: packages, helper functions, ggplot/knitr settings, and project paths.
repo_root <- local({
  p <- normalizePath(getwd(), winslash = "/", mustWork = FALSE)
  while (!file.exists(file.path(p, ".here")) && !identical(dirname(p), p)) p <- dirname(p)
  p
})
source(file.path(repo_root, "R", "functions", "_setup.R"))

```

# Load and preprocess data

Loads the MS_KR_1 / MS_SS / PNAS stoichiometry tables and harmonises sample labels.
The grouped zero-fill / reshape used below is the shared `contrast_hydroxylation()`
helper in `R/functions/2-useful_functions.R` (p2-06 uses the same helper, grouped by
oxygen status).

```{r load_data}

# Load and preprocess the MS_KR_1, MS_SS and PNAS stoichiometry tables
# (see load_stoichiometry_datasets() in R/functions/2-useful_functions.R).
.stoic_data       <- load_stoichiometry_datasets(results.dir, data.dir)
ms_kr1_stoic_dt   <- .stoic_data$MS_KR1_stoic_dt
ms_ss_stoic_dt    <- .stoic_data$MS_SS_stoic_dt
pnas2022_stoic_dt <- .stoic_data$pnas2022_stoic_dt
pnas2022_dt       <- .stoic_data$pnas2022_dt

```


```{r preprocess}

# Remove BRD2/3 samples reporting BRD4 stoichiometry and vice versa: these are
# cross-protein false positives (e.g. from shared / homologous peptides).
ms_ss_stoic_dt <- ms_ss_stoic_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Combine MS_SS and MS_KR_1 data (PNAS optionally added below)
ms_ss_kr_pnas_dt <- rbindlist(list(ms_ss_stoic_dt,
                         ms_kr1_stoic_dt#,
                         #pnas2022_stoic_dt
                         ),
                         use.names = TRUE)

ms_ss_kr_pnas_dt <- ms_ss_kr_pnas_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Column "sample group" created to simplify sample names
ms_ss_kr_pnas_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

# Normoxia vs hypoxia

Plots per-site stoichiometry under normoxia vs hypoxia with JMJD6 re-expression.

```{r merge_di_sites}

# Diagnostic-ion (or PNAS-curated) sites across MS_SS, MS_KR_1 and PNAS
di_sites_dt <- rbind(
  ms_ss_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  ms_kr1_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_dt[curated_oxK_site == TRUE, .(protein_accession, aa_pos)]
)
di_sites_dt <- di_sites_dt[!duplicated(paste(protein_accession, aa_pos))]

# Keep only those unique diagnostic-ion sites
ms_ss_kr_pnas_dt <- merge(
  di_sites_dt,
  ms_ss_kr_pnas_dt[aa == "K"],
  by = c("protein_accession", "aa_pos")
)

# Using contrast function on MS_SS_KR raw stoic data
c_ms_ss_kr_pnas_dt <- contrast_hydroxylation(ms_ss_kr_pnas_dt, group_col = "sample_group")

# Factor the gene names
c_ms_ss_kr_pnas_dt[, `:=`(
  gene_name =  
    factor(gene_name, levels = c("BRD2", "BRD3", "BRD4"))
)]


```


```{r reshape_long}

# Change wide format to long format 
lc_ms_ss_kr_pnas_dt <- melt(
  c_ms_ss_kr_pnas_dt,
  id.vars = c("protein_accession", "gene_name", "aa_pos"),
  value.name = "Stoichiometry", 
  variable.name = "sample_group"
  #value.var = grep("_", colnames(c_MS_SS_KR_all_dt), value = TRUE)
)

# Separate the contents of sample_group column into new columns 
lc_ms_ss_kr_pnas_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2] %>%
    factor(levels = c("Inf", "0h", "4h", "8h", "18h", "24h")),
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# reorder the oxygen levels
lc_ms_ss_kr_pnas_dt <-  lc_ms_ss_kr_pnas_dt[, oxygen := factor(
  oxygen,
  levels = c("01pc", "1pc", "4pc", "21pc")
)]

# subset rows that do not contain NA values 
lc_ms_ss_kr_pnas_dt <- lc_ms_ss_kr_pnas_dt[!is.na(Stoichiometry)]

# counting the data per sample and size 
lc_ms_ss_kr_pnas_dt[, data_size_per_sample := .N, by = list(sample_group)]
lc_ms_ss_kr_pnas_dt[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]

```

# Normoxia dox time course

Examines stoichiometry under normoxia across the doxycycline induction time course.

```{r plot_dox_timecourse_normoxia}

# Subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- lc_ms_ss_kr_pnas_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stoichiometry")

# Sub-setting and combining data tables 
long_pc2wt_21pc <- rbind(
  lc_ms_ss_kr_pnas_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  lc_ms_ss_kr_pnas_dt[sample_group == "WT_Inf_21pc_KR"]
)
long_pc2wt_21pc <- long_pc2wt_21pc[sample_group != "iJ6_0h_21pc_SS"]

# Here, we combine induction times 18h and 24h into category 18-24h 
lc_ms_ss_kr_pnas_dt[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  induction == "24h" ~ "18-24h",
  #induction == "Inf" ~ "18-24h",
  TRUE ~ induction
)][, induction2 := factor(induction2, levels = c("Inf", "0h", "4h", "8h", "18-24h"))]

# Count data per protein accession and aa position
plot_long_pc2wt_21pc <- copy(long_pc2wt_21pc)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
plot_long_pc2wt_21pc <- plot_long_pc2wt_21pc[data_size_per_site == max(plot_long_pc2wt_21pc[, data_size_per_site])] # keep only max value of data

# Keep only sites which are WT and stoichiometry > 0 
plot_long_pc2wt_21pc <- plot_long_pc2wt_21pc[
  paste(protein_accession, aa_pos) %in% plot_long_pc2wt_21pc[cell == "WT"][Stoichiometry > 0, paste(protein_accession, aa_pos)]
]

# Boxplot - comparing normoxia in different dox incubation timings
ggplot(
  plot_long_pc2wt_21pc,
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 1.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggtitle("All the sites")

plot_long_pc2wt_21pc[, .N, by = induction]


# Plot ratio
# Merge MS_SS normoxia data with MS_KR_1 WT normoxia data
pc2wt_21pc <- merge(
  lc_ms_ss_kr_pnas_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)

# Subset data to WT_stoichiometry > 0 and exclude sample group = iJ6_0h_21pc_SS
pc2wt_21pc_sub <- pc2wt_21pc[WT_stoichiometry > 0][sample_group != "iJ6_0h_21pc_SS"]

pc2wt_21pc_sub <- copy(pc2wt_21pc_sub)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
pc2wt_21pc_sub <- pc2wt_21pc_sub[data_size_per_site == max(pc2wt_21pc_sub[, data_size_per_site])]

# Boxplot - comparing normoxia in different dox incubation timings (relative to WT stoichiometry)
ggplot(
  data = pc2wt_21pc_sub,
  aes(
    x = induction,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Create data table  
induction_t_test <- data.table()

# One sample two-sided t.test compared to mu = 1
for(i in pc2wt_21pc_sub[, unique(induction)]){
  t.out <- t.test(
    pc2wt_21pc_sub[induction == i][, Stoichiometry / WT_stoichiometry],
    mu = 1,
    alternative = "two.sided"
  )
  induction_t_test <- rbind(
    induction_t_test,
    data.table(induction = i, p_value = t.out$p.value, n = nrow(pc2wt_21pc_sub[induction == i]))
  )  
}

# Adjusted P-value
induction_t_test[, padj := p.adjust(p_value, method = "holm")]
induction_t_test[, significance := case_when(
  padj < 0.005 ~ "**",
  padj < 0.05 ~ "*",
  TRUE ~ "N.S."
)]

induction_t_test

```

# t50 (time to half-maximal)

Estimates t50 — the induction time to reach half-maximal stoichiometry — per site.

```{r compute_t50}

# helper: convert time to numeric and replace Inf by a large finite value
time_to_num <- function(x, inf_value = 100) {
  x_chr <- as.character(x)
  out <- suppressWarnings(as.numeric(gsub("[^0-9.]+", "", x_chr)))
  out[grepl("^inf$", x_chr, ignore.case = TRUE)] <- inf_value
  out[is.infinite(out)] <- inf_value
  out
}

# Calculate t50 based on induction time 
calc_t50 <- function(dt, time_col = "induction", y_col = "Stoichiometry", inf_value = 100) {
  t <- time_to_num(dt[[time_col]], inf_value = inf_value)
  y <- as.numeric(dt[[y_col]])

  keep <- is.finite(t) & is.finite(y)
  t <- t[keep]
  y <- y[keep]

  if (length(t) < 2L || uniqueN(t) < 2L || all(y <= 0, na.rm = TRUE)) {
    return(NA_real_)
  }

  # average replicates at the same time
  d <- data.table(t = t, y = y)[, .(y = mean(y, na.rm = TRUE)), by = t][order(t)]

  # monotone increasing fit
  iso <- isoreg(d$t, d$y)
  y_fit <- iso$yf

  ymax <- max(y_fit, na.rm = TRUE)
  if (!is.finite(ymax) || ymax <= 0) return(NA_real_)

  y_half <- ymax / 2

  # first point where fitted curve reaches half-max
  idx <- which(y_fit >= y_half)[1]
  if (is.na(idx)) return(NA_real_)

  if (idx == 1L) return(d$t[1])

  t1 <- d$t[idx - 1L]
  t2 <- d$t[idx]
  y1 <- y_fit[idx - 1L]
  y2 <- y_fit[idx]

  if (y2 == y1) return(mean(c(t1, t2)))

  t1 + (y_half - y1) * (t2 - t1) / (y2 - y1)
}

# Calculate t50 for normoxia data 
t50_dt <- lc_ms_ss_kr_pnas_dt[oxygen == "21pc"][, .(
  t50 = calc_t50(.SD, time_col = "induction", y_col = "Stoichiometry", inf_value = 100),
  n = .N
), by = .(gene_name, aa_pos)]

t50_dt[, `:=`(
  t50_bin = case_when(
    t50 <= 4 ~ "0-4h",
    t50 <= 8 ~ "4-8h",
    TRUE ~ ">8h"
  ) %>% factor(levels = c("0-4h", "4-8h", ">8h"))
)]

t50_dt

```

# Kinetics

Summarises per-site hydroxylation kinetics (t50 distributions across BRD2/3/4).

```{r kinetics}

# Kinetics uses the induction time course excluding 0h and 18h, and requires
# detectable WT stoichiometry (> 0).
kinetics_data <- pc2wt_21pc[!(induction %in% c("0h", "18h"))][WT_stoichiometry > 0]

# Boxplot - WT stoichiometry versus normalized stoichiometry 
ggplot(
  data = kinetics_data,
  aes(
    x = WT_stoichiometry,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_point() +
  facet_grid(~ induction2) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))

# Spearman correlation between WT stoichiometry and normalized stoichiometry 
kinetics_out <- data.table()

for(i2 in kinetics_data[, unique(induction2)]){
  c1 <- kinetics_data[induction2 == i2] %$%
    cor.test(x = WT_stoichiometry, y = Stoichiometry / WT_stoichiometry, method = "spearman")
  
  kinetics_out <- rbind(kinetics_out, data.table(induction2 = i2, rho = c1$estimate, p_val = c1$p.value, n = nrow(kinetics_data[induction2 == i2])))
}

# Adjusted p-value
kinetics_out[, padj := p.adjust(p_val, method = "holm")] # Holm correction for multiple comparisons

kinetics_out


kinetics_data <- merge(
  kinetics_data,
  t50_dt,
  by = c("gene_name", "aa_pos")
)

# Boxplot - half-time versus stoichiometry 
ggplot(
  data = kinetics_data[!duplicated(paste(gene_name, aa_pos))],
  aes(
    x = t50_bin,
    y = WT_stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

kinetics_data[!duplicated(paste(gene_name, aa_pos))][, table(gene_name, t50_bin)]

```

# XIC vs stoichiometry

Benchmarks extracted-ion-chromatogram (XIC) intensities against precursor-intensity stoichiometry across datasets.

```{r compare_xic_stoichiometry}

# Load XIC data
xic_ms_ss_dt <- fread(file.path(data.dir, "XIC_MS_SS/xic_MS_SS.csv"))
xic_ms_ss_dt[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  TRUE ~ induction
)]

on_lc_ms_ss_kr_pnas_dt <- copy(lc_ms_ss_kr_pnas_dt)[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  induction == "24h" ~ "18-24h",
  #induction == "Inf" ~ "18-24h",
  TRUE ~ induction
)]

# on_lc_ms_ss_kr_pnas_dt <- on_lc_ms_ss_kr_pnas_dt[induction != "24h"]

# Merge XIC data and stoichiometry data
xic_stoic <- merge(on_lc_ms_ss_kr_pnas_dt,
                   xic_ms_ss_dt,
                   by = c("induction2", "oxygen", "gene_name", "aa_pos"))

xic_stoic[, XIC := XIC/100]

# Scatter plot - XIC vs stoichiometry
ggplot(
  xic_stoic,
  aes(
    x = XIC,
    y = Stoichiometry
  )
) + 
  geom_point(size = 2) +
  theme_classic_2() +
  theme(aspect.ratio = 1) +
  coord_cartesian(xlim = c(0, 0.7), ylim = c(0, 0.7)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))

xic_stoic %$%
cor.test(
  XIC,
  Stoichiometry
)

m_xic_stoic <- melt(
  xic_stoic,
  id.vars = c("gene_name", "aa_pos", "oxygen", "induction2"),
  measure.vars = c("Stoichiometry", "XIC"),
  variable.name = "type",
  value.name = "stoichiometry"
)

m_xic_stoic[, `:=`(
  oxygen = factor(oxygen, levels = rev(c("1pc", "4pc", "21pc"))),
  induction2 = factor(induction2, levels = c("4h", "8h", "18h"))
)]

```


```{r plot_xic_per_pos, fig.width=14, fig.height=4.5}

ggplot(
  m_xic_stoic,
  aes(
    x = induction2,
    y = stoichiometry,
    color = type
  )
) + 
  facet_grid(~ gene_name + aa_pos + oxygen, scale = "free", space = "free") +
  geom_point(size = 4) +
  theme_classic_2() +
  coord_cartesian(ylim = c(0, 0.7)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = c("XIC" = "black", "Stoichiometry" = "darkred"))

```

# Neighbouring-site interaction

Tests for interdependence between neighbouring hydroxylation sites: within peptides covering a pair of lysines, is the Hyl stoichiometry of the higher-t50 residue different when the neighbouring lower-t50 residue is hydroxylated?

```{r analyse_site_interaction}

# Apparent kinetics (t50) of the BRD4 lysine-rich region, used to assign the
# residues below to lower- and higher-t50 categories.
t50_dt[gene_name == "BRD4"][aa_pos > 534 & aa_pos < 560]

# Hydroxylation assignments supported by diagnostic immonium ions
di_ptms <- c("[Oxidation (K) DI]", "[Oxidised Propionylation (K) DI]")

# For each pair of a lower- and a higher-t50 residue, compare the Hyl
# stoichiometry of the higher-t50 residue between peptides in which the
# lower-t50 residue is hydroxylated and peptides in which it is not.
calc_neighbouring_site_interaction <- function(per_pos_data, site_pairs, sample_group) {

  # One row per (peptide, residue), restricted to the residues of interest.
  site_dt <- per_pos_data[
    protein_accession %in% site_pairs$protein_accession &
      aa_pos %in% c(site_pairs$low_t50_aa_pos, site_pairs$high_t50_aa_pos),
    .(peptide_id, protein_accession, aa_pos, peak_intensity, hyl = ptm %in% di_ptms)
  ]

  summarise_pair <- function(acc, low_pos, high_pos) {

    pair_dt <- site_dt[protein_accession == acc & aa_pos %in% c(low_pos, high_pos)]

    # Use only peptides covering both residues, so that the two groups compared
    # below differ solely in the modification state of the lower-t50 residue.
    # This is what excludes pairs whose residues fall in different tryptic
    # peptides. The `< 2` guard only stops the with/without split degenerating
    # to a single peptide; it does not bind on these data, where every retained
    # pair is supported by 13-157 peptides and every excluded pair by none.
    both_peptides <- pair_dt[, .N, by = peptide_id][N == 2L, peptide_id]
    if (length(both_peptides) < 2L) return(NULL)
    pair_dt <- pair_dt[peptide_id %in% both_peptides]

    # Intensity-weighted Hyl stoichiometry of the higher-t50 residue
    high_t50_stoic <- function(peptides) {
      x <- pair_dt[aa_pos == high_pos & peptide_id %in% peptides]
      x[hyl == TRUE, sum(peak_intensity)] / x[, sum(peak_intensity)]
    }
    low_t50_hyl_peptides <- pair_dt[aa_pos == low_pos & hyl == TRUE, peptide_id]

    data.table(
      protein_accession   = acc,
      low_t50_aa_pos      = low_pos,
      high_t50_aa_pos     = high_pos,
      stoic_all           = high_t50_stoic(both_peptides),
      stoic_low_t50_hyl   = high_t50_stoic(low_t50_hyl_peptides),
      stoic_low_t50_unmod = high_t50_stoic(setdiff(both_peptides, low_t50_hyl_peptides)),
      n_peptides          = length(both_peptides),
      sample_group        = sample_group
    )
  }

  # Pairs without enough supporting peptides, or with no hydroxylation at the
  # higher-t50 residue, drop out here.
  out <- rbindlist(Map(
    summarise_pair,
    site_pairs$protein_accession, site_pairs$low_t50_aa_pos, site_pairs$high_t50_aa_pos
  ))
  out[stoic_all > 0]
}

# Candidate BRD4 residues by t50 category. K535 is intermediate and therefore
# appears in both lists: it is the higher-t50 partner of K538 and the lower-t50
# partner of K537. Self-pairs are dropped.
site_pairs <- CJ(
  low_t50_aa_pos  = c(535L, 538L, 548L),
  high_t50_aa_pos = c(535L, 537L, 550L, 552L)
)[low_t50_aa_pos != high_t50_aa_pos][, protein_accession := "O60885"][]

# Datasets contributing to the comparison: three independent experiments, with
# MS_KR_1 providing both a WT and a JMJD6-induced sample.
brd_genes <- c("BRD2", "BRD3", "BRD4")
ms_kr1_per_pos <- fread(file.path(results.dir, "p2-analysis-setting/MS_KR_1/MS_KR_1_all_per_pos_data.csv"))[gene_name %in% brd_genes]
ms_ss_per_pos <- fread(file.path(results.dir, "p2-analysis-setting/MS_SS/MS_SS_all_per_pos_data.csv"))

per_pos_data_list <- list(
  # PNAS data-A (MQ_DI, no water loss)
  WT_PNAS2022 = fread(file.path(results.dir, "p2-analysis-setting/MQ_DI/DI_data-A_trp_m7_v7_def_all_per_pos_data.csv"))[
    gene_name %in% brd_genes & sample_name == "JQ1_HeLaWT_derivatised"],
  WT = ms_kr1_per_pos[sample_name == "HeLaWT_NA_N_NA"],
  `iJMJD6 24h` = ms_kr1_per_pos[sample_name == "HeLaiJMJD6_Dox_N_NA"],
  # In MS_SS, BRD4 and BRD2/3 were acquired in separate runs
  `iJMJD6 18h` = rbind(
    ms_ss_per_pos[gene_name == "BRD4" & sample_name == "18h_21pc_BRD4"],
    ms_ss_per_pos[gene_name %in% c("BRD2", "BRD3") & sample_name == "18h_21pc_BRD23"]
  )
)

all_interaction_summary <- rbindlist(Map(
  function(per_pos_data, sample_group) {
    calc_neighbouring_site_interaction(per_pos_data, site_pairs, sample_group)
  },
  per_pos_data_list, names(per_pos_data_list)
))

# Per-pair paired t-tests across datasets (exploratory, unadjusted; pairs
# quantified in fewer than three datasets are skipped).
interaction_out_data <- all_interaction_summary[
  , if (.N > 2) {
    .(p_val = t.test(stoic_low_t50_unmod, stoic_low_t50_hyl, paired = TRUE)$p.value, n = .N)
  },
  by = .(protein_accession, low_t50_aa_pos, high_t50_aa_pos)
]


# Paired t-test pooling all residue pairs and datasets (reported in the paper)
all_interaction_summary %$%
  t.test(stoic_low_t50_unmod, stoic_low_t50_hyl, paired = TRUE)

# The 20 values above are 5 residue pairs measured in each of 4 samples, and so
# are not mutually independent. As a robustness check, average the samples
# within each residue pair so that every pair contributes one value, and repeat
# the test (also reported in the paper).
all_interaction_summary[
  , .(stoic_low_t50_hyl = mean(stoic_low_t50_hyl),
      stoic_low_t50_unmod = mean(stoic_low_t50_unmod)),
  by = .(protein_accession, low_t50_aa_pos, high_t50_aa_pos)
] %$%
  t.test(stoic_low_t50_unmod, stoic_low_t50_hyl, paired = TRUE)

# Convert table into long format for plotting
m_all_interaction_summary <- melt(
  all_interaction_summary,
  id.vars = c("protein_accession", "low_t50_aa_pos", "high_t50_aa_pos", "n_peptides", "sample_group"),
  measure.vars = c("stoic_low_t50_hyl", "stoic_low_t50_unmod"),
  variable.name = "type",
  value.name = "stoichiometry"
)

m_all_interaction_summary[
  , type := factor(type, labels = c("low_t50_hyl", "low_t50_unmodified"))]

# Plot
ggplot(
  data = m_all_interaction_summary,
  aes(
    x = type,
    y = stoichiometry,
    group = paste(sample_group, low_t50_aa_pos, high_t50_aa_pos),
    color = paste(low_t50_aa_pos, high_t50_aa_pos)
  )
) +
  geom_line(color = "black") +
  geom_point(size = 4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_bright() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme_classic_2() +
  theme(aspect.ratio = 4)


```


# Oxygen sensitivity

Compares hydroxylation under hypoxia across doxycycline incubation times to assess oxygen sensitivity.

```{r plot_oxygen_sensitivity}

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
# Data subset to overnight (18/24h) dox induction

on_data <- lc_ms_ss_kr_pnas_dt[induction2 %in% c("18-24h", "Inf")][
  sample_group != "iJ6_18h_21pc_SS" # As iJ6_24h_21pc_KR exists
]

#---------------------------------------------
# Stoichiometry relative to WT values
#---------------------------------------------

# subset data to WT normoxia (MS_KR_1 sample)
wt_on <- on_data[induction2 == "Inf", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_on, old = "Stoichiometry", "WT_stoichiometry")

# Merge MS_SS normoxia data with MS_KR_1 WT normoxia data
pc2wt_on <- merge(
  on_data[induction2 != "Inf"],
  wt_on,
  by = c("protein_accession", "aa_pos")
)

pc2wt_on <- copy(pc2wt_on)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
pc2wt_on <- pc2wt_on[data_size_per_site == max(pc2wt_on[, data_size_per_site])]

pc2wt_on <- pc2wt_on[WT_stoichiometry > 0]
pc2wt_on[, `:=`(
  oxygen = factor(oxygen, c("21pc", "4pc", "1pc", "01pc"))
)]

ggplot(
  data = pc2wt_on,
  aes(
    x = oxygen,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggtitle("Min WT stoichiometry > 0")

pc2wt_on[, .N, by = oxygen]


# Boxplot - comparing normoxia versus hypoxia
ggplot(
  data = pc2wt_on,
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

oxygen_t_test <- data.table()

for(i in pc2wt_on[, unique(oxygen)]){
  t.out <- t.test(
    pc2wt_on[oxygen == i][, Stoichiometry/WT_stoichiometry],
    pc2wt_on[oxygen == "21pc"][, Stoichiometry/WT_stoichiometry],
    paired = TRUE
  )
  oxygen_t_test <- rbind(
    oxygen_t_test,
    data.table(induction = i, p_value = t.out$p.value, n = nrow(pc2wt_on[data_size_per_site == max(pc2wt_on[, data_size_per_site])][oxygen == i]))
  )  
}

oxygen_t_test[, padj := p.adjust(p_value, method = "holm")]
oxygen_t_test[, significance := case_when(
  padj < 0.005 ~ "**",
  padj < 0.05 ~ "*",
  TRUE ~ "N.S."
)]

oxygen_t_test


```

# Baseline vs hypoxic change

Relates baseline (WT/normoxia) stoichiometry to the magnitude of hypoxic suppression per site.

```{r plot_baseline_vs_hypoxia}

ggplot(
  data = pc2wt_on[oxygen != "21pc"],
  aes(
    x = WT_stoichiometry,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(size = 2) +
  facet_grid(~ oxygen) +
  theme(
    aspect.ratio = 1,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))


oxygen_out <- data.table()

for(o2 in pc2wt_on[oxygen != "21pc"][, unique(oxygen)]){
  c2 <- pc2wt_on[oxygen == o2] %$%
    cor.test(
      WT_stoichiometry, Stoichiometry / WT_stoichiometry, method = "spearman"
    )
  
    oxygen_out <- rbind(oxygen_out, data.table(oxygen = o2, rho = c2$estimate, p_val = c2$p.value, n = nrow(pc2wt_on[oxygen == o2])))
}

oxygen_out[, padj := p.adjust(p_val, method = "holm")]

oxygen_out

```


```{r plot_hypoxic_sensitivity}


# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
on_01pc_ms_kr1_data <- lc_ms_ss_kr_pnas_dt[induction2 %in% c("18-24h", "Inf")][
  sample_group == "iJ6_24h_01pc_KR"
]
setnames(on_01pc_ms_kr1_data, old = "Stoichiometry", "ON01_stoichiometry")

wt_on <- lc_ms_ss_kr_pnas_dt[sample_group == "WT_Inf_21pc_KR"]
setnames(wt_on, old = "Stoichiometry", "WT_stoichiometry")

merge_dts <- function(dt_list, by_columns){Reduce(function(...) merge(..., by = by_columns), dt_list)}

sl_columns <- c("protein_accession", "aa_pos")

ref_dts <- merge_dts(
  dt_list = list(
    on_01pc_ms_kr1_data[, c(sl_columns, "ON01_stoichiometry"), with = FALSE],
    wt_on[, c(sl_columns, "WT_stoichiometry"), with = FALSE]
  ),
  by = sl_columns
)

non_ref_lc_ms_ss_kr_pnas_dt <- lc_ms_ss_kr_pnas_dt[
  !(sample_group %in% c("iJ6_24h_01pc_KR", "WT_Inf_21pc_KR", "iJ6_0h_21pc_SS", "iJ6_0h_21pc_KR", "iJ6_18h_21pc_SS"))
]

induction2vsref_21pc <- merge(
  non_ref_lc_ms_ss_kr_pnas_dt,
  ref_dts,
  by = sl_columns
)

induction2vsref_21pc <- induction2vsref_21pc[WT_stoichiometry > 0]
## induction2vsref_21pc <- induction2vsref_21pc[sample_group %in% c("iJ6_4h_21pc_SS", "iJ6_8h_21pc_SS", "iJ6_24h_21pc_KR")]
induction2vsref_21pc[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
induction2vsref_21pc <- induction2vsref_21pc[data_size_per_site == max(induction2vsref_21pc[, data_size_per_site])]

induction2vsref_21pc[, .N, by = list(sample_group, gene_name)][order(sample_group)]

ggplot(
  data = induction2vsref_21pc,
  aes(
    x = ON01_stoichiometry,
    y = Stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_abline(intercept = 0, slope = 1) +
  geom_point() +
  facet_grid(oxygen ~ induction2) +
  theme(aspect.ratio = 1) +
  coord_cartesian(ylim = c(0, 1), xlim = c(0, 1))

rsq_out_dt <- data.table()

for(o2 in induction2vsref_21pc[, unique(oxygen)]){
  for(i2 in induction2vsref_21pc[, unique(induction2)]){
    rsq_dt <- induction2vsref_21pc[
      oxygen == o2 & induction2 == i2
    ]
    rsq_dt[, mean_St := mean(Stoichiometry)]
    rsq <- 1 - sum(rsq_dt[, (Stoichiometry - ON01_stoichiometry)^2]) / sum(rsq_dt[, (Stoichiometry - mean_St)^2])
    rsq_out_dt <- rbind(
      rsq_out_dt, data.table(oxygen = o2, induction2 = i2, rsq = rsq, n = nrow(rsq_dt))
    )
  }
}

rsq_out_dt

pc2wt_on_kinetics <- merge(
  pc2wt_on,
  t50_dt,
  by = c("gene_name", "aa_pos")
)

library("ggbeeswarm")

ggplot(
  pc2wt_on_kinetics[n == 7][oxygen != "21pc"],
  aes(
    x = t50,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_point(size = 2) +
  facet_grid(~ oxygen) +
  theme(
    aspect.ratio = 1,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  coord_cartesian(xlim = c(0, 24)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

ggplot(
  pc2wt_on_kinetics[n == 7][oxygen != "21pc"],
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  geom_beeswarm() +
  facet_grid(~ t50_bin) +
  theme(
    aspect.ratio = 3,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  coord_cartesian(ylim = c(0, 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# QC
ggplot(
  pc2wt_on_kinetics[n == 7],
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ t50_bin) +
  theme(
    aspect.ratio = 3,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  coord_cartesian(ylim = c(0, 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


pc2wt_on_kinetics[n == 7][, .N, by = list(oxygen, t50_bin)]

```

# Session information

```{r session_info}

sessioninfo::session_info()

```
