---
title: "2-1. Calculate PTM stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

# This script calculates the 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Create project directory - contains the R scripts, data and result folders
 project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
)

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```


# Install,load essential functions and libraries
```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
temp <- sapply(list.files(file.path(project.dir, "R/functions"), pattern="*.R", full.names = TRUE), source)

## Install private packages 
# Install ptm.stiochiometry package - package installed 24.09.2025
#install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries - ptm.stiochiometry and readxl
library(ptm.stoichiometry)
library("readxl")

```

# Import data 

```{r import_data}

# Import human protein reference data from specified file path 
ref_protein_data <- import_reference_fasta(file.path
("/fast/AG_Sugimoto/reference/uniprot/human", 
  "UP000005640_9606.fasta")) 

## Create results directory within the project directory (project.dir)
#  p2_analysis in results, results in 
p2_results_dir <- file.path(project.dir,
                        "results",
                        "p2_analysis_setting")

dir.create(p2_results_dir, recursive = TRUE, showWarnings = FALSE)

# Create a other results directory to check the reflections of code changes. 
# This is a test results directory and will be removed or moved eventually

p2_results_test_dir <- file.path(project.dir,
                        "results_test",
                        "p2_analysis_setting_test")

dir.create(p2_results_test_dir, recursive = TRUE, showWarnings = FALSE)

```


# Calculation of Stoichiometry without diagnostic ion consideration 
```{r data_directory_wo_di}

# Load data into environment
pnas2022_data <- file.path(
  project.dir,
  "data/MQ_standard/PNAS2022" 
)

# Load sample info data 
all_sample_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_standard"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each row (1:N), 
#where N is the total number of rows in the data.table
all_sample_run_info[, sample_id := 1:.N]

```

```{r calculate_stoichiometry_wo_di}

## Stochiomtery analysis using with MULTI-MSMS combined with 
# MULTI-SECPEP (secondary peptide) and MULTI-MSMS only

# Retrieve data id along with its matching sample ID for data-A and -D
# Note: data-A and -D are considered, as these samples contain fixed PTM of 
#carbamidomethyl
for(i in all_sample_run_info[data %in% 
                             c("data-A", "data-D"), sample_id]){
  
# Build full path to MaxQuant evidence file
   mq_evidence_data = file.path(pnas2022_data, 
                               "evidence",
                               paste0(all_sample_run_info[sample_id == i, 
                                                          prefix], 
                                      "evidence.txt"))
   
# Print the name of the evidence file currently being processed
   print(paste0("Processing: ", basename(mq_evidence_data)))
  
# Calculate stochiomtery for MULTI-MSMS and MULTI-SECPEP
    stoic.dt <- calculate_stoichiometry(
      mq_evidence_data = mq_evidence_data,
      sample_info_file = file.path( 
        pnas2022_data,
        "sample_info",
          paste0(
          "MS_dataset_overview_PXD031221_",
          all_sample_run_info[sample_id == i, data],
          ".csv"
        )
      ),
      ref_protein_dt = ref_protein_data,
      output_prefix = file.path(p2_results_dir, paste0(
        "including_SECPEP_", all_sample_run_info[sample_id == i, prefix] 
        )),
    ptm_mapping_file = file.path(
      project.dir, 
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
      ),
      K_only = FALSE,
      selected_type = NA
    )
  
  gc() # Run garbage collection to free up unused memory
  
}


## Code simplification_test! 
# To improve code efficiency and run faster 
samples_to_process <- all_sample_run_info[data %in% c("data-A", "data-D")]
ptm_mapping_file <- file.path(
  project.dir,
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
)

for (i in samples_to_process[, sample_id]) {
  sample_row <- samples_to_process[sample_id == i]   

  prefix    <- sample_row$prefix[1]
  data_name <- sample_row$data[1]

  mq_evidence_data <- file.path(pnas2022_data, "evidence", paste0(prefix, 
                                                                "evidence.txt"
                                                                  ))
  message("Processing: ", basename(mq_evidence_data))
  
  if(file.exists(mq_evidence_data)) {
  
  sample_info_file <- file.path(
    pnas2022_data, "sample_info",
    paste0("MS_dataset_overview_PXD031221_", data_name, ".csv")
  )
  output_prefix <- file.path(p2_results_test_dir, paste0("including_SECPEP_", 
                                                         prefix))

 stoic_MSMS_SECEP_2.dt <- calculate_stoichiometry(
    mq_evidence_data = mq_evidence_data,
    sample_info_file = sample_info_file,
    ref_protein_dt   = ref_protein_data,
    output_prefix    = output_prefix,
    ptm_mapping_file = ptm_mapping_file,
    K_only           = FALSE,
    selected_type    = NA
  )
 
   else {
    print(paste0("File does not exist: ", mq_evidence_data))
   }
 
    gc() # Run garbage collection to free up unused memory  
  }  
}


# Calculate stochiomtery for MULTI-MSMS only

ptm_mapping_file <- file.path(
  project.dir,
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv")

for(i in 1:nrow(all_sample_run_info)) {

  mq_evidence_data = file.path(pnas2022_data, "evidence", 
                               paste0(all_sample_run_info[i, prefix], 
                                      "evidence.txt"))
    message("Processing: ", basename(mq_evidence_data))

  sample_info_file = file.path(
    pnas2022_data,"sample_info",
        paste0("MS_dataset_overview_PXD031221_",  all_sample_run_info[i, data],
               ".csv"))
  if(file.exists(mq_evidence_data)) {
  
   stoic_MSMS.dt <- calculate_stoichiometry(
    mq_evidence_data = mq_evidence_data,
    sample_info_file = sample_info_file,
    ref_protein_dt   = ref_protein_data,
    output_prefix    = output_prefix,
    ptm_mapping_file = ptm_mapping_file,
    K_only           = FALSE,
    selected_type    = c("MULTI-MSMS")
  )
   
     else {
    print(paste0("File does not exist: ", mq_evidence_data))
   }
   
   gc()
  }
}

```

# Calculation of Stoichiometry with diagnostic ion consideration 

```{r data_directory_w_di}

# Load data into environment
pnas2022_DI_data <- file.path(
  project.dir,
  "data/MQ_DI_output/PNAS2022" 
)

# Load sample info data 
di_sample_run_info <- read_excel(
  file.path(pnas2022_DI_data, "PXD031221_sample_matrix.xlsx"),
  sheet = "run_setting"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
di_sample_run_info[, sample_id := 1:.N]

```

```{r calculate_stoichiometry_w_di}

for(i in 1:nrow(di_sample_run_info)) {
  
  sample_info_file = file.path(pnas2022_DI_data, "sample_info",
        paste0("MS_dataset_overview_PXD031221_", di_sample_run_info[i, data],
          ".csv"))

  mq_evidence_data = file.path(pnas2022_DI_data,
                               "evidence",
                               paste0(di_sample_run_info[i, prefix], 
                                      "evidence.txt"))
  
   message("Processing: ", basename(mq_evidence_data))
  
  if (file.exists(mq_evidence_data)) {
    
    ptm_files <- list.files(file.path(
      pnas2022_DI_data, "ptm", 
      paste0(di_sample_run_info[i, prefix], "ptm")
    ), full.names = TRUE)
    
    ptm_names <- paste0(
      "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
    )
    
    stoic_MQ_DI.dt <- calculate_stoichiometry2(
      mq_evidence_data = mq_evidence_data,
      sample_info_file = sample_info_file,
      ref_protein_dt = ref_protein_dt,
      ptm_files = ptm_files,
      ptm_names = ptm_names,
      output_prefix = file.path(p2.res.dir, paste0("DI_", di_sample_run_info[i, 
                                                                  prefix])),
      ptm_mapping_file = file.path(
        project.dir, 
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
      ),
      K_only = FALSE,
      selected_type = c("MULTI-MSMS")
    )
  } else {
    print(paste0("File does not exist: ", mq_evidence_data))
  }
  
  gc()
}

```

# Calculation of Stoichiometry with diagnostic ion and iterative 

```{r data_directory_w_di_it}

```

```{r calculate_stoichiometry_w_di_it}

```


# Calculation of Stoichiometry with diagnostic ion, iterative and additional PTMs  

```{r data_directory_w_di_it_aPTMs}

```

```{r alculate_stoichiometry_w_di_it_aPTMs}

```


# Session information

```{r sessionInfo}

## sessioninfo::session_info()

```
