---
title: "2-1. Calculate PTM stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

# This script calculates the 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )


project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```


# Install and load essentials functions 
```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
temp <- sapply(list.files(file.path(project.dir, "R/functions"), pattern="*.R", full.names = TRUE), source)

## Install private packages 
# Install ptm.stiochiometry package
install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries - ptm.stiochiometry
library(ptm.stoichiometry)

```

# Import data 

```{r import_data}

# Import human protien reference data from specified file path 
ref_protein_data <- import_reference_fasta(file.path
("/fast/AG_Sugimoto/reference/uniprot/human", 
  "UP000005640_9606.fasta")) 

```


# Calculation of Stoichiometry without diagnostic ion consideration 
```{r data_directory_wo_di}


```

```{r calculate_stoichiometry_wo_di}

```

# Calculation of Stoichiometry with diagnostic ion consideration 

```{r data_directory_w_di}

```

```{r calculate_stoichiometry_w_di}

```

# Calculation of Stoichiometry with diagnostic ion and iterative 

```{r data_directory_w_di_it}

```

```{r calculate_stoichiometry_w_di_it}

```


# Calculation of Stoichiometry with diagnostic ion, iterative and additional PTMs  

```{r data_directory_w_di_it_aPTMs}

```

```{r alculate_stoichiometry_w_di_it_aPTMs}

```


# Session information

```{r sessionInfo}

## sessioninfo::session_info()

```
