---
title: "2-1. Calculate PTM stoichiometry"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

# This script calculates the 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Create project directory - contains the R scripts, data and result folders
 project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
)

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```


# Install,load essential functions and libraries
```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
temp <- sapply(list.files(file.path(project.dir, "R/functions"), pattern="*.R", full.names = TRUE), source)

## Install private packages 
# Install ptm.stiochiometry package - package installed 24.09.2025
#install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries - ptm.stiochiometry and readxl
library(ptm.stoichiometry)
library("readxl")

```

# Import data 

```{r import_data}

# Import human protein reference data from specified file path 
ref_protein_data <- import_reference_fasta(file.path
("/fast/AG_Sugimoto/reference/uniprot/human", 
  "UP000005640_9606.fasta")) 

## Create results directory within the project directory (project.dir)
#  p2_analysis in results, results in 
p2_results_dir <- file.path(project.dir,
                        "results",
                        "p2_analysis_setting")

dir.create(p2_results_dir, recursive = TRUE, showWarnings = FALSE)

```


# Calculation of Stoichiometry without diagnostic ion consideration 
```{r data_directory_wo_di}

# Load data into environment
pnas2022_data <- file.path(
  project.dir,
  "data/MQ_standard/PNAS2022" 
)

# Load sample info data 
all_sample_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_standard"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each row (1:N), where N is the total number of rows in the data.table
all_sample_run_info[, sample_id := 1:.N]

```

```{r calculate_stoichiometry_wo_di}

## Stochiomtery analysis using with MULTI-MSMS combined with MULTI-SECPEP (secondary peptide) and MULTI-MSMS only

# Retrieve data id along with its matching sample ID for data-A and -D
# Note: data-A and -D are considered, as these samples contain fixed PTM of carbamidomethyl
for(i in all_sample_run_info[data %in% 
                             c("data-A", "data-D", sample_id), sample_id]){
  
# Build full path to MaxQuant evidence file
   mq_evidence_data = file.path(pnas2022_data, 
                               "evidence",
                               paste0(all_sample_run_info[sample_id == i, prefix], "evidence.txt"))
   
# Print the name of the evidence file currently being processed
   print(paste0("Processing: ", basename(mq_evidence_data)))
  
# Calculate stochiomtery for MULTI-MSMS and MULTI-SECPEP
    stoic.dt <- calculate_stoichiometry(
      mq_evidence_data = mq_evidence_data,
      sample_info_file = file.path( 
        pnas2022_data,
        "sample_info",
          paste0(
          "MS_dataset_overview_PXD031221_",
          all_sample_run_info[sample_id == i, data],
          ".csv"
        )
      ),
      ref_protein_dt = ref_protein_data,
      output_prefix = file.path(p2_results_dir, paste0(
        "including_SECPEP_", all_sample_run_info[sample_id == i, prefix] 
        )),
      ptm_mapping_file = file.path(
        project.dir, 
        "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
      ),
      K_only = FALSE,
      selected_type = NA
    )
  
  gc() # Run garbage collection to free up unused memory
  
}


# Calculate stochiomtery for MULTI-MSMS only
for(i in 1:nrow(all_sample_run_info)) {

  mq_evidence_data = file.path(pnas2022_data,
                               "evidence",
                               paste0(all_sample_run_info[i, prefix], "evidence.txt"))
  
  print(paste0("Processing: ", basename(mq_evidence_data)))
  
      # Calculate stochiomtery for MULTI-MSMS 
    stoic.dt <- calculate_stoichiometry(
      mq_evidence_data = mq_evidence_data,
      sample_info_file = file.path(
        pnas2022_data,
        "sample_info",
        paste0(
          "MS_dataset_overview_PXD031221_",
          all_sample_run_info[i, data],
          ".csv"
        )
      ),
      ref_protein_dt = ref_protein_data,
      output_prefix = file.path(p2_results_dir, all_sample_run_info[i, prefix]),
      ptm_mapping_file = file.path(
        project.dir, 
        "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
      ),
      K_only = FALSE,
      selected_type = c("MULTI-MSMS")
    )
  
  gc()
  
}

```

# Calculation of Stoichiometry with diagnostic ion consideration 

```{r data_directory_w_di}

```

```{r calculate_stoichiometry_w_di}

```

# Calculation of Stoichiometry with diagnostic ion and iterative 

```{r data_directory_w_di_it}

```

```{r calculate_stoichiometry_w_di_it}

```


# Calculation of Stoichiometry with diagnostic ion, iterative and additional PTMs  

```{r data_directory_w_di_it_aPTMs}

```

```{r alculate_stoichiometry_w_di_it_aPTMs}

```


# Session information

```{r sessionInfo}

## sessioninfo::session_info()

```
