---
title: "2-2. Optimise MQ parameters"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script determines the optimal settings for the database search that analyses the LC-MS data of lysine-derivatised samples.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))
#renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)


## Install private packages 
# Install ptm.stoichiometry package
#install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries - ptm.stoichiometry and readxl
library(ptm.stoichiometry)
library("readxl")

# Define path to result directory 
results.dir <- file.path(project.dir, "results")

p2_results_dir <- file.path(results.dir,
                        "p2-analysis-setting")

```

# 2.2.1 Import basic data

```{r import_basic_data}

# Define path for data directory
data.dir <- file.path(project.dir, "data")

# Define file path for PNAS2022 data 
pnas2022_data <- file.path(
  project.dir,
  "data/PNAS2022" 
)

# Load sample run info using defined file path
all_sample_run_info <- read_excel(
  file.path(data.dir, "analysis_setting", "PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_standard"
) %>% data.table

# Load protein feature data
protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))

# Change column names
setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

```


# 2.2.2 MS/MS count by different setting

```{r MS_MS_count}

# Define "read_evidence" function 
read_evidence <- function(prefix, dir_path){
  input.file <- file.path(
    dir_path, 
    paste0("including_SECPEP_", prefix, "all_processed_evidence_data.csv")
  )
  if(file.exists(input.file)){
    dt <- fread(input.file)
    dt[, condition := gsub("_$", "", prefix)]
  } else {
    dt <- data.table()
  }
  return(dt)
}

# Apply the function 'read_evidence' to each selected 'prefix' value
evidence.dt <- lapply( # Filter data-A and data-D and retrieve corresponding prefix
  all_sample_run_info[data %in% c("data-A", "data-D") & !is.na(prefix), prefix], 
  read_evidence,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS_SECPEP")
) %>% rbindlist

data.count.dt <- evidence.dt[, list(
  total_peptide_count = .N, # Count total number of rows (peptides)
  total_msms_count = sum(psm_mapped), # Sum of 'psm_mapped'
  total_intensity = sum(peak_intensity) # Sum of 'peak_intensity'
), by = list(file_name, sample_name, condition, type)] 

data.count.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2]
)]

data_col = c("total_peptide_count", "total_msms_count", "total_intensity")

m2_v2.dt <- data.count.dt[
  MQ_setting == "m2_v2" & type == "MULTI-MSMS"
][, c("file_name", data_col), with = FALSE]
setnames(m2_v2.dt, old = data_col, new = paste0("m2_v2_", data_col))

rel_count.dt <- data.count.dt

rel_count.dt <- merge(
  rel_count.dt,
  m2_v2.dt,
  by = c("file_name")
)

rel_count.dt[, `:=`(
  rel_total_peptide_count = total_peptide_count / m2_v2_total_peptide_count,
  rel_msms_count = total_msms_count / m2_v2_total_msms_count,
  rel_intensity = total_intensity / m2_v2_total_intensity 
)]

rel_count_stat.dt <- rel_count.dt[, list(
  mean_rel_total_peptide_count = mean(rel_total_peptide_count),
  sd_rel_total_peptide_count = sd(rel_total_peptide_count),
  mean_rel_msms_count = mean(rel_msms_count),
  sd_rel_msms_count = sd(rel_msms_count),
  mean_rel_intensity = mean(rel_intensity),
  sd_rel_intensity = sd(rel_intensity)
), by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_stat.dt[, `:=`(
  protease_setting = factor(protease_setting, levels = c("trp", "argC")),
  type = factor(type, levels = c("MULTI-MSMS", "MULTI-SECPEP"))
)]


ggplot(
  rel_count_stat.dt[type == "MULTI-MSMS"],
  aes(
    x = MQ_setting,
    y = mean_rel_msms_count,
    fill = type
  )
) +
  geom_bar(stat = "identity") +
  geom_errorbar(aes(
    ymin = mean_rel_msms_count - sd_rel_msms_count,
    ymax = mean_rel_msms_count + sd_rel_msms_count
  ), width = 0.5, position = position_dodge(width = 0.9)) +
  facet_grid(~ propionylation + protease_setting, scales = "free", space = "free") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Relative MS/MS count") +
  scale_fill_manual(values = c("MULTI-MSMS" = "#4477AA")) #+
# scale_color_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE"))


```

The results indicate that the setting with 7 miscleavage and 7 variable modifications identify the highest MS/MS count.


# 2.2.3 QC by the position of propionylated lysines

```{r QC_by_propionylated_lysine_position}

read_all_per_pos_data_with_SECPEP <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0("including_SECPEP_", prefix, "all_per_pos_data.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_per_pos_with_SECPEP.dt <- lapply(
  all_sample_run_info[data %in% c("data-A") & grepl("m7_v7", prefix) & !is.na(prefix), prefix],
  read_all_per_pos_data_with_SECPEP,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS_SECPEP")
) %>% rbindlist

all_per_pos_with_SECPEP.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2],
  is_peptide_C_term = peptide_end == aa_pos,
  is_propionylated = ptm %in% c("[Propionylation]", "[Oxidised Propionylation]")
)]

all_per_K_pos_with_SECPEP.dt <- all_per_pos_with_SECPEP.dt[aa == "K"]

propionylated_K_count.dt <- all_per_K_pos_with_SECPEP.dt[
  , list(total_MS_MS_count = sum(psm_mapped)),
  by = list(is_propionylated, is_peptide_C_term, type)
]

propionylated_K_count.dt

ggplot(
  propionylated_K_count.dt,
  aes(
    x = is_peptide_C_term,
    y = total_MS_MS_count,
    fill = is_propionylated
  )
) +
  geom_bar(stat = "identity", position="fill") +
  scale_fill_manual(values = c("TRUE" = "#4477AA", "FALSE" = "#BBBBBB")) +
  facet_grid(~ type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


Based on the analyses here, we decided to focus on using MULTI-MSMS data.


# 2.2.4 The effect of MQ setting by K score

```{r MQ_setting_by_k_score}

read_stoic_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(prefix, "PTM_stoichiometry.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_stoic_pos.dt <- lapply(
  all_sample_run_info[
    data != "data-D" & 
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS")
) %>% rbindlist

all_stoic_pos.dt[, `:=`(
  MQ_setting = str_extract(condition, "m\\d+_v\\d+")
)]

ms_ms_count.dt <- all_stoic_pos.dt[
  , list(total_ms_ms_count = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos, MQ_setting)
]

ms_ms_count.dt <- merge(
  ms_ms_count.dt,
  protein.feature.dt[, .(protein_accession, aa_pos, K_ratio, K_ratio_score, IUPRED2)],
  by = c("protein_accession", "aa_pos")
)

d.ms_ms_count.dt <- dcast(
  copy(ms_ms_count.dt)[, IUPRED2 := NULL],
  protein_accession + aa_pos + K_ratio + K_ratio_score ~ MQ_setting,
  value.var = "total_ms_ms_count", fill = 0
  
)

library("RColorBrewer")
library("patchwork")

d.ms_ms_count.dt[, table(K_ratio)]

p1 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m2_v2,
    y = m5_v5
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p2 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m5_v5,
    y = m7_v7
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p1 + p2

p3 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m5_v5 - m2_v2
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25)) +
  theme(aspect.ratio = 1)

p4 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m7_v7 - m5_v5
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25))

p3 + p4

```


# 2.2.5 The effect of MQ setting on runtime

```{r MQ_setting_on_runtime}

read_runtime_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(prefix, "runningTimes.txt")))
  dt[, condition := gsub("_$", "", prefix)]
  
  dt <- janitor::clean_names(dt)
  
  dt[, job := 
       janitor::clean_names(setNames(nm = job)) %>%
       names %>% as.character
  ]
  
  dt[, `:=`(
    job = dplyr::case_when(
      job == "ms_ms_first_search" ~ "First_search",
      job == "ms_ms_main_search" ~ "Main_search",
      job == "second_peptide_search" ~ "Second_peptide_search",
      TRUE ~ job
    )
  )]
  return(dt)
}

all_runtime.dt <- lapply(
  all_sample_run_info[
    data == "data-A" & 
      !grepl("including_SECPEP", prefix) &
      grepl("m[2-8]_v[2-8]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_runtime_data,
  dir_path = file.path(data.dir, "MQ_standard/PNAS2022/runtime")
) %>% rbindlist

job_names <- all_runtime.dt[, unique(job)]

all_runtime.dt[, `:=`(
  job = factor(job, levels = rev(job_names)),
  condition = factor(condition, levels = c(
    all_sample_run_info[
      data == "data-A" & 
        !grepl("including_SECPEP", prefix) &
        grepl("m[2-8]_v[2-8]_(def|mCC)", prefix) & 
        !is.na(prefix), prefix
    ] %>% {gsub("_$", "", .)}
  ))
)]

all_runtime.dt[, `:=`(
  protease_setting = 
    str_split_fixed(condition, "_", n = 5)[, 2] %>%
    factor(levels = c("trp", "argC"))
)]

runtime_plot_colors <- setNames(
  rep("gray80", times = length(job_names)), nm = job_names
) 

runtime_plot_colors[c("First_search", "Main_search", "Second_peptide_search")] <- c("royalblue1", "royalblue2", "royalblue3")

ggplot(
  data = all_runtime.dt,
  aes(
    x = condition,
    y = running_time_min,
    fill = job
  ),
  color = NA
) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(
    breaks = c("First_search", "Main_search", "Second_peptide_search"),
    values = runtime_plot_colors
  ) +
  facet_grid(~ protease_setting, scales = "free_x", space = "free") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Run time [min]")

```

# 2.2.6 Coverage in BRD2, 3, and 4

```{r coverage}

all_stoic_pos.dt <- lapply(
  all_sample_run_info[
    data == "data-A" & 
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = file.path(p2_results_dir, "MQ_Std_MSMS")
) %>% rbindlist

list.files(data.dir)
ref_protein_bs <- Biostrings::readAAStringSet(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta"))

plot_coverage <- function(stoichiometry_data, sl.accession, sl.plot_range){
  coverage_dt <- 
    lapply(
      c("data-A_trp_m2_v2_def", "data-A_trp_m5_v5_def", "data-A_trp_m7_v7_def"),
      function(x){
        preprocess_stoichiometry_data_for_plotting(
          stoichiometry_data = stoichiometry_data[sample_name == "JQ1_HeLaWT_derivatised" & condition == x], 
          accession = sl.accession, 
          plot_range = sl.plot_range, 
          all.protein.bs = ref_protein_bs
        )$coverage_data %>%
          {.[, condition := x]}
      }
    ) %>% rbindlist
  
  g <- ggplot(
    data = coverage_dt,
    aes(
      x = aa_pos,
      y = psm_coverage,
      color = condition
    )
  ) +
    geom_line() +
    scale_x_continuous(
      breaks = sl.plot_range[1]:sl.plot_range[2],
      labels = ref_protein_bs[
        str_split_fixed(names(ref_protein_bs), "\\|", n = 3)[, 2] == sl.accession
      ] %>%
        {strsplit(as.character(.), "")[[1]]} %>%
        {.[sl.plot_range[1]:sl.plot_range[2]]}      # Use amino acids as labels
    ) +
    scale_color_bright() +
    theme(
      axis.ticks.x = element_blank(),
      aspect.ratio = 0.3
    )
  
  return(g)
}

# BRD4
plot_coverage(
  stoichiometry_data = all_stoic_pos.dt, sl.accession = "O60885", sl.plot_range = c(531, 581)
)

# BRD3
plot_coverage(
  stoichiometry_data = all_stoic_pos.dt, sl.accession = "Q15059", sl.plot_range = c(483, 533)
)

# BRD2
plot_coverage(
  stoichiometry_data = all_stoic_pos.dt, sl.accession = "P25440", sl.plot_range = c(540, 590)
) 





```


# Notes

The following sites were identified by non-unique peptides (therefore total number used here is 153 instead of 150 in the paper).

           Accession position curated_oxK_site    screen
1:  Q14331|FRG1_HUMAN       27  JMJD6_substrate FLAGJMJD6
2:  Q14331|FRG1_HUMAN       29  JMJD6_substrate FLAGJMJD6
3:  Q14331|FRG1_HUMAN       30  JMJD6_substrate FLAGJMJD6
4: Q9UQ35|SRRM2_HUMAN      241  JMJD6_substrate FLAGJMJD6
5: Q9UQ35|SRRM2_HUMAN      243  JMJD6_substrate FLAGJMJD6
6: Q9UQ35|SRRM2_HUMAN      244  JMJD6_substrate FLAGJMJD6

In the end, 91% of hydroxylated sites were covered by the data, and 80% (120 / 150) of reported hydroxylation sites were identified by this workflow.



# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
