---
title: "2-2. Optimise MQ parameters"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

This script calculates the stoichiometry of PTMs.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

temp <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)

# install.packages("/fast/AG_Sugimoto/home/users/yoichiro/projects/ptm.stoichiometry", repos = NULL, type = "source")
library("readxl")

p2.res.dir <- file.path(project.dir,
                        "results",
                        "p2-analysis-setting")


```

# Import basic data

```{r import_basic_data}

pnas2022_data <- file.path(
  project.dir,
  "data/MQ_output/PNAS2022" 
)

all_sample_run_info <- read_excel(
  file.path(pnas2022_data, "PXD031221_sample_matrix.xlsx"),
  sheet = "run_setting"
) %>% data.table

data.dir <- file.path(project.dir, "data")
protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))

setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

```


# MS/MS count by different setting


```{r MS_MS_count}

read_evidence <- function(prefix, dir_path){
  input.file <- file.path(
    dir_path, 
    paste0("including_SECPEP_", prefix, "all_processed_evidence_data.csv")
  )
  if(file.exists(input.file)){
    dt <- fread(input.file)
    dt[, condition := gsub("_$", "", prefix)]
  } else {
    dt <- data.table()
  }
  return(dt)
}

evidence.dt <- lapply(
  all_sample_run_info[data %in% c("data-A", "data-D") & !is.na(prefix), prefix],
  read_evidence,
  dir_path = p2.res.dir
  ) %>% rbindlist

data.count.dt <- evidence.dt[, list(
  total_peptide_count = .N,
  total_msms_count = sum(psm_mapped),
  total_intensity = sum(peak_intensity) 
), by = list(file_name, sample_name, condition, type)]

data.count.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2]
)]

data_col = c("total_peptide_count", "total_msms_count", "total_intensity")

m2_v2.dt <- data.count.dt[
  MQ_setting == "m2_v2" & type == "MULTI-MSMS"
  ][, c("file_name", data_col), with = FALSE]
setnames(m2_v2.dt, old = data_col, new = paste0("m2_v2_", data_col))

rel_count.dt <- data.count.dt

rel_count.dt <- merge(
  rel_count.dt,
  m2_v2.dt,
  by = c("file_name")
)

rel_count.dt[, `:=`(
  rel_total_peptide_count = total_peptide_count / m2_v2_total_peptide_count,
  rel_msms_count = total_msms_count / m2_v2_total_msms_count,
  rel_intensity = total_intensity / m2_v2_total_intensity 
)]

rel_count_stat.dt <- rel_count.dt[, list(
  meab_rel_total_peptide_count = mean(rel_total_peptide_count),
  sd_rel_total_peptide_count = sd(rel_total_peptide_count),
  mean_rel_msms_count = mean(rel_msms_count),
  sd_rel_msms_count = sd(rel_msms_count),
  mean_rel_intensity = mean(rel_intensity),
  sd_rel_intensity = sd(rel_intensity)
), by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_stat.dt[, `:=`(
  protease_setting = factor(protease_setting, levels = c("trp", "argC")),
  type = factor(type, levels = c("MULTI-MSMS", "MULTI-SECPEP"))
)]

ggplot(
  rel_count_stat.dt,
  aes(
    x = MQ_setting,
    y = mean_rel_msms_count,
    fill = type
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(
    ymin = mean_rel_msms_count - sd_rel_msms_count,
    ymax = mean_rel_msms_count + sd_rel_msms_count
  ), width = 0.5, position = position_dodge(width = 0.9)) +
  facet_grid(~ propionylation + protease_setting, scales = "free", space = "free") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Relative MS/MS count") +
  scale_fill_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE")) #+
  # scale_color_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE"))

```

The results indicate that the setting with 7 miscleavage and 7 variable modifications identify the highest MS/MS count.


# QC by the position of propionylated lysines


```{r QC_by_propionylated_lysine_position}

read_all_per_pos_data_with_SECPEP <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0("including_SECPEP_", prefix, "all_per_pos_data.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_per_pos_with_SECPEP.dt <- lapply(
  all_sample_run_info[data %in% c("data-A") & grepl("m7_v7", prefix) & !is.na(prefix), prefix],
  read_all_per_pos_data_with_SECPEP,
  dir_path = p2.res.dir
  ) %>% rbindlist

all_per_pos_with_SECPEP.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2],
  is_peptide_C_term = peptide_end == aa_pos,
  is_propionylated = ptm %in% c("[Propionylation]", "[Oxidised Propionylation]")
)]

all_per_K_pos_with_SECPEP.dt <- all_per_pos_with_SECPEP.dt[aa == "K"]

propionylated_K_count.dt <- all_per_K_pos_with_SECPEP.dt[
  , list(total_MS_MS_count = sum(psm_mapped)),
  by = list(is_propionylated, is_peptide_C_term, type)
]

propionylated_K_count.dt

ggplot(
  propionylated_K_count.dt,
  aes(
    x = is_peptide_C_term,
    y = total_MS_MS_count,
    fill = is_propionylated
  )
) +
  geom_bar(stat = "identity", position="fill") +
  scale_fill_manual(values = c("TRUE" = "#4477AA", "FALSE" = "#BBBBBB")) +
  facet_grid(~ type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


Based on the analyses here, we decided to focus on using MULTI-MSMS data.


# The effect of MQ setting by K score

```{r MQ_setting_by_k_score}

read_stoic_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(prefix, "PTM_stoichiometry.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_stoic_pos.dt <- lapply(
  all_sample_run_info[
    data != "data-D" & 
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = p2.res.dir
) %>% rbindlist

all_stoic_pos.dt[, `:=`(
  MQ_setting = str_extract(condition, "m\\d+_v\\d+")
)]

ms_ms_count.dt <- all_stoic_pos.dt[
  , list(total_ms_ms_count = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos, MQ_setting)
]

ms_ms_count.dt <- merge(
  ms_ms_count.dt,
  protein.feature.dt[, .(protein_accession, aa_pos, K_ratio, K_ratio_score)],
  by = c("protein_accession", "aa_pos")
)

d.ms_ms_count.dt <- dcast(
  ms_ms_count.dt,
  protein_accession + aa_pos + K_ratio + K_ratio_score ~ MQ_setting,
  value.var = "total_ms_ms_count", fill = 0
  
)

library("RColorBrewer")
library("patchwork")

d.ms_ms_count.dt[, table(K_ratio)]

p1 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m2_v2,
    y = m5_v5
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p2 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m5_v5,
    y = m7_v7
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p1 + p2

p3 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m5_v5 - m2_v2
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25)) +
  theme(aspect.ratio = 1)

p4 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m7_v7 - m5_v5
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25))

p3 + p4

```

# The effect of MQ setting on runtime

```{r MQ_setting_on_runtime}

read_runtime_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(prefix, "runningTimes.txt")))
  dt[, condition := gsub("_$", "", prefix)]
  
  dt <- janitor::clean_names(dt)
  
  dt[, job := 
       janitor::clean_names(setNames(nm = job)) %>%
       names %>% as.character
  ]
  
  dt[, `:=`(
    job = dplyr::case_when(
      job == "ms_ms_first_search" ~ "First_search",
      job == "ms_ms_main_search" ~ "Main_search",
      job == "second_peptide_search" ~ "Second_peptide_search",
      TRUE ~ job
    )
  )]
  return(dt)
}

all_runtime.dt <- lapply(
  all_sample_run_info[
    data == "data-A" & 
      !grepl("including_SECPEP", prefix) &
      grepl("m[2-8]_v[2-8]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_runtime_data,
  dir_path = file.path(data.dir, "MQ_output/PNAS2022/runtime")
) %>% rbindlist

job_names <- all_runtime.dt[, unique(job)]

all_runtime.dt[, `:=`(
  job = factor(job, levels = rev(job_names)),
  condition = factor(condition, levels = c(
    all_sample_run_info[
      data == "data-A" & 
        !grepl("including_SECPEP", prefix) &
        grepl("m[2-8]_v[2-8]_(def|mCC)", prefix) & 
        !is.na(prefix), prefix
    ] %>% {gsub("_$", "", .)}
  ))
)]

all_runtime.dt[, `:=`(
  protease_setting = 
    str_split_fixed(condition, "_", n = 5)[, 2] %>%
    factor(levels = c("trp", "argC"))
)]

runtime_plot_colors <- setNames(
  rep("gray80", times = length(job_names)), nm = job_names
) 

runtime_plot_colors[c("First_search", "Main_search", "Second_peptide_search")] <- c("royalblue1", "royalblue2", "royalblue3")

ggplot(
  data = all_runtime.dt,
  aes(
    x = condition,
    y = running_time_min,
    fill = job
  ),
  color = NA
) + 
  geom_bar(stat = "identity") +
  scale_fill_manual(
    breaks = c("First_search", "Main_search", "Second_peptide_search"),
    values = runtime_plot_colors
  ) +
  facet_grid(~ protease_setting, scales = "free_x", space = "free") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Run time [min]")

```


# The hydroxylation sites reported in the PNAS2022 paper

```{r hydroxylation_site_reported_in_PNAS2022}

## Hydroxylation stoichiometry comparisons
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))
setnames(pnas2022.stoic.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)]

# PNAS paper reported 153 sites
nrow(pnas2022.stoic.dt[curated_oxK_site == TRUE][!duplicated(paste(accession_position))])

# Check how many hydroxylation sites that were reported by PNAS2022 are identified by the new workflow

all_stoic_pos.dt[, `:=`(
  curated_oxK_site = 
    paste0(protein_accession, "_", aa_pos) %in%
    pnas2022.stoic.dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)],
  genotype = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)")
)]

# PNAS paper reported 153 sites
nrow(pnas2022.stoic.dt[curated_oxK_site == TRUE][!duplicated(paste(protein_accession, aa_pos))])

# The number of hydroxylated sites identified by 
all_stoic_pos.dt[
  curated_oxK_site == TRUE & genotype == "WT" & ptm == "[Oxidation (K)]"
][order(stoichiometry, MQ_setting, decreasing = TRUE)][
  !duplicated(paste(genotype, MQ_setting, protein_accession, aa_pos, ptm))
  ][,list(.N), by = list(genotype, MQ_setting, ptm)][order(MQ_setting, ptm)]

## The sites not identified by the new workflow
print("Unidentified sites out of 150")
pnas2022.stoic.dt[curated_oxK_site == TRUE][!duplicated(paste(accession_position))][!(accession_position %in% all_stoic_pos.dt[MQ_setting == "m7_v7" & curated_oxK_site == TRUE & grepl("[Oxidation (K)]", ptm, fixed = TRUE) & genotype == "WT", paste0(protein_accession, "_", aa_pos)]), .(Accession, aa_pos)]

```

The following sites were identified by non-unique peptides (therefore total number used here is 153 instead of 150 in the paper).

           Accession position curated_oxK_site    screen
1:  Q14331|FRG1_HUMAN       27  JMJD6_substrate FLAGJMJD6
2:  Q14331|FRG1_HUMAN       29  JMJD6_substrate FLAGJMJD6
3:  Q14331|FRG1_HUMAN       30  JMJD6_substrate FLAGJMJD6
4: Q9UQ35|SRRM2_HUMAN      241  JMJD6_substrate FLAGJMJD6
5: Q9UQ35|SRRM2_HUMAN      243  JMJD6_substrate FLAGJMJD6
6: Q9UQ35|SRRM2_HUMAN      244  JMJD6_substrate FLAGJMJD6

In the end, 91% of hydroxylated sites were covered by the data, and 80% (120 / 150) of reported hydroxylation sites were identified by this workflow.

# Comparison of WT and KO data

```{r comparison_WT_KO}

## Compare WT vs KO data
wt_ko_comparison.dt <- all_stoic_pos.dt[
  sample_name %in% c(
    "JMJD6peptide_HeLaWT_derivatised",
    "JMJD6peptide_HeLaJMJD6KO_derivatised",
    "JQ1_HeLaWT_derivatised",
    "JQ1_HeLaJMJD6KO_derivatised"
  ) &
  MQ_setting == "m7_v7"
]

wt_ko_comparison.dt[, `:=`(
  total_sum_psm_mapped = sum_psm_mapped
)]

no_hydroxyK.stoic.dt <- copy(wt_ko_comparison.dt[aa == "K"])
# Analyse positions for which at least one sample identified hydroxylation
no_hydroxyK.stoic.dt <- no_hydroxyK.stoic.dt[
  paste0(protein_accession, "_", aa_pos) %in% 
    wt_ko_comparison.dt[
      grepl("[Oxidation (K)]", ptm, fixed = TRUE), 
      paste0(protein_accession, "_", aa_pos)
    ]
]

# Assign 0 for the hydroxylation stoichiometry for the position which only identified unmodified K
no_hydroxyK.stoic.dt <- no_hydroxyK.stoic.dt[
  !(paste0(sample_name, "_", protein_accession, "_", aa_pos) %in% 
      wt_ko_comparison.dt[
        grepl("[Oxidation (K)]", ptm, fixed = TRUE), 
        paste0(sample_name, "_", protein_accession, "_", aa_pos)
      ]
  )
]

no_hydroxyK.stoic.dt[, `:=`(
  sum_psm_mapped = 0,
  stoichiometry = 0,
  ptm = "[Oxidation (K)]"
)]

hydroxyK.stoic.dt <- rbind(
  wt_ko_comparison.dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
  no_hydroxyK.stoic.dt
)

hydroxyK.stoic.dt <- hydroxyK.stoic.dt[, list(
  stoichiometry = sum(stoichiometry),
  sum_psm_mapped_per_position = max(sum_psm_mapped_per_position)
), by = list(sample_name, protein_accession, gene_name, aa_pos, genotype)]

hydroxyK.stoic.dt <- hydroxyK.stoic.dt[order(stoichiometry, decreasing = TRUE)][
  !duplicated(paste0(protein_accession, gene_name, aa_pos, genotype))
]

hydroxyK.stoic.dt <- hydroxyK.stoic.dt[
  sum_psm_mapped_per_position > 2
]

d.hydroxyK.stoic.dt <- dcast(
  hydroxyK.stoic.dt,
  protein_accession + gene_name + aa_pos ~ genotype,
  value.var = "stoichiometry"
)

d.hydroxyK.stoic.dt[, `:=`(
  curated_oxK_site = 
    paste0(protein_accession, "_", aa_pos) %in%
    pnas2022.stoic.dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)]
)]

ggplot(
  data = d.hydroxyK.stoic.dt[order(curated_oxK_site)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = curated_oxK_site
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

d.hydroxyK.stoic.dt[WT > 0.5 & JMJD6KO < 0.01 & curated_oxK_site == FALSE]

```

The sites for BRD2 (551, 552, 555), ARL6IP4 (120, 122, 123), and maybe NIPBL (1023, 1034, 1923) might be novel hydroxylation sites.


# Comparison of the results of PNAS paper (hydoxylation stoichometry)

```{r comparison_with_PNAS_paper}

non.duplicated.pnas2022.stoic.dt <- pnas2022.stoic.dt[
  (data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep")) &
    (curated_oxK_site == TRUE) &
    total_n_feature_K > 2
][
  order(
    data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep"),
    curated_oxK_site == TRUE,
    oxK_ratio,
    decreasing = TRUE
  )][
    !duplicated(accession_position)
]

selected_stoic_pos.dt <- all_stoic_pos.dt[
  sample_name %in% c("JQ1_HeLaWT_derivatised", "JMJD6peptide_HeLaWT_derivatised") &
    aa == "K" & grepl("[Oxidation (K)]", ptm, fixed = TRUE)
][order(stoichiometry, decreasing = TRUE)][!duplicated(paste0(protein_accession, "_", aa_pos))]

pnas2022.mq1.stoic.dt <- merge(
  non.duplicated.pnas2022.stoic.dt,
  selected_stoic_pos.dt[, .(protein_accession, aa_pos, stoichiometry)],
  by = c("protein_accession", "aa_pos")
)

pnas2022.mq1.stoic.dt[, cor.test(oxK_ratio, stoichiometry)]

ggplot(
  pnas2022.mq1.stoic.dt,
  aes(
    x = oxK_ratio,
    y = stoichiometry
  )
) +
  geom_point() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(aspect.ratio = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab("Stoichiometry in PNAS2022") +
  ylab("Stoichiometry in new workflow")

```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
