---
title: "2-3. Optimise MQ parameters"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

This script calculates the stoichiometry of PTMs.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

temp <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)

# install.packages("/fast/AG_Sugimoto/home/users/yoichiro/projects/ptm.stoichiometry", repos = NULL, type = "source")
library("readxl")

p2.res.dir <- file.path(project.dir,
                        "results",
                        "p2-analysis-setting")


```

# Import basic data

```{r import_basic_data}

pnas2022_data <- file.path(
  project.dir,
  "data/MQ_output/PNAS2022" 
)

all_sample_run_info <- read_excel(
  file.path(pnas2022_data, "PXD031221_sample_matrix.xlsx"),
  sheet = "run_setting"
) %>% data.table

data.dir <- file.path(project.dir, "data")
protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))

setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

```


# MS/MS count by different setting


```{r MS_MS_count}

read_evidence <- function(prefix, dir_path){
  input.file <- file.path(
    dir_path, 
    paste0("including_SECPEP_", prefix, "all_processed_evidence_data.csv")
  )
  if(file.exists(input.file)){
    dt <- fread(input.file)
    dt[, condition := gsub("_$", "", prefix)]
  } else {
    dt <- data.table()
  }
  return(dt)
}

evidence.dt <- lapply(
  all_sample_run_info[data %in% c("data-A", "data-D") & !is.na(prefix), prefix],
  read_evidence,
  dir_path = p2.res.dir
  ) %>% rbindlist

data.count.dt <- evidence.dt[, list(
  total_peptide_count = .N,
  total_msms_count = sum(psm_mapped),
  total_intensity = sum(peak_intensity) 
), by = list(file_name, sample_name, condition, type)]

data.count.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2]
)]

data_col = c("total_peptide_count", "total_msms_count", "total_intensity")

m2_v2.dt <- data.count.dt[
  MQ_setting == "m2_v2" & type == "MULTI-MSMS"
  ][, c("file_name", data_col), with = FALSE]
setnames(m2_v2.dt, old = data_col, new = paste0("m2_v2_", data_col))

rel_count.dt <- data.count.dt

rel_count.dt <- merge(
  rel_count.dt,
  m2_v2.dt,
  by = c("file_name")
)

rel_count.dt[, `:=`(
  rel_total_peptide_count = total_peptide_count / m2_v2_total_peptide_count,
  rel_msms_count = total_msms_count / m2_v2_total_msms_count,
  rel_intensity = total_intensity / m2_v2_total_intensity 
)]

rel_count_stat.dt <- rel_count.dt[, list(
  meab_rel_total_peptide_count = mean(rel_total_peptide_count),
  sd_rel_total_peptide_count = sd(rel_total_peptide_count),
  mean_rel_msms_count = mean(rel_msms_count),
  sd_rel_msms_count = sd(rel_msms_count),
  mean_rel_intensity = mean(rel_intensity),
  sd_rel_intensity = sd(rel_intensity)
), by = list(propionylation, MQ_setting, protease_setting, type)]

rel_count_stat.dt[, `:=`(
  protease_setting = factor(protease_setting, levels = c("trp", "argC")),
  type = factor(type, levels = c("MULTI-MSMS", "MULTI-SECPEP")),
  stacked_mean_rel_msms_count = mean_rel_msms_count
)]

rel_count_stat.dt[type == "MULTI-SECPEP", `:=`(
  stacked_mean_rel_msms_count = 
    with(rel_count_stat.dt, mean_rel_msms_count[type == "MULTI-MSMS"]) +
    with(rel_count_stat.dt, mean_rel_msms_count[type == "MULTI-SECPEP"])
)]

ggplot(
  rel_count_stat.dt,
  aes(
    x = MQ_setting,
    y = mean_rel_msms_count,
    fill = type
  )
) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_errorbar(aes(
    ymin = mean_rel_msms_count - sd_rel_msms_count,
    ymax = mean_rel_msms_count + sd_rel_msms_count
  ), width = 0.5, position = position_dodge(width = 0.9)) +
  facet_grid(~ propionylation + protease_setting, scales = "free", space = "free") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  ylab("Relative MS/MS count") +
  scale_fill_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE")) #+
  # scale_color_manual(values = c("MULTI-MSMS" = "#4477AA", "MULTI-SECPEP" = "#66CCEE"))

```

The results indicate that the setting with 7 miscleavage and 7 variable modifications identify the highest MS/MS count.


# QC by the position of propionylated lysines


```{r QC_by_propionylated_lysine_position}

read_all_per_pos_data_with_SECPEP <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0("including_SECPEP_", prefix, "all_per_pos_data.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_per_pos_with_SECPEP.dt <- lapply(
  all_sample_run_info[data %in% c("data-A") & grepl("m7_v7", prefix) & !is.na(prefix), prefix],
  read_all_per_pos_data_with_SECPEP,
  dir_path = p2.res.dir
  ) %>% rbindlist

all_per_pos_with_SECPEP.dt[, `:=`(
  propionylation = case_when(
    grepl("^data-D", condition) ~ "no_propionylation",
    TRUE ~ "propionylation"
  ),
  MQ_setting = str_extract(condition, "m\\d+_v\\d+"),
  protease_setting = str_split_fixed(condition, "_", n = 3)[, 2],
  is_peptide_C_term = peptide_end == aa_pos,
  is_propionylated = ptm %in% c("[Propionylation]", "[Oxidised Propionylation]")
)]

all_per_K_pos_with_SECPEP.dt <- all_per_pos_with_SECPEP.dt[aa == "K"]

propionylated_K_count.dt <- all_per_K_pos_with_SECPEP.dt[
  , list(total_MS_MS_count = sum(psm_mapped)),
  by = list(is_propionylated, is_peptide_C_term, type)
]

propionylated_K_count.dt

ggplot(
  propionylated_K_count.dt,
  aes(
    x = is_peptide_C_term,
    y = total_MS_MS_count,
    fill = is_propionylated
  )
) +
  geom_bar(stat = "identity", position="fill") +
  scale_fill_manual(values = c("TRUE" = "#4477AA", "FALSE" = "#BBBBBB")) +
  facet_grid(~ type) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


Based on the analyses here, we decided to focus on using MULTI-MSMS data.


# The effect of MQ setting by K score

```{r high_confident_PTM_sites}

read_stoic_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(prefix, "PTM_stoichiometry.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_stoic_pos.dt <- lapply(
  all_sample_run_info[
    data != "data-D" & 
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m[257]_v[257]_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  dir_path = p2.res.dir
) %>% rbindlist

all_stoic_pos.dt[, `:=`(
  MQ_setting = str_extract(condition, "m\\d+_v\\d+")
)]

ms_ms_count.dt <- all_stoic_pos.dt[
  , list(total_ms_ms_count = sum(sum_psm_mapped)), by = list(protein_accession, aa_pos, MQ_setting)
]

ms_ms_count.dt <- merge(
  ms_ms_count.dt,
  protein.feature.dt[, .(protein_accession, aa_pos, K_ratio, K_ratio_score)],
  by = c("protein_accession", "aa_pos")
)

d.ms_ms_count.dt <- dcast(
  ms_ms_count.dt,
  protein_accession + aa_pos + K_ratio + K_ratio_score ~ MQ_setting,
  value.var = "total_ms_ms_count", fill = 0
  
)

library("RColorBrewer")
library("patchwork")

d.ms_ms_count.dt[, table(K_ratio)]

p1 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m2_v2,
    y = m5_v5
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p2 <- ggplot(
  d.ms_ms_count.dt[order(K_ratio)],
  aes(
    x = m5_v5,
    y = m7_v7
  )
) +
  geom_abline(slope = 1, intercept = 0, color = "gray60") +
  geom_point() +
  coord_cartesian(xlim = c(0, 1000), ylim = c(0, 1000)) +
  theme(aspect.ratio = 1)

p1 + p2

p3 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m5_v5 - m2_v2
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25)) +
  theme(aspect.ratio = 1)

p4 <- ggplot(
  d.ms_ms_count.dt,
  aes(
    x = factor(K_ratio),
    y = m7_v7 - m5_v5
  )
) +
  geom_hline(yintercept = 0, color = "gray60") +
  geom_boxplot(outlier.shape = NA, fill = "steelblue") +
  coord_cartesian(ylim = c(-15, 25))

p3 + p4

```

# Comparison of the results of PNAS paper (hydoxylation stoichometry)

```{r comparison_with_PNAS_paper}

all_stoic_pos.dt[, table(sample_name)]

wt_ko_comparison.dt <- all_stoic_pos.dt[
  sample_name %in% c(
    "JMJD6peptide_HeLaWT_derivatised",
    "JMJD6peptide_HeLaJMJD6KO_derivatised",
    "JQ1_HeLaWT_derivatised",
    "JQ1_HeLaJMJD6KO_derivatised"
  )
]

wt_ko_comparison.dt[, `:=`(
  JMJD6_GT = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)")
)]


zero.stoic.dt <- wt_ko_comparison.dt[aa == "K" & grepl("[Oxidation (K)]", ptm, fixed = TRUE)]


zero.stoic.dt[, `:=`(
  sum_psm_mapped = 0,
  stoichiometry = 0,
  ptm = "[Oxidation (K)]"
)]

wt_ko_comparison.dt[aa == "K" & grepl("[Oxidation (K)]", ptm, fixed = TRUE)]


## Hydroxylation stoichiometry comparisons
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))
setnames(pnas2022.stoic.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)]

non.duplicated.pnas2022.stoic.dt <- pnas2022.stoic.dt[
  (data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep")) &
    (curated_oxK_site == TRUE) &
    total_n_feature_K > 2
][
  order(
    data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep"),
    curated_oxK_site == TRUE,
    oxK_ratio,
    decreasing = TRUE
  )][
    !duplicated(accession_position)
]

selected_stoic_pos.dt <- all_stoic_pos.dt[
  sample_name %in% c("JQ1_HeLaWT_derivatised", "JMJD6peptide_HeLaWT_derivatised") &
    aa == "K" & grepl("[Oxidation (K)]", ptm, fixed = TRUE)
][order(stoichiometry, decreasing = TRUE)][!duplicated(paste0(protein_accession, "_", aa_pos))]

pnas2022.mq1.stoic.dt <- merge(
  non.duplicated.pnas2022.stoic.dt,
  selected_stoic_pos.dt[, .(protein_accession, aa_pos, stoichiometry)],
  by = c("protein_accession", "aa_pos")
)

pnas2022.mq1.stoic.dt[, cor.test(oxK_ratio, stoichiometry)]

ggplot(
  pnas2022.mq1.stoic.dt,
  aes(
    x = oxK_ratio,
    y = stoichiometry
  )
) +
  geom_point() +
  coord_cartesian(xlim = c(0, 1), ylim = c(0, 1)) +
  theme(aspect.ratio = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  xlab("Stoichiometry in PNAS2022") +
  ylab("Stoichiometry in new workflow")

```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
