---
title: "2-4. Analysis of lysine hydroxylations using diagnostic ions"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

This script examines how the use of diagnostic ions improve the analysis of lysine hydroxylations.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

# Define project directory - contains R scripts, data and results folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

# Load all R scripts from the 'functions' folder into the current session
P2_functions <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)

### Install private packages 
# Install ptm.stiochiometry package - package installed 24.09.2025
#install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries - ptm.stiochiometry,readxl and janitor
library("readxl")
library("janitor")
library("ptm.stoichiometry")

# Create and define path to result directory 
p2_results_dir <- file.path(project.dir,
                        "results",
                        "p2_DI_comparisons")
dir.create(file.path(project.dir,
                        "results",
                        "p2_DI_comparisons"), showWarnings = FALSE, recursive = TRUE)


```

# Import basic data

```{r import_basic_data, warning=FALSE}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Load sample run info data (MQ Standard)
MQ_std_sample_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_standard"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_std_sample_run_info[, sample_id := 1:.N]

# Load sample run info data (DI + iterative)
MQ_DI_it_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_with_DI_iterative" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_it_run_info[, sample_id := 1:.N]
```

# Definition of functions

```{r definition_of_functions}

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

# This function 
contrast_hydroxylation_by_genotype <- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    genotype = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)"), # new column - Genotype, extract WT or JMJD6KO from sample_name column
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_name, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  ## Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  ## Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, genotype)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ genotype, # values in genotype become into separate columns (WT & JMJD6KO)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```

# Analyse hydroxylation sites in the data of PNAS paper

```{r hydroyxlation_sites_PNAS}

# Load PNAS stoichiometry data
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022.stoic.dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022.stoic.dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022.stoic.dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

## Add columns to pnas2022.stoic.dt metadata
# new columns: accession_position, ptm, sample_name 
pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos), #this column combines data from protein_accession and aa_pos
  ptm = "[Oxidation (K)]", 
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>% #replace sample names as mentioned (from "Hela_" to "HeLa")
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)} # Take response from previous line and change accordingly 
)]

# PNAS paper reported 153 sites
# Create a data table with protein_accession, aa_pos, curated_oxK_site columns from metadata 
# order the curated oxK sites in descending order and remove duplicated rows from 
pnas2022_curated_hydroxylysine_dt <- pnas2022.stoic.dt[, .(protein_accession, aa_pos, curated_oxK_site)] %>%
  {.[order(curated_oxK_site, decreasing = TRUE)][!duplicated(paste(protein_accession, aa_pos))]}

# Subset curated_oxK_site and count number of TURE or FALSE oxK_sites
pnas2022_curated_hydroxylysine_dt[, table(curated_oxK_site)]


# Contrast non hydroxylated and hydroxylated lysines
# Take only the HeLa samples with JQ1 or J6 peptide pull-down
d.pnas.hydroxyK_dt <- contrast_hydroxylation_by_genotype(
  pnas2022.stoic.dt[
    grepl("HeLa", sample_name) & (grepl("JQ1", sample_name) | grepl("J6pep", sample_name))
  ]
)

# Retain only samples with stoichiometry values of JMJD6 and WT greater than 0 
d.pnas.hydroxyK_dt <- d.pnas.hydroxyK_dt[JMJD6KO != 0 | WT != 0]

d.pnas.hydroxyK_dt <- merge(
  d.pnas.hydroxyK_dt,
  pnas2022_curated_hydroxylysine_dt,
  by = c("protein_accession", "aa_pos")
)

# Plot - Scatterplot of hydroxylations sites indentified in the PNAS2022 paper
ggplot(
  data = d.pnas.hydroxyK_dt[order(curated_oxK_site)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = curated_oxK_site
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

```


# Analyse hydroxylation sites without DI data

```{r hydroyxlation_sites}

MQ_std_stoic_dt <- lapply(
  MQ_std_sample_run_info[
    data != "data-D" & 
      grepl("_trp_", prefix) &
      grepl("m7_v7_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  pre_prefix = "",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS")
) %>% rbindlist

MQ_std_d.hydroxyK_dt <- contrast_hydroxylation_by_genotype(MQ_std_stoic_dt)

ggplot(
  data = MQ_std_d.hydroxyK_dt,
  aes(
    x = JMJD6KO,
    y = WT
  )
) + geom_point() +
  theme(aspect.ratio = 1) 

```


# Analyse hydroxylation sites with DI data

```{r analyse_hydroxylation_sites}

MQ_DI_it_stoic_dt <- lapply(
  MQ_DI_it_run_info[, data_dir],
  read_stoic_data,
  pre_prefix = "DI_",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_iterative")
) %>% rbindlist

MQ_DI_it_d.hydroxyK_dt <- contrast_hydroxylation_by_genotype(MQ_DI_it_stoic_dt)

DI_info <- MQ_DI_it_stoic_dt[
  , .(protein_accession, aa_pos, diagnostic_peak)
  ][
    order(protein_accession, aa_pos, -diagnostic_peak)
    ][
      !duplicated(paste(protein_accession, aa_pos))
    ]

MQ_DI_it_d.hydroxyK_dt <- merge(MQ_DI_it_d.hydroxyK_dt, DI_info, by = c("protein_accession", "aa_pos"))

MQ_DI_it_d.hydroxyK_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+"
)]

MQ_DI_it_d.hydroxyK_dt[diagnostic_peak == "+"]

MQ_DI_it_d.hydroxyK_dt[diagnostic_peak == "+"][JMJD6KO < 0.0001 & WT > 0.001][, .N, by = gene_name][order(N)]


#Plot before JMJD6 value filtered 
ggplot(
  data = MQ_DI_it_d.hydroxyK_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

#Plot - filtering JMJD6 to >0.1 and presence of diagnostic ions 
ggplot(
  data = MQ_DI_it_d.hydroxyK_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  ggrepel::geom_text_repel(aes(label = ifelse(JMJD6KO > 0.1 & diagnostic_peak == "+", gene_name, NA))) +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

#CHD3 outlier 
MQ_DI_it_stoic_dt[gene_name == "CHD3" & aa_pos == 111]

```

# Comparison of data without and with DI consideration


```{r comparison_nonDI_and_DI_data, fig.width = 5, fig.height = 5}

MQ_DI_it_d.hydroxyK_dt[, `:=`(
  hydroxylation_site_class = case_when(
    JMJD6KO < 0.00001 & WT > 0.01 ~ "class_A",
    WT < 0.00001 & JMJD6KO > 0.01 ~ "class_B",
    JMJD6KO > 0.00001 & WT > 0.00001 ~ "class_C",
    TRUE ~ "others"
  )
)]

ggplot(
  MQ_DI_it_d.hydroxyK_dt[hydroxylation_site_class != "others"],
  aes(
    x = hydroxylation_site_class,
    fill = is_diagnostic_peak
  )
) +
  geom_bar() +
  scale_fill_manual(values = c("TRUE" = "#A50026", "FALSE" = "#DDDDDD"))

MQ_DI_it_d.hydroxyK_dt[, table(diagnostic_peak, hydroxylation_site_class) %>%
                   addmargins]

protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))
setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

feature_MQ_DI_it_hydroxyK_dt <- merge(
  MQ_DI_it_d.hydroxyK_dt,
  protein.feature.dt,
  by = c("protein_accession", "aa_pos")
)

```

```{r methionine_cont}

feature_MQ_DI_it_hydroxyK_dt[, `:=`(
  met_within_2 = case_when(
    nchar(Window) == 11 & grepl("M", substr(Window, start = 4, stop = 8)) ~ "Yes",
    nchar(Window) == 11 ~ "No",
    TRUE ~ "edge"
  )
)]

feature_MQ_DI_it_hydroxyK_dt[
  met_within_2 == "Yes" & is_diagnostic_peak == TRUE
]

ggplot(
  feature_MQ_DI_it_hydroxyK_dt[
    met_within_2 != "edge" & hydroxylation_site_class != "others" & CenterResidue == "K"
  ],
  aes(
    x = hydroxylation_site_class,
    fill = met_within_2
  )
) +
  geom_bar() +
  facet_grid(~ is_diagnostic_peak) +
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB"))



```

# Comparison with previous PNAS 2022 paper

## 

```{r pnas2022_comparison}

pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))
setnames(pnas2022.stoic.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)]

# PNAS paper reported 153 sites
nrow(pnas2022.stoic.dt[curated_oxK_site == TRUE][!duplicated(paste(accession_position))])

# Check how many hydroxylation sites that were reported by PNAS2022 are identified by the new workflow

MQ_DI_it_d.hydroxyK_dt[, `:=`(
 curated_oxK_site = 
    paste0(protein_accession, "_", aa_pos) %in%
    pnas2022.stoic.dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)] 
)]

MQ_DI_it_d.hydroxyK_dt[, table(is_diagnostic_peak, curated_oxK_site) %>% addmargins]

MQ_DI_it_d.hydroxyK_dt[, table(is_diagnostic_peak, hydroxylation_site_class, curated_oxK_site) %>% addmargins]

j6_target_protein <- MQ_DI_it_d.hydroxyK_dt[
  hydroxylation_site_class == "class_A" &
    is_diagnostic_peak == TRUE
][!duplicated(protein_accession)][, .(protein_accession, gene_name)]

j6_target_protein[, `:=`(
  known_target = protein_accession %in% pnas2022.stoic.dt[curated_oxK_site == TRUE, protein_accession]
)]

j6_target_protein[, table(known_target)]
j6_target_protein[known_target == FALSE]


## Correlation
non.duplicated.pnas2022.stoic.dt <- pnas2022.stoic.dt[
  grepl("HeLa", data_source) &
  # (data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep")) &
    (curated_oxK_site == TRUE) &
    total_n_feature_K > 2
][
  order(
    #data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep"),
    curated_oxK_site == TRUE,
    oxK_ratio,
    decreasing = TRUE
  )][
    !duplicated(accession_position)
]

non.duplicated.pnas2022.stoic.dt[, stoichiometry_PNAS2022 := oxK_ratio]

d.pnas2022.hydroxyK_DI_dt <- merge(
  MQ_DI_it_d.hydroxyK_dt,
  non.duplicated.pnas2022.stoic.dt[, .(protein_accession, aa_pos, stoichiometry_PNAS2022)],
  by = c("protein_accession", "aa_pos")
)

ggplot(
  d.pnas2022.hydroxyK_DI_dt,
  aes(
    x = stoichiometry_PNAS2022,
    y = WT
  )
) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = gene_name))
  theme(aspect.ratio = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

d.pnas2022.hydroxyK_DI_dt %$%
cor.test(
  stoichiometry_PNAS2022, WT
)


# Identify unique PTMs present in new data compared to PNAS2022
MQ2025_vs_PNAS2022_dt <- merge(
  MQ_DI_it_d.hydroxyK_dt,
  non.duplicated.pnas2022.stoic.dt %>% select(protein_accession, stoichiometry_PNAS2022),
  by = c("protein_accession")
)


MQ2025_vs_PNAS2022_dt <- MQ2025_vs_PNAS2022_dt %>%
  mutate(
    point_type = case_when(
      !is.na(WT) & !is.na(stoichiometry_PNAS2022) ~ "common",
      !is.na(WT) & is_diagnostic_peak == TRUE ~ "WT_only_DI",
      !is.na(WT) & (is.na(stoichiometry_PNAS2022) | is_diagnostic_peak == FALSE) ~ "unique_WT",
      is.na(WT) & !is.na(stoichiometry_PNAS2022) ~ "unique_PNAS2022"
    )
  )

ggplot(MQ2025_vs_PNAS2022_dt, aes(x = stoichiometry_PNAS2022, y = WT, color = point_type)) +
  geom_point(size = 3) +
  geom_text(aes(label = gene_name), vjust = -0.5, hjust = 0.5) +
  scale_color_manual(
    values = c(
      "common" = "black",
      "WT_only_DI" = "red",
      "unique_WT" = "orange",
      "unique_PNAS" = "blue"
    )
  ) +
  theme_minimal() +
  labs(
    x = "stoichiometry_PNAS2022",
    y = "stoichiometry_WT_2025",
    color = "Protein type"
  )

```



# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
