---
title: "2-4. Analysis of lysine hydroxylations using diagnostic ions"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

This script examines how the use of diagnostic ions improve the analysis of lysine hydroxylations.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

# Define project directory - contains R scripts, data and results folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

# renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

# Load all R scripts from the 'functions' folder into the current session
P2_functions <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)

### Install private packages 
# Install ptm.stiochiometry package - package installed 16.12.2025
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")


# Load Libraries - ptm.stiochiometry,readxl and janitor
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.4.1 Import basic data

```{r import_basic_data, warning=FALSE}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

#------------
# MQ_Standard
#------------

# Load sample run info data (MQ Standard)
MQ_std_sample_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_standard"
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_std_sample_run_info[, sample_id := 1:.N]

#-------
# MQ_DI
#-------

# Load sample run info data (MQ DI)
MQ_DI_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_with_DI" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_run_info[, sample_id := 1:.N]


#------------
# MQ_DI_noH2O
#------------

# Load sample run info data (MQ DI with no waterloss)
MQ_DI_noH2O_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_with_DI_noH2O" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_noH2O_run_info[, sample_id := 1:.N]

```

# 2.4.2 Definition of functions

```{r definition_of_functions}

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

## This function creates a data table with stoichiomtery values of WT and JMJD6KO samples for each gene_name
contrast_hydroxylation_by_genotype <- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    genotype = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)"), # new column - Genotype, extract WT or JMJD6KO from sample_name column
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_name, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, genotype)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ genotype, # values in genotype become into separate columns (WT & JMJD6KO)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```

# 2.4.3 Analyse hydroxylation sites in the data of PNAS paper

```{r hydroyxlation_sites_PNAS}

# Load PNAS stoichiometry data
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022.stoic.dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022.stoic.dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022.stoic.dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

## Add columns to pnas2022.stoic.dt metadata
# new columns: accession_position, ptm, sample_name 
pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos), #this column combines data from protein_accession and aa_pos
  ptm = "[Oxidation (K)]", 
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>% #replace sample names as mentioned (from "Hela_" to "HeLa")
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)} # Take response from previous line and change accordingly 
)]

# Create a data table with protein_accession, aa_pos, curated_oxK_site columns from metadata 
# order the curated oxK sites in descending order and remove duplicated rows from 
pnas2022_curated_hydroxylysine_dt <- pnas2022.stoic.dt[, .(protein_accession, aa_pos, curated_oxK_site)] %>%
  {.[order(curated_oxK_site, decreasing = TRUE)][!duplicated(paste(protein_accession, aa_pos))]}

# Subset curated_oxK_site and count number of TURE or FALSE oxK_sites
pnas2022_curated_hydroxylysine_dt[, table(curated_oxK_site)]

# Hydroxylations sites identified in PNAS2022 - 153 sites
pnas2022.stoic.dt[
  curated_oxK_site == TRUE & aa == "K" & ptm == "[Oxidation (K)]"
][
  order(curated_oxK_site, decreasing = TRUE)
][
  !duplicated(paste(protein_accession, aa_pos))
][
  , .N
]

# Contrast non hydroxylated and hydroxylated lysines
# Take only the HeLa samples with JQ1 or J6 peptide pull-down
d.pnas.hydroxyK_dt <- contrast_hydroxylation_by_genotype(
  pnas2022.stoic.dt[
    grepl("HeLa", sample_name) & (grepl("JQ1", sample_name) | grepl("J6pep", sample_name))
  ]
)

# Retain only samples with stoichiometry values of JMJD6 and WT greater than 0 
d.pnas.hydroxyK_dt <- d.pnas.hydroxyK_dt[JMJD6KO != 0 | WT != 0]

d.pnas.hydroxyK_dt <- merge(
  d.pnas.hydroxyK_dt,
  pnas2022_curated_hydroxylysine_dt,
  by = c("protein_accession", "aa_pos")
)

# Plot - Scatterplot of hydroxylation sites identified in the PNAS2022 paper
ggplot(
  data = d.pnas.hydroxyK_dt[order(curated_oxK_site)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = curated_oxK_site
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

```

# 2.4.4 Analyse hydroxylation sites without DI data

```{r analyse_hydroyxlation_sites_without_DI}

## Read stoichiometry data for MQ_standard 
# Results taken from calculating stoichiometry  
MQ_std_stoic_dt <- lapply(
  MQ_std_sample_run_info[
    data != "data-D" & # do not consider data-D
      grepl("_trp_", prefix) & # fetch data prefix with 'trp'
      grepl("m7_v7_(def|mCC)", prefix) & # fetch data prefix with 'm7_v7_(def or mCC)'
      !is.na(prefix), prefix # filters rows with prefix values
  ],
  read_stoic_data,
  pre_prefix = "",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_Std_MSMS")
) %>% rbindlist

# Apply 'contrast_hydroxylation_by_genotype' function to create data table with stoichiometry data for WT and JMJD6KO per gene_name 
MQ_std_d.hydroxyK_dt <- contrast_hydroxylation_by_genotype(MQ_std_stoic_dt)

# Plot - Scatter plot of hydroxylation events in MQ_standard
ggplot(
  data = MQ_std_d.hydroxyK_dt,
  aes(
    x = JMJD6KO,
    y = WT
  )
) + geom_point() +
  theme(aspect.ratio = 1) 

```

# 2.4.5 Analyse hydroxylation sites with DI (wihtout waterloss)

```{r hydroxylation_sites_DI_no_waterloss}

## Read stoichiometry data for MQ_DI without waterloss
MQ_DI_noH20_stoic_dt <- lapply(
  MQ_DI_noH2O_run_info[, prefix],
  read_stoic_data,
  pre_prefix = "DI_", 
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI")
) %>% rbindlist

# Generate data table with stoichiometry values for WT and JMJD6KO 
MQ_DI_d.hydroxyK_dt <- contrast_hydroxylation_by_genotype(MQ_DI_noH20_stoic_dt)

# Generate data table with WT and JMJD6KO with diagnostic peaks
DI_info <- MQ_DI_noH20_stoic_dt[
  , .(protein_accession, aa_pos, diagnostic_peak) # filter these columns and create a separate data table 
][
  order(protein_accession, aa_pos, -diagnostic_peak) # order the data based on the highest diagnostic peak 
][                                              
  !duplicated(paste(protein_accession, aa_pos)) # Keep only the first row for each unique protein and amino acid position,
  # removing duplicates while preserving the row with the highest diagnostic_peak (from previous sorting)
]

# merge the modified data table 'DI_info' with the metadata 'MQ_DI_d.hydroxyK_dt'
MQ_DI_d.hydroxyK_dt <- merge(MQ_DI_d.hydroxyK_dt, DI_info, by = c("protein_accession", "aa_pos"))

#Filter data table with presence of diagnostic_peak and WT stoichiometry values
MQ_DI_d.hydroxyK_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

MQ_DI_d.hydroxyK_dt[diagnostic_peak == "+"][JMJD6KO < 0.0001 & WT > 0.001][, .N, by = gene_name][order(N)] # only includes data WT values

# Plot - WT versus JMJD6KO with presence of diagnostic ions
ggplot(
  data = MQ_DI_d.hydroxyK_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

```

```{r hydroxylation_sites_DI_JMJD6KO_no_waterloss}

# Plot - WT versus JMJD6KO with presence of diagnostic ions (label points that are JMJD6KO > 0.1 with diagnostic ion)
# This plot identifies lysine hydroxylations events in the absence of JMJD6
ggplot(
  data = MQ_DI_d.hydroxyK_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  ggrepel::geom_text_repel(aes(label = ifelse(JMJD6KO > 0.1 & diagnostic_peak == "+", gene_name, NA)), max.overlaps = 50) +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

```

# 2.4.6 Comparison of data without and with DI consideration

```{r comparison_nonDI_and_DI_data, fig.width = 5, fig.height = 5}

# Categorize hydroxylation site based on JMJD6KO and WT stoichiometry values
MQ_DI_d.hydroxyK_dt[, `:=`(   # Create new column "hydroxylation_site_class"
  hydroxylation_site_class = case_when( 
    JMJD6KO < 0.00001 & WT > 0.01 ~ "class_A", # low in JMJD6KO and high in WT, then classify as class_A
    WT < 0.00001 & JMJD6KO > 0.01 ~ "class_B", # low in WT and high in JMJD6KO, then classify as class_B
    JMJD6KO > 0.00001 & WT > 0.00001 ~ "class_C", # present in JMJD6KO and WT, then classify as class_C
    TRUE ~ "others" # values that equal to 0 and missing values, classify as "others"
  )
)]

#Plot - Bar chart of hydroxylation site based on JMJD6KO and WT stoichiometry values
ggplot(
  MQ_DI_d.hydroxyK_dt[hydroxylation_site_class != "others"], # exclude "others" category 
  aes(
    x = hydroxylation_site_class,
    fill = is_diagnostic_peak
  )
) +
  geom_bar() +
  scale_fill_manual(values = c("TRUE" = "#A50026", "FALSE" = "#DDDDDD"))


# Count the total hydroxylation site per class with diagnostic peak 
MQ_DI_d.hydroxyK_dt[, table(diagnostic_peak, hydroxylation_site_class) %>%
                      addmargins]

```


# 2.4.7 Comparison of hydroxylation site class (MQ_DI) vs methionine presence with diagnostic peak 

```{r methionine_cont_MQ_DI_no_waterloss}

# Load protein feature data 
protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))
setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos")) # change column names 

# Merge data tables by protein_accession and aa_pos (MQ_DI)
feature_MQ_DI_hydroxyK_dt <- merge(
  MQ_DI_d.hydroxyK_dt,
  protein.feature.dt,
  by = c("protein_accession", "aa_pos")
)

# Create column on the presence of M in the sequence (MQ_DI) 
feature_MQ_DI_hydroxyK_dt[, `:=`( #create a new column 'met_within_2'
  met_within_2 = case_when(
    nchar(Window) == 11 & grepl("M", substr(Window, start = 4, stop = 8)) ~ "Yes", # if M is in the middle, Yes
    nchar(Window) == 11 ~ "No", # if M is not in the sequence, then No 
    TRUE ~ "edge" # if M is in the sequence but not in the middle, then edge
  )
)]

# Filter feature data 
feature_data <- feature_MQ_DI_hydroxyK_dt[
  met_within_2 != "edge" & hydroxylation_site_class != "others" & CenterResidue == "K"
]

feature_data_plot <- rbind(
  feature_data[, .(hydroxylation_site_class, met_within_2, group = "All")],
  feature_data[is_diagnostic_peak == TRUE, .(hydroxylation_site_class, met_within_2, group = "Diagnostic peak")] 
)

# Plot - new 
ggplot(feature_data_plot, aes(x = hydroxylation_site_class, fill = met_within_2)) +
  geom_bar() +
  facet_wrap(~group) +  
  scale_fill_manual(values = c("Yes" = "#4477AA", "No" = "#BBBBBB")) +
  labs(
    x = "Hydroxylation Site Class",
    y = "Count",
    fill = "Met within 2"
  )


```

# 2.4.8 Comparison with previous PNAS 2022 paper

```{r pnas2022_comparison_MQ_DI}

# Define file path to load 'long_K_stiochiometry_data'  
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

#change the column names
setnames(pnas2022.stoic.dt, old = c("uniprot_id", "position", "residue"), new = c("protein_accession", "aa_pos", "aa")) 

# Create new column 'accession_position'
pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos) # paste protein_accession and aa_pos together '_'
)]

# Check how many hydroxylation sites that were reported by PNAS2022 are identified by the new workflow
MQ_DI_d.hydroxyK_dt[, `:=`( 
  curated_oxK_site = 
    paste0(protein_accession, "_", aa_pos) %in%
    pnas2022.stoic.dt[curated_oxK_site == TRUE, paste0(protein_accession, "_", aa_pos)] 
)]

# Convert the data table into table and sum the number of diagnostic peak and curated oxK sites
MQ_DI_d.hydroxyK_dt[, table(is_diagnostic_peak, curated_oxK_site) %>% addmargins]

# Additionally, sum the number of hydroxylation site 
MQ_DI_d.hydroxyK_dt[, table(is_diagnostic_peak, hydroxylation_site_class, curated_oxK_site) %>% addmargins]

## Data table containing WT and JMJD6KO samples isolated by J6 peptide
# Filter rows and columns to create a new data table 
j6_target_protein <- MQ_DI_d.hydroxyK_dt[
  hydroxylation_site_class == "class_A" & # keep only rows where hydroxylation_site_class is "class_A"
    is_diagnostic_peak == TRUE # keep only rows where is_diagnostic_peak is TRUE
][!duplicated(protein_accession)][, .(protein_accession, gene_name)] # remove duplicated protein_accession values (keep first occurrence), select and return columns 'protein_accession', 'gene_name'

# Create a new column 'known_target'
j6_target_protein[, `:=`( 
  # check if protein_accession in j6_target_protein exists in pnas2022.stoic.dt filtered data 
  known_target = protein_accession %in% pnas2022.stoic.dt[curated_oxK_site == TRUE, protein_accession] 
)] # filter rows where curated_oxK_site is TRUE, extract protein_accession column based on curated_oxK_site 

# Create table showing number of TRUE and FALSE values in known_target
j6_target_protein[, table(known_target)]

# Return all rows where 'known_target' is FALSE
j6_target_protein[known_target == FALSE]


## Correlation between pnas2022 stoic and new stoic data  
non.duplicated.pnas2022.stoic.dt <- pnas2022.stoic.dt[
  grepl("HeLa", data_source) & # extract sample rows using HeLa cell line 
    # (data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep")) &
    (curated_oxK_site == TRUE) & # keep only rows where curated_oxK_site is TRUE
    total_n_feature_K > 2 # filter rows where total_n_feature_K is greater than 2
][
  order(
    #data_source %in% c("HeLa_WT_JQ1", "HeLa_WT_J6pep"),
    curated_oxK_site == TRUE, # sort first by 'curated_oxK_site'
    oxK_ratio, # sort by oxK_ratio
    decreasing = TRUE # arrange oxK_ratio in descending order (from largest to smallest)
  )][
    !duplicated(accession_position) # remove duplicated accession_position, (keep first occurrence value only)
  ]

# Create a new column 'stoichiometry_PNAS2022' and assign it the values of 'oxK_ratio'
non.duplicated.pnas2022.stoic.dt[, stoichiometry_PNAS2022 := oxK_ratio]

# Merge MQ_DI stoic and pnas2022 by 'protein_accession' and 'aa_pos'
d.pnas2022.hydroxyK_DI_dt <- merge(
  MQ_DI_d.hydroxyK_dt,
  non.duplicated.pnas2022.stoic.dt[, .(protein_accession, aa_pos, stoichiometry_PNAS2022)],
  by = c("protein_accession", "aa_pos")
)

# Plot - Scatterplot of PNAS2022 stoichiometry versus MQ_DI stoichiometry data 
ggplot(
  d.pnas2022.hydroxyK_DI_dt,
  aes(
    x = stoichiometry_PNAS2022,
    y = WT
  )
) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = gene_name), size = 3, force = 1, max.overlaps = 7)+
  theme(aspect.ratio = 1) +
  labs(title = "PNAS2022 stoichiometry versus MQ_DI (noH20) stoichiometry data ") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Correlation test between pnas2022 stoic and new stoic data 
d.pnas2022.hydroxyK_DI_dt %$%
  cor.test(
    stoichiometry_PNAS2022, WT
  )

```

```{r pnas2022_comparison_MQ_standard}

# Merge MQ_stoic data and PNAS2022 by 'protein_accession' and 'aa_pos' 
pnas2022_MQStd_hydroxyK_dt <- merge(
  MQ_std_d.hydroxyK_dt,
  non.duplicated.pnas2022.stoic.dt[, .(protein_accession, aa_pos, stoichiometry_PNAS2022)],
  by = c("protein_accession", "aa_pos")
)

# Plot - Scatterplot of PNAS2022 stoichiometry versus MQ_Std stoichiometry data 
ggplot(
  pnas2022_MQStd_hydroxyK_dt,
  aes(
    x = stoichiometry_PNAS2022,
    y = WT
  )
) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = gene_name), size = 3, force = 1, max.overlaps = 7)+
  theme(aspect.ratio = 1) +
  labs(title = "PNAS2022 stoichiometry versus MQ_DI (noH20) stoichiometry data ") +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Correlation test between PNAS2022 stoichiometry versus MQ_Std stoichiometry data 
pnas2022_MQStd_hydroxyK_dt %$%
  cor.test(
    stoichiometry_PNAS2022, WT
  )

```

# 2.4.9 Number of hydroxylation sites and proteins identified in PNAS2022 and new workflow

```{r comparison_PNAS2022_vs_MQ}

#----------
## MQ_All
#----------

# MQ data including all lysine hydroxylation sites
MQ_all_hoxy_site_count <- MQ_DI_noH20_stoic_dt[aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, aa_pos))]

# Add new column to data table 
MQ_all_hoxy_site_count[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)
]

# MQ data including all lysine hydroxylation proteins
MQ_all_hoxy_protein_count <- MQ_DI_noH20_stoic_dt[aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, gene_name))]

#-----------
## MQ_DI
#-----------

# Count the total number of lysine hydroxylations in MQ_DI (aa_pos)
MQ_DI_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, aa_pos))][, .N]

# create data table with hydroxylated sites 
MQ_DI_site_count <- MQ_DI_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, aa_pos))]

# Add new column to data table 
MQ_DI_site_count[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)
]

# Count the total number of lysine hydroxylations in MQ_DI (gene_name)
MQ_DI_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, gene_name))][, .N]

# create data table with hydroxylated proteins
MQ_DI_protein_count <- MQ_DI_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, gene_name))]


#------------
# MQ_DI_noH2O
#------------

# Count the total number of lysine hydroxylations in MQ_DI (aa_pos)
MQ_DI_noH20_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, aa_pos))][, .N]

# create data table with hydroxylated sites 
MQ_DI_noH2O_site_count <- MQ_DI_noH20_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, aa_pos))]

# Add new column to data table 
MQ_DI_noH2O_site_count[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos)
)
]

# Count the total number of lysine hydroxylations in MQ_DI (gene_name)
MQ_DI_noH20_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, gene_name))][, .N]

# create data table with hydroxylated proteins
MQ_DI_noH2O_protein_count <- MQ_DI_noH20_stoic_dt[diagnostic_peak == "+"][aa == "K"][ptm == "[Oxidation (K)]"][order(protein_accession, aa_pos, -diagnostic_peak)][!duplicated(paste(protein_accession, gene_name))]


#-----------
## PNAS2022
#-----------

# Define file path to load 'long_K_stiochiometry_data'  
pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022.stoic.dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022.stoic.dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022.stoic.dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

# Create new column 'accession_position'
pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos) # paste protein_accession and aa_pos together '_'
)]

# Remove duplicated protein and aa_pos - FRG1 gene 
# Note: in the PNAS2022 paper, during protein assignment, 3 aa sequences were assigned to two different proteins by MaxQuant, which resulted 3 duplicated hydroxylated sites. In order to have a fair comparison, the 3 duplicated sites have been removed from PNAS2022 data. 

pnas2022.stoic.dt <- pnas2022.stoic.dt[
  protein_accession != "Q14331" &
    !(aa_pos %in% c(27, 29, 30))
]

# Create data table with hydroxylated sites only
# Note: this data set already contains the accession_protein column 
PNAS2022_site_count <- pnas2022.stoic.dt[
  curated_oxK_site == TRUE & aa == "K"
][
  order(curated_oxK_site, decreasing = TRUE)
][
  !duplicated(paste(protein_accession, aa_pos))
]

PNAS2022_protein_count <- pnas2022.stoic.dt[
  curated_oxK_site == TRUE & aa == "K"
][
  order(curated_oxK_site, decreasing = TRUE)
][
  !duplicated(paste(protein_accession, gene_name))
]

```

```{r Venn_diagram_hyoxy_sites}

# Plot - Venn Diagram of comparing hydroxylated sites between PNAS2022, MQ_DI and MQ_DI_it

# Install package "ggVennDiagram"
# install.packages("ggVennDiagram")
# install.packages("eulerr")

# load libraries
library(ggVennDiagram)
library(ggplot2)
library(eulerr)
library(RColorBrewer)

# Extract the accession position vectors 
vennlist <-  list(
  PNAS2022_sites = as.vector(PNAS2022_site_count[, accession_position]),
  MQ_all_hoxy_sites = as.vector(MQ_all_hoxy_site_count[, accession_position]),
  MQ_DI_noH2O_sites = as.vector(MQ_DI_noH2O_site_count[, accession_position])
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(3, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated Sites Overlap")


vennlist <-  list(
  PNAS2022_sites = as.vector(PNAS2022_site_count[, accession_position]),
  MQ_DI_noH2O_sites = as.vector(MQ_DI_noH2O_site_count[, accession_position])
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(2, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated Sites Overlap")



# Duplicated sites check PNAS 
# MQ_all_hoxy_site_count[protein_accession == "Q14331" & aa_pos %in% c(27, 29, 30)]
# MQ_all_hoxy_site_count[protein_accession == "Q9UQ35" & aa_pos %in% c(241, 243, 244)]

```

```{r Venn_diagram_hyoxy_proteins}

# Plot - Venn Diagram of comparing hydroxylated proteins between PNAS2022, MQ_DI and MQ_DI_it

# Extract the accession position vectors 
vennlist <-  list(
  PNAS2022_sites = as.vector(PNAS2022_protein_count[, gene_name]),
    MQ_all_sites = as.vector(MQ_all_hoxy_protein_count[, gene_name]),
  MQ_DI_noH2O_sites = as.vector(MQ_DI_noH2O_protein_count[, gene_name])
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(3, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated proteins overlap")



# Extract the accession position vectors 
vennlist <-  list(
  PNAS2022_sites = as.vector(PNAS2022_protein_count[, gene_name]),
  MQ_DI_noH2O_sites = as.vector(MQ_DI_noH2O_protein_count[, gene_name])
)

# Changing the circle size based on the number in the circle
venn_size_based <- euler(vennlist)

cols <- brewer.pal(2, "Set2")

# Plot the diagram
plot(venn_size_based,
     fills = list(fill = cols, alpha = 0.4),
     legend = list(side = "right"),
     quantities = TRUE,
     main = "Hydroxylated proteins overlap")
```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
