---
title: "2-4. Analysis of lysine hydroxylations using diagnostic ions"
author: "Yoichiro Sugimoto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

This script examines how the use of diagnostic ions improve the analysis of lysine hydroxylations.

# Environment setup

```{r renv_set, warning=FALSE}

# renv::init(
#           "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains/R"
#       )

project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/yoichiro/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

renv::restore(file.path(project.dir, "R"))

```


```{r set parameters}

temp <-
  sapply(list.files(
    file.path(project.dir, "R/functions"),
    pattern = "*.R",
    full.names = TRUE
  ),
  source)

library("readxl")
library("janitor")
library("ptm.stoichiometry")

p2.res.dir <- file.path(project.dir,
                        "results",
                        "p2-analysis-setting")


```

# Import basic data

```{r import_basic_data}

data.dir <- file.path(project.dir, "data")

ref_protein_dt <- import_reference_fasta(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

pnas2022_DI_data <- file.path(
  project.dir,
  "data/MQ_DI_output/PNAS2022" 
)

pnas2022_data <- file.path(
  project.dir,
  "data/MQ_output/PNAS2022" 
)

all_sample_run_info <- read_excel(
  file.path(pnas2022_data, "PXD031221_sample_matrix.xlsx"),
  sheet = "run_setting"
) %>% data.table

all_sample_run_info[, sample_id := 1:.N]

di_sample_run_info <- all_sample_run_info[grepl("data-[A-C]_trp_m7_v7_", prefix)]

```

# Definition of functions


```{r definition_of_functions}

read_stoic_data <- function(prefix, pre_prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

contrast_hydroxylation_by_genotype <- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  wt_ko_dt[, `:=`(
    genotype = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)"),
    pos_id = paste0(protein_accession, "_", aa_pos),
    sample_pos_id = paste0(sample_name, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(pos_id)
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessay
  
  ## Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  ## Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, genotype))
  ]
  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ genotype,
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```

# Analyse hydroxylation sites in the data of PNAS paper

```{r hydroyxlation_sites_PNAS}

pnas2022.stoic.dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))
setnames(
  pnas2022.stoic.dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

pnas2022.stoic.dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)],
  pnas2022.stoic.dt,
  by = "protein_accession"
)

pnas2022.stoic.dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos),
  ptm = "[Oxidation (K)]",
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>%
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)}
)]

# PNAS paper reported 153 sites
pnas2022_curated_hydroxylysine_dt <- pnas2022.stoic.dt[, .(protein_accession, aa_pos, curated_oxK_site)] %>%
  {.[order(curated_oxK_site, decreasing = TRUE)][!duplicated(paste(protein_accession, aa_pos))]}

pnas2022_curated_hydroxylysine_dt[, table(curated_oxK_site)]


# Contrast non hydroxylated and hydroxylated lysines
d.pnas.hydroxyK_dt <- contrast_hydroxylation_by_genotype(
  pnas2022.stoic.dt[
    grepl("HeLa", sample_name) & (grepl("JQ1", sample_name) | grepl("J6pep", sample_name))
  ]
)

d.pnas.hydroxyK_dt <- d.pnas.hydroxyK_dt[JMJD6KO != 0 | WT != 0]

d.pnas.hydroxyK_dt <- merge(
  d.pnas.hydroxyK_dt,
  pnas2022_curated_hydroxylysine_dt,
  by = c("protein_accession", "aa_pos")
)

ggplot(
  data = d.pnas.hydroxyK_dt[order(curated_oxK_site)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = curated_oxK_site
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))

```


# Analyse hydroxylation sites without DI data

```{r hydroyxlation_sites}

all_stoic_dt <- lapply(
  all_sample_run_info[
    data != "data-D" & 
      grepl("_trp_", prefix) &
      !grepl("including_SECPEP", prefix) &
      grepl("m7_v7_(def|mCC)", prefix) & 
      !is.na(prefix), prefix
  ],
  read_stoic_data,
  pre_prefix = "",
  dir_path = p2.res.dir
) %>% rbindlist

d.hydroxyK_dt <- contrast_hydroxylation_by_genotype(all_stoic_dt)

ggplot(
  data = d.hydroxyK_dt,
  aes(
    x = JMJD6KO,
    y = WT
  )
) + geom_point() +
  theme(aspect.ratio = 1) 

```


# Analyse hydroxylation sites with DI data

```{r analyse_hydroxylation_sites}

all_stoic_with_di_dt <- lapply(
  di_sample_run_info[grepl("data-[A|C]", prefix), prefix],
  read_stoic_data,
  pre_prefix = "DI_",
  dir_path = p2.res.dir
) %>% rbindlist

read_di_data <- function(prefix, dir_path){
  dt <- fread(file.path(dir_path, paste0("DI_", prefix, "ptm_site.csv")))
  dt[, condition := gsub("_$", "", prefix)]
  return(dt)
}

all_ptm_dt <- lapply(
  di_sample_run_info[grepl("data-[A|C]", prefix), prefix],
  read_di_data,
  dir_path = p2.res.dir
) %>% rbindlist

all_ptm_dt <- all_ptm_dt[order(
    diagnostic_peak == "+",
    score_for_localization,
    decreasing = TRUE
  )][!duplicated(paste(protein_accession, aa_pos, ptm))]


d.hydroxyK_DI_dt <- contrast_hydroxylation_by_genotype(all_stoic_with_di_dt)

d.hydroxyK_DI_dt <- merge(
  d.hydroxyK_DI_dt,
  all_ptm_dt,
  by = c("protein_accession", "aa_pos")
)

d.hydroxyK_DI_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+"
)]

d.hydroxyK_DI_dt[diagnostic_peak == "+"]

ggplot(
  data = d.hydroxyK_DI_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))


ggplot(
  data = d.hydroxyK_DI_dt[order(is_diagnostic_peak)],
  aes(
    x = JMJD6KO,
    y = WT,
    color = is_diagnostic_peak
  )
) + geom_point() +
  ggrepel::geom_text_repel(aes(label = ifelse(JMJD6KO > 0.1 & diagnostic_peak == "+", gene_name, NA))) +
  theme(aspect.ratio = 1) +
  scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black"))



d.hydroxyK_DI_dt[JMJD6KO > 0.1 & diagnostic_peak == "+"]

```

# Comparison of data without and with DI consideration


```{r comparison_nonDI_and_DI_data}

protein.feature.dt <- fread(file.path(data.dir, "PNAS2022/all_protein_feature_per_position.csv"))
setnames(protein.feature.dt, old = c("uniprot_id", "position"), new = c("protein_accession", "aa_pos"))

```

# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
