---
title: "2-5. Lysine hydroxylations in hypoxia and normoxia"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```


# 2.5.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files(file.path(project.dir, "R/functions"), pattern="*.R", full.names = TRUE), source)

## Install private package 
# Install ptm.stiochiometry package
#install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```


# 2.5.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

# 2.5.3 Calculation of Stoichiometry with diagnostic ion in hypoxia and normoxia 

```{r data_directory_MS_KR_1}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

# Load data into environment
MS_KR1_data <- file.path(
  project.dir,
  "data/MQ_with_DI/MS_KR_1" 
)

# Create file path for results
MS_KR_1_dir <- file.path(project.dir, "results", "p2-analysis-setting", "MS_KR_1")
# dir.create(MS_KR_1_dir, recursive = TRUE)

# Define file path common PTM mapping file
ptm_mapping_file <- file.path(
  project.dir,
  "data/analysis_setting/ptm_replacement_de-propionylate_for_hydroxylysine.csv"
)
```


```{r calculate_stoichiometry_MS_KR_1}
  
# Define file path to MaxQuant evidence files
mq_evidence_data <- file.path(MS_KR1_data, 
                              "MS_KR_1_evidence.txt")

# Print which sample id is being processed
message("Processing: ", basename(mq_evidence_data)) 

# Checks whether the sample ID has the corresponding evidence.txt file. If yes, then proceed
if (file.exists(mq_evidence_data)) {
  
  # Locate and fetch all PTM site files within the ptm folder
  ptm_files <- list.files(file.path(
    MS_KR1_data, "ptm"), 
    full.names = TRUE)
  
  # Generate PTM names based on file names, add brackets and remove "Sites.txt"
  ptm_names <- paste0(
    "[", str_replace_all(basename(ptm_files), "Sites.txt", ""), "]"
  )
  
  # Run the stoichiometry calculation
  stoic.dt <- calculate_stoichiometry2(
    mq_evidence_data = mq_evidence_data,
    sample_info_file = file.path(
      MS_KR1_data,
      "sample_info.csv"
    ),
    ref_protein_dt = ref_protein_dt,
    ptm_files = ptm_files,
    ptm_names = ptm_names,
    output_prefix = file.path(MS_KR_1_dir, "MS_KR_1_"),
    ptm_mapping_file = ptm_mapping_file,
    K_only = FALSE,
    selected_type = "MULTI-MSMS",
    parse_protein_accession_function = NA
  )
} else { # If evidence.txt does not exist, skip the function and print below message
  print(paste0("File does not exist: ", mq_evidence_data))
}

# Garbage collection - to free up memory
gc()


```

# 2.5.4 Plotting Stoichiometry values of hypoxia and normoxia data (+ diagnostic ion)

```{r hydroxylation_hypoxia_vs_normoxia}

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condiiton
  return(dt)
}

## Read stoichiometry data for MS_KR1 data 
MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


#Filter data table with presence of diagnostic_peak and WT stoichiometry values
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# categorize data according to sample names 
MS_KR1_stoic_dt[, `:=`(
  sample_name =  
    factor(sample_name, levels = c("HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA"))
)]

# BRD4 
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "O60885", plot_range = c(531, 581), all.protein.bs, sample_colors = NA,
)

# BRD3
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "Q15059", plot_range = c(483, 533), all.protein.bs, sample_colors = NA
)

# BRD2
plot_ptm_stoichiometry(
  MS_KR1_stoic_dt[sample_name %in% c("HeLaWT_NA_N_NA","HeLaiJMJD6_noDox_N_NA", "HeLaiJMJD6_Dox_N_NA", "HeLaiJMJD6_Dox_01O224h_NA")], accession = "P25440", plot_range = c(540, 590), all.protein.bs, sample_colors = NA
)
```

# 2.5.5 Plotting Stoichiometry of hypoxia and normoxia data with re-expression of JMJD6 (+ diagnostic peak)

```{r JMJD6_reexpression}

# Remove non JMJD6 re expression data (HeLaWT_NA_N_NA, HeLaiJMJD6_noDox_N_NA)
MS_KR1_stoic_dt <-  droplevels(MS_KR1_stoic_dt[!sample_name %in% c("HeLaWT_NA_N_NA", "HeLaiJMJD6_noDox_N_NA")])

# Filter MS_KR_1 data to retain only BRD2, 3, 4 data
MS_KR1_stoic_dt <-  MS_KR1_stoic_dt[gene_name %in% c("BRD2", "BRD3", "BRD4")]

# Filter data to have a + diagnostic peak 
MS_KR1_stoic_dt <-  MS_KR1_stoic_dt[is_diagnostic_peak == TRUE]

# Add new column combining protein accession and ass_pos to crerate unique id for each aa_pos
#MS_KR1_stoic_dt <-  MS_KR1_stoic_dt[, `:=`(
 # accession_protein = paste0(protein_accession, "_", aa_pos, "_", Oxygen_levels)
#)
#]

# converting the data format from long to wide to calculate the stoichiometry differences 
# between JMJD6 re expression and hypoxia

MS_KR1_stoic_dt <-  dcast(MS_KR1_stoic_dt, gene_name + protein_accession + aa_pos ~ sample_name, value.var = "stoichiometry")
MS_KR1_stoic_dt <- MS_KR1_stoic_dt[!is.na(HeLaiJMJD6_Dox_01O224h_NA)]

# Calculate the difference in stoichiometric values of JMJD6 re expression and Hypoxia
MS_KR1_stoic_dt <- MS_KR1_stoic_dt[, `:=`(
  Stoic_diff = (HeLaiJMJD6_Dox_N_NA - HeLaiJMJD6_Dox_01O224h_NA),
  Stoic_ratio = (HeLaiJMJD6_Dox_01O224h_NA/HeLaiJMJD6_Dox_N_NA)
)]


# Plot - Stoic difference between BRD2, 3 and 4
ggplot(
  data = MS_KR1_stoic_dt,
  aes(
    x = gene_name,
    y = Stoic_diff
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point() +
  labs(title = "Stoichiometry difference between BRD2, 3 and 4") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(legend.position="none")



ggplot(
  data = MS_KR1_stoic_dt,
  aes(
    x = gene_name,
    y = Stoic_ratio
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point() +
  labs(title = "Stoichiometry difference between BRD2, 3 and 4") +
  theme(axis.text.x = element_text(angle = 0, hjust = 1)) + 
  theme(legend.position="none")





# Plot - JMJD6 re-expression - BRD4
ggplot(
  data = MS_KR1_stoic_dt[gene_name %in% c("BRD4") & diagnostic_peak == "+"],
  aes(
    x = Oxygen_levels,
    y = stoichiometry,
    group = paste(protein_accession, aa_pos)
  )
) +
  geom_point(aes(colour = paste(protein_accession, aa_pos))) +
  geom_line(linewidth = 0.4) +
  labs(title = "BRD4") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position="none") +
  scale_x_discrete(limits = rev(levels(MS_KR1_stoic_dt$Oxygen_levels)))


## Boxplot

install.packages("viridis")
library(viridis)

# Plot - JMJD6 re-expression - BRD2
ggplot(
  data = MS_KR1_stoic_dt[gene_name %in% c("BRD2") & diagnostic_peak == "+"],
  aes(
    x = Oxygen_levels,
    y = stoichiometry
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point(aes(colour = paste(protein_accession, aa_pos))) +
  geom_line(aes(group =  aa_pos), size=0.3, colour='black') +
  labs(title = "BRD2") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position="none") +
  scale_x_discrete(limits = rev(levels(MS_KR1_stoic_dt$Oxygen_levels)))


# Plot - JMJD6 re-expression - BRD3
ggplot(
  data = MS_KR1_stoic_dt[gene_name %in% c("BRD3") & diagnostic_peak == "+"],
  aes(
    x = Oxygen_levels,
    y = stoichiometry
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point(aes(colour = paste(protein_accession, aa_pos))) +
  geom_line(aes(group =  aa_pos), size=0.3, colour='black') +
  labs(title = "BRD3") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position="none") +
  scale_x_discrete(limits = rev(levels(MS_KR1_stoic_dt$Oxygen_levels)))


# Plot - JMJD6 re-expression - BRD4
ggplot(
  data = MS_KR1_stoic_dt[gene_name %in% c("BRD4") & diagnostic_peak == "+"],
  aes(
    x = Oxygen_levels,
    y = stoichiometry
  )
) +
  stat_boxplot(geom = "errorbar", width = 0.25) + 
  geom_boxplot() +
  geom_point(aes(colour = paste(protein_accession, aa_pos))) +
  geom_line(aes(group =  aa_pos), size=0.3, colour='black') +
  labs(title = "BRD4") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  theme(legend.position="none") +
  scale_x_discrete(limits = rev(levels(MS_KR1_stoic_dt$Oxygen_levels)))

```


# Session information

```{r sessionInfo}

sessioninfo::session_info()

```
