---
title: "2-9. Site-specific stoichiometry of lysine hydroxylations in varying O2% and dox induction time"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines site-specific stoichiometry of lysine hydroxylations in BRD proteins across varying oxygen percentages and doxycycline induction times. The site-specific stoichiometries are visualised using the plot_ptm_stoichiometry function from the ptm.stoichiometry package

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```

# 2.9.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.9.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p2_MS_SS_KR <- file.path(results.dir, "p2_MS_SS_KR")
# create.dirs(c(results.dir, p2_MS_SS_KR))

```

# 2.9.3 Defining functions and data preprocessing

```{r Define_important_functions}

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

```


```{r load_PNAS_MS_KR_1_MS_SS_data}

#-----------------------------
# Load MS_KR_1 stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_KR1 data (noH2O loss)
MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_SS data 
MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS_noH2O"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant (dataset A)
#--------------------------------------------------------------

pnas2022_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
pnas2022_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#------------------------------------------------------
# Load PNAS stoichiometry data manually curated sites
#------------------------------------------------------

# Load PNAS stoichiometry data
pnas2022_dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022_dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

```


```{r data_preprocessing}

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_stoic_dt <- MS_SS_stoic_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_PNAS_dt <- rbindlist(list(MS_SS_stoic_dt,
                         MS_KR1_stoic_dt,
                         pnas2022_stoic_dt),
                         use.names = TRUE)

MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Column "sample group" created to simplify sample names
MS_SS_KR_PNAS_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

# 2.9.4 Plotting raw stoichiometry values of normoxia data with re-expression of JMJD6 

```{r stoichiometry_BRD2_at_21pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented 
# in the PNAS2022 paper. 

# subset data to specific BRD2 protein
MS_SS_KR_PNAS_BRD2_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
                                            gene_name == "BRD2"]


# reorder sample names
MS_SS_KR_PNAS_BRD2_dt <-  MS_SS_KR_PNAS_BRD2_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD3_at_21pc}

# subset data to specific BRD3 protein
MS_SS_KR_PNAS_BRD3_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD3"]



MS_SS_KR_PNAS_BRD3_dt <-  MS_SS_KR_PNAS_BRD3_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD4_at_21pc}

# subset data to specific BRD4 protein
MS_SS_KR_PNAS_BRD4_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD4", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD4"]


# reorder sample names
MS_SS_KR_PNAS_BRD4_dt <-  MS_SS_KR_PNAS_BRD4_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD4",
             "8h_21pc_BRD4",
             "18h_21pc_BRD4",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```

# 2.9.5 Plotting raw stoichiometry values of normoxia and hypoxia data with re-expression of JMJD6 

```{r stoichiometry_BRD2_at_hypoxia_pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented in the PNAS2022 paper. 

# HeLaWT_NA_N_NA - WT_Inf_21pc_KR
# JQ1_HeLaWT_derivatised - WT_Inf_21pc_PNAS
# JQ1_HeLaJMJD6KO_derivatised - iJ6_Inf_21pc_PNAS
# HeLaiJMJD6_Dox_N_NA - iJ6_24h_21pc_KR
# 18h_4pc_BRD23 - iJ6_18h_4pc_SS
# 18h_1pc_BRD23 - iJ6_18h_1pc_SS
# HeLaiJMJD6_Dox_01O224h_NA - iJ6_24h_01pc_KR
# minusDox_BRD23 - iJ6_0h_21pc_SS
# HeLaiJMJD6_noDox_N_NA - iJ6_0h_21pc_KR


# HeLaWT_NA_N_NA - WT_Inf_21pc_KR
# JQ1_HeLaWT_derivatised - WT_Inf_21pc_PNAS
# JQ1_HeLaJMJD6KO_derivatised - iJ6_Inf_21pc_PNAS
# HeLaiJMJD6_Dox_N_NA - iJ6_24h_21pc_KR
# 18h_4pc_BRD23 - iJ6_18h_4pc_SS
# 18h_1pc_BRD23 - iJ6_18h_1pc_SS
# HeLaiJMJD6_Dox_01O224h_NA - iJ6_24h_01pc_KR
# HeLaiJMJD6_noDox_N_NA - iJ6_0h_21pc_KR

#------
# BRD2
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD2_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD2"]

# reorder sample names
MS_SS_KR_PNAS_BRD2_O2pc_dt <-  MS_SS_KR_PNAS_BRD2_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_O2pc_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD3_at_hypoxia_pc}

#------
# BRD3
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD3_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD3"]


# reorder sample names
MS_SS_KR_PNAS_BRD3_O2pc_dt <-  MS_SS_KR_PNAS_BRD3_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]


# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_O2pc_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r stoichiometry_BRD4_at_hypoxia_pc}

#------
# BRD4
#------

# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD4_O2pc_dt <- MS_SS_KR_PNAS_dt[  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD4", "18h_21pc_BRD4","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD4"]


#iJ6_0h_21pc_SS - minusDox_BRD23
# HeLaiJMJD6_noDox_N_NA

# reorder sample names
MS_SS_KR_PNAS_BRD4_O2pc_dt <-  MS_SS_KR_PNAS_BRD4_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD4",
             "18h_1pc_BRD4",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_O2pc_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```

# Session information

```{r session_Info}

sessioninfo::session_info()

```
