---
title: "2-8. Lysine hydroxylations in varying O2 pc and dox induction time - data visualisation"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines the oxygen sensitivity of lysine hydroxylations in BRD proteins and their respective sites. This script is for visualizing the calculated stoichiometry. 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```

# 2.7.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.7.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

```

# 2.7.3 Plotting Stoichiometry values of hypoxia and normoxia data with re-expression of JMJD6 (+ diagnostic ion)

```{r Define_important_functions}

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

#-------------------------------------
# Contrast function - by sample group
#-------------------------------------

contrast_hydroxylation_by_sample_group<- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_group, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, sample_group)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ sample_group, # values in oxygen_stat become into separate columns (Normoxia & Hypoxia)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```


```{r load_MS_KR_1_MS_SS_data}

#-----------------------------
# Load MS_KR_1 stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_KR1 data (noH2O loss)
 MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_SS data 
 MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS_noH2O"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant (dataset A)
#--------------------------------------------------------------

 pnas2022_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
pnas2022_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]
  
```

```{r data_preprocessing}

# Filter MS_KR_1 data to retain only BRD2, 3, 4 data
MS_KR1_BRD_dt <- MS_KR1_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Filter MS_SS data to retain only BRD2, 3, 4 data 
MS_SS_BRD_dt <-  MS_SS_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Filter PNAS2022 data to retain only BRD2, 3, 4 data 
pnas2022_BRD_dt <-  pnas2022_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_PNAS_dt <- rbindlist(list(MS_SS_BRD_dt,
                         MS_KR1_BRD_dt,
                         pnas2022_BRD_dt),
                         use.names = TRUE)

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Column "sample group" created to simplify sample names
MS_SS_KR_PNAS_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

```{r stoichiometry_BRD_proteins_at_21pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented 
# in the PNAS2022 paper. 

# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD2_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                     grepl("21pc", sample_group) & 
                     !sample_name %in% c("HeLaiJMJD6_noDox_N_NA") & 
                     gene_name == "BRD2"]


# reorder sample names
MS_SS_KR_PNAS_BRD2_dt <-  MS_SS_KR_PNAS_BRD2_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "JQ1_HeLaWT_derivatised",
             "JQ1_HeLaJMJD6KO_derivatised",
             "minusDox_BRD23",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)


# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD3_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                     grepl("21pc", sample_group) & 
                     !sample_name %in% c("HeLaiJMJD6_noDox_N_NA") & 
                     gene_name == "BRD3"]


# reorder sample names
MS_SS_KR_PNAS_BRD3_dt <-  MS_SS_KR_PNAS_BRD3_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "JQ1_HeLaWT_derivatised",
             "JQ1_HeLaJMJD6KO_derivatised",
             "minusDox_BRD23",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD4_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                     grepl("21pc", sample_group) & 
                     !sample_name %in% c("HeLaiJMJD6_noDox_N_NA") & 
                     gene_name == "BRD4"]


# reorder sample names
MS_SS_KR_PNAS_BRD4_dt <-  MS_SS_KR_PNAS_BRD4_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "JQ1_HeLaWT_derivatised",
             "JQ1_HeLaJMJD6KO_derivatised",
             "minusDox_BRD4",
             "4h_21pc_BRD4",
             "8h_21pc_BRD4",
             "18h_21pc_BRD4",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r merge_MS_SS_KR_data}

# subset rows with diagnostic ion from MS_SS and MS_KR data
DI_sites <- rbind(
  MS_KR1_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  MS_SS_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)]
  ############ Add DI sites of dataset A
)

# remove duplicate sites 
DI_sites <- DI_sites[!duplicated(paste(protein_accession, aa_pos))]

# Merge MS_SS_KR data to only have unique diagnostic ion sites
MS_SS_KR_PNAS_dt <- merge(
  DI_sites,
  MS_SS_KR_PNAS_dt[aa == "K"],
  by = c("protein_accession", "aa_pos")
)

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Using contrast function on MS_SS_KR raw stoic data
c_MS_SS_KR_PNAS_dt <- contrast_hydroxylation_by_sample_group(MS_SS_KR_PNAS_dt)

# Factor the gene names
c_MS_SS_KR_PNAS_dt[, `:=`(
  gene_name =  
    factor(gene_name)
)]

# WT values subset to greater than 0 
c_MS_SS_KR_PNAS_dt <- c_MS_SS_KR_PNAS_dt[WT_Inf_21pc_KR > 0]

```


```{r overall_MS_SS_KR_comparison}

# Change wide format to long format 
lc_MS_SS_KR_PNAS_dt <- melt(
  c_MS_SS_KR_PNAS_dt,
  id.vars = c("protein_accession", "gene_name", "aa_pos"),
  value.name = "Stoichiometry", 
  variable.name = "sample_group"
  #value.var = grep("_", colnames(c_MS_SS_KR_all_dt), value = TRUE)
)

# Separate the contents of sample_group column into new columns 
lc_MS_SS_KR_PNAS_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2] %>%
    factor(levels = c("Inf", "0h", "4h", "8h", "18h", "24h")),
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# subset rows that do not contain NA values 
lc_MS_SS_KR_PNAS_dt <- lc_MS_SS_KR_PNAS_dt[!is.na(Stoichiometry)]

# counting the data per sample and size 
lc_MS_SS_KR_PNAS_dt[, data_size_per_sample := .N, by = list(sample_group)]
lc_MS_SS_KR_PNAS_dt[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]

# reorder the oxygen levels
lc_MS_SS_KR_PNAS_dt <-  lc_MS_SS_KR_PNAS_dt[, oxygen := factor(
  oxygen,
  levels = c("01pc", "1pc", "4pc", "21pc")
)]


# subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- lc_MS_SS_KR_PNAS_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stoichiometry")

wt_21pc <- wt_21pc[WT_stoichiometry > 0]

# Merge WT data with lc_MS_SS_KR_PNAS_dt dtata
lc_MS_SS_KR_PNAS_WT_dt <- merge(
  lc_MS_SS_KR_PNAS_dt[
    cell %in% c("iJ6", "WT")
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)


# Boxplot - Stoichiometry of Dox + and - JMJD6 under O2 levels
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```

```{r Dox_plus_minus_JMJD6_O2_levels}

# Boxplot - Stoichiometry of Dox + and - JMJD6 under O2 levels
ggplot(
  data = lc_MS_SS_KR_PNAS_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#---------------------

# O2 levels subset to 21pc 
lc_MS_SS_KR_PNAS_WT_dt <- lc_MS_SS_KR_PNAS_WT_dt[oxygen == "21pc"]


# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2
ggplot(
  data = lc_MS_SS_KR_PNAS_21pc_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_21pc_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
 # facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


#------------------------

# O2 levels subset to 21pc and (N >= 0.1)
lc_MS_SS_KR_PNAS_21pc_N01_dt <- lc_MS_SS_KR_PNAS_WT_dt[WT_stoichiometry >= 0.1]


# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2
ggplot(
  data = lc_MS_SS_KR_PNAS_21pc_N01_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_21pc_N01_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("(N >= 0.1) Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r normoxia_varying_dox_induction}

#---------------------------------------------
# Normoxia samples under varying dox induction
#---------------------------------------------

# subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- long_c_MS_SS_KR_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stioichiomtry")

# Merge WT data with MS_SS_KR normoxia data
pc2wt_21pc <- merge(
  long_c_MS_SS_KR_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)

wt_21pc <- wt_21pc[WT_stioichiomtry > 0]

# Boxplot - comparing normoxia in different dox incubation timings
ggplot(
  data = pc2wt_21pc[data_size_per_site == max(pc2wt_21pc[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry / WT_stioichiomtry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


```

# 2.7.5 XIC values against MS_SS and MS_KR_1 stoichiometry data. 

```{r XIC_vs_stoic}

#-------------
# XIC vs stoic
#-------------

# Load XIC data
xic_MS_SS <- fread(file.path(data.dir, "xic_MS_SS.csv"))

# Merge XIC data and stoichiometry data
xic_stoic <- merge(long_c_MS_SS_KR_dt, 
                   xic_MS_SS,
                   by = c("induction", "oxygen", "gene_name", "aa_pos"))

# Scatter plot - XIC vs stoichiometry
ggplot(
  xic_stoic,
  aes(
    x = XIC,
    y = Stoichiometry
  )
) + 
  geom_point() +
  theme_classic()

```

# Oxygen sensitivity


```{r normoxia_vs_varying_hypoxia_pc_overnightdox}

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
# Data subset to overnight (18/24h) dox induction

long_MS_SS_KR_O2p_dt <- long_c_MS_SS_KR_dt[grepl("18h|24h|WT", sample_group)]

# reorder the sample_group according to oxygen levels
long_MS_SS_KR_O2p_dt[, sample_group :=
                       factor(sample_group,
                              levels = c("WT_Inf_21pc_KR",
                                         "iJ6_24h_21pc_KR",
                                         # "iJ6_18h_21pc_SS",
                                         "iJ6_18h_4pc_SS",
                                         "iJ6_18h_1pc_SS",
                                         "iJ6_24h_01pc_KR"))
]

#### Add WT stoichiometry columns as did above

long_MS_SS_KR_O2p_dt ############ Subset the sites with data in all the samples

# Plot
ggplot(
  data = long_MS_SS_KR_O2p_dt, ############ Subset the sites with data in all the samples
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))



#-----------------------------



# binning normoxia values 

# subset data to WT normoxia (MS_SS_KR sample)
MS_SS_KR_21pc <- long_c_MS_SS_KR_dt[grepl("18h|24h", sample_group) & grepl("21pc", oxygen),  .(protein_accession, aa_pos, Stoichiometry)]

MS_SS_KR_21pc[
  ,
  normoxia_bin := cut(
    Stoichiometry,
    breaks = c(0, 0.3, 0.6, 1),
    labels = c("0–0.3", "0.3–0.6", "0.6–1"),
    include.lowest = TRUE
  )
]

```

```{r normoxia_vs_hypoxia_scatterplot}

# Scatter plot - Normoxia stoichiometry versus hypoxia at 4% - 18h dox incubation
ggplot(
  data = c_MS_SS_KR_dt, 
  aes(
    x = iJ6_18h_21pc_SS,
    y = iJ6_18h_4pc_SS
  )
) +
  geom_point(aes(colour = gene_name)) +
  #facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(c_MS_SS_KR_dt[, iJ6_18h_21pc_SS], c_MS_SS_KR_dt[, iJ6_18h_4pc_SS], method = "spearma")

wilcox.test(
  c_MS_SS_KR_dt[, iJ6_18h_4pc_SS],
  c_MS_SS_KR_dt[, iJ6_18h_21pc_SS],
  paired = TRUE
)

# Scatter plot - Normoxia stoichiometry versus hypoxia at 1% - 18h dox incubation
ggplot(
  data = c_MS_SS_KR_dt, 
  aes(
    x = iJ6_18h_21pc_SS,
    y = iJ6_18h_1pc_SS
  )
) +
  geom_point(aes(colour = gene_name)) +
#  facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(c_MS_SS_KR_dt[, iJ6_18h_21pc_SS], c_MS_SS_KR_dt[, iJ6_18h_1pc_SS], method = "spearma")

wilcox.test(
  c_MS_SS_KR_dt[, iJ6_18h_1pc_SS],
  c_MS_SS_KR_dt[, iJ6_18h_21pc_SS],
  paired = TRUE
)

# Scatter plot - Normoxia stoichiometry versus hypoxia at 0.1% - 24h dox incubation
ggplot(
  data = c_MS_SS_KR_dt, 
  aes(
    x = iJ6_24h_21pc_KR,
    y = iJ6_24h_01pc_KR
  )
) +
  geom_point(aes(colour = gene_name)) +
  #facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right")

# correlation test 
cor.test(c_MS_SS_KR_dt[, iJ6_24h_21pc_KR], c_MS_SS_KR_dt[, iJ6_24h_01pc_KR], method = "spearma")

wilcox.test(
  c_MS_SS_KR_dt[, iJ6_24h_01pc_KR],
  c_MS_SS_KR_dt[, iJ6_24h_21pc_KR],
  paired = TRUE
)

```

# 2.7.4 Comparing hypoxia stoichiometry under varying dox incubation (+ diagnostic ion)

```{r hypoxia_vs_dox_incubation}

# box_plot - hypoxia stoichiometry in diff dox incubation
ggplot(
  data = long_c_MS_SS_KR_dt[induction != "Inf" & oxygen != "21pc"], 
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen)+
  ggtitle("Hypoxia samples in diff dox incubation") +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#--------------------

# Subset normoxia data to >= 0.1
O2_21pc_N01_dt <- long_c_MS_SS_KR_dt[
  oxygen == "21pc" & Stoichiometry >= 0.1
]
  
# Combine subset normoxia 
MS_hypo_Nsub_dt <- rbindlist(list(O2_21pc_N01_dt,
                                     long_c_MS_SS_KR_dt),
                                use.names = TRUE)

# box_plot - Hypoxia samples in diff dox incubation (N =>0.1)
ggplot(
  data = MS_hypo_Nsub_dt[induction != "Inf" & oxygen != "21pc"], 
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
 facet_grid(~ oxygen)+
  ggtitle("Hypoxia samples in diff dox incubation (N subset)") +
  ylab("Stoichiometry [%] (N =>0.1)") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```





# Session information

```{r session_Info}

sessioninfo::session_info()

```
