---
title: "2-10. Lysine hydroxylations in varying O2 pc and dox induction time - data visualisation"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines the oxygen sensitivity of lysine hydroxylations in BRD proteins and their respective sites. This script is for visualizing the calculated stoichiometry. 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```

# 2.10.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.10.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p2_MS_SS_KR <- file.path(results.dir, "p2_MS_SS_KR")
# create.dirs(c(results.dir, p2_MS_SS_KR))

```

# 2.10.3 Defining functions and data preprocessing

```{r Define_important_functions}

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

#-------------------------------------
# Contrast function - by sample group
#-------------------------------------

contrast_hydroxylation_by_sample_group<- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_group, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, sample_group)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ sample_group, # values in oxygen_stat become into separate columns (Normoxia & Hypoxia)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```


```{r load_PNAS_MS_KR_1_MS_SS_data}

#-----------------------------
# Load MS_KR_1 stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_KR1 data (noH2O loss)
MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_SS data 
MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS_noH2O"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant (dataset A)
#--------------------------------------------------------------

pnas2022_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
pnas2022_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#------------------------------------------------------
# Load PNAS stoichiometry data manually curated sites
#------------------------------------------------------

# Load PNAS stoichiometry data
pnas2022_dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022_dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

```


```{r data_preprocessing}

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_stoic_dt <- MS_SS_stoic_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_PNAS_dt <- rbindlist(list(MS_SS_stoic_dt,
                         MS_KR1_stoic_dt#,
                         #pnas2022_stoic_dt
                         ),
                         use.names = TRUE)

MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Column "sample group" created to simplify sample names
MS_SS_KR_PNAS_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

# 2.10.3 Plotting Stoichiometry values of hypoxia and normoxia data with re-expression of JMJD6 (+ diagnostic ion)

```{r merge_MS_SS_KR_PNAS_data}

# subset rows with diagnostic ion from MS_SS and MS_KR data
DI_sites <- rbind(
  MS_SS_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  MS_KR1_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_stoic_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_dt[curated_oxK_site == TRUE, .(protein_accession, aa_pos)]
)

# remove duplicate sites 
DI_sites <- DI_sites[!duplicated(paste(protein_accession, aa_pos))]

# Merge MS_SS_KR data to only have unique diagnostic ion sites
MS_SS_KR_PNAS_dt <- merge(
  DI_sites,
  MS_SS_KR_PNAS_dt[aa == "K"],
  by = c("protein_accession", "aa_pos")
)

# Using contrast function on MS_SS_KR raw stoic data
c_MS_SS_KR_PNAS_dt <- contrast_hydroxylation_by_sample_group(MS_SS_KR_PNAS_dt)

# Factor the gene names
c_MS_SS_KR_PNAS_dt[, `:=`(
  gene_name =  
    factor(gene_name, levels = c("BRD2", "BRD3", "BRD4"))
)]


```


```{r overall_MS_SS_KR_PNAS_comparison}

# Change wide format to long format 
lc_MS_SS_KR_PNAS_dt <- melt(
  c_MS_SS_KR_PNAS_dt,
  id.vars = c("protein_accession", "gene_name", "aa_pos"),
  value.name = "Stoichiometry", 
  variable.name = "sample_group"
  #value.var = grep("_", colnames(c_MS_SS_KR_all_dt), value = TRUE)
)

# Separate the contents of sample_group column into new columns 
lc_MS_SS_KR_PNAS_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2] %>%
    factor(levels = c("Inf", "0h", "4h", "8h", "18h", "24h")),
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# reorder the oxygen levels
lc_MS_SS_KR_PNAS_dt <-  lc_MS_SS_KR_PNAS_dt[, oxygen := factor(
  oxygen,
  levels = c("01pc", "1pc", "4pc", "21pc")
)]

# subset rows that do not contain NA values 
lc_MS_SS_KR_PNAS_dt <- lc_MS_SS_KR_PNAS_dt[!is.na(Stoichiometry)]

# counting the data per sample and size 
lc_MS_SS_KR_PNAS_dt[, data_size_per_sample := .N, by = list(sample_group)]
lc_MS_SS_KR_PNAS_dt[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]

```

# 2.10.4 Normoxia samples under varying dox induction

```{r Dox_plus_minus_JMJD6_21_O2pc}

# Subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- lc_MS_SS_KR_PNAS_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stoichiometry")

# Sub-setting and combining data tables 
long_pc2wt_21pc <- rbind(
  lc_MS_SS_KR_PNAS_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  lc_MS_SS_KR_PNAS_dt[sample_group == "WT_Inf_21pc_KR"]
)
long_pc2wt_21pc <- long_pc2wt_21pc[sample_group != "iJ6_0h_21pc_SS"]

# Here, we combine induction times 18h and 24h into category 18-24h 
lc_MS_SS_KR_PNAS_dt[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  induction == "24h" ~ "18-24h",
  #induction == "Inf" ~ "18-24h",
  TRUE ~ induction
)][, induction2 := factor(induction2, levels = c("Inf", "0h", "4h", "8h", "18-24h"))]

# Count data per protein accession and aa position
plot_long_pc2wt_21pc <- copy(long_pc2wt_21pc)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
plot_long_pc2wt_21pc <- plot_long_pc2wt_21pc[data_size_per_site == max(plot_long_pc2wt_21pc[, data_size_per_site])] # keep only max value of data

# Keep only sites which are WT and stoichiometry > 0 
plot_long_pc2wt_21pc <- plot_long_pc2wt_21pc[
  paste(protein_accession, aa_pos) %in% plot_long_pc2wt_21pc[cell == "WT"][Stoichiometry > 0, paste(protein_accession, aa_pos)]
]

# Boxplot - comparing normoxia in different dox incubation timings
ggplot(
  plot_long_pc2wt_21pc,
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 1.5) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggtitle("All the sites")

plot_long_pc2wt_21pc[, .N, by = induction]


# Plot ratio
# Merge MS_SS normoxia data with MS_KR_1 WT normoxia data
pc2wt_21pc <- merge(
  lc_MS_SS_KR_PNAS_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)

# Subset data to WT_stoichiometry > 0 and exclude sample group = iJ6_0h_21pc_SS
pc2wt_21pc_sub <- pc2wt_21pc[WT_stoichiometry > 0][sample_group != "iJ6_0h_21pc_SS"]

pc2wt_21pc_sub <- copy(pc2wt_21pc_sub)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
pc2wt_21pc_sub <- pc2wt_21pc_sub[data_size_per_site == max(pc2wt_21pc_sub[, data_size_per_site])]

# Boxplot - comparing normoxia in different dox incubation timings (relative to WT stoichiometry)
ggplot(
  data = pc2wt_21pc_sub,
  aes(
    x = induction,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Create data table  
induction_t_test <- data.table()

# Paired t-test 
for(i in pc2wt_21pc_sub[, unique(induction)]){
  t.out <- t.test(
    pc2wt_21pc_sub[induction == i][, Stoichiometry],
    pc2wt_21pc_sub[induction == i][, WT_stoichiometry],
    paired = TRUE
  )
  induction_t_test <- rbind(
    induction_t_test,
    data.table(induction = i, p_value = t.out$p.value, n = nrow(pc2wt_21pc_sub[induction == i]))
  )  
}

# Adjusted P-value
induction_t_test[, padj := p.adjust(p_value, method = "holm")]
induction_t_test[, significance := case_when(
  padj < 0.005 ~ "**",
  padj < 0.05 ~ "*",
  TRUE ~ "N.S."
)]

induction_t_test

```

# 2.10.5 t50 - Time to half value 

```{r time_to_half_value}

# helper: convert time to numeric and replace Inf by a large finite value
time_to_num <- function(x, inf_value = 100) {
  x_chr <- as.character(x)
  out <- suppressWarnings(as.numeric(gsub("[^0-9.]+", "", x_chr)))
  out[grepl("^inf$", x_chr, ignore.case = TRUE)] <- inf_value
  out[is.infinite(out)] <- inf_value
  out
}

calc_t50 <- function(dt, time_col = "induction", y_col = "Stoichiometry", inf_value = 100) {
  t <- time_to_num(dt[[time_col]], inf_value = inf_value)
  y <- as.numeric(dt[[y_col]])

  keep <- is.finite(t) & is.finite(y)
  t <- t[keep]
  y <- y[keep]

  if (length(t) < 2L || uniqueN(t) < 2L || all(y <= 0, na.rm = TRUE)) {
    return(NA_real_)
  }

  # average replicates at the same time
  d <- data.table(t = t, y = y)[, .(y = mean(y, na.rm = TRUE)), by = t][order(t)]

  # monotone increasing fit
  iso <- isoreg(d$t, d$y)
  y_fit <- iso$yf

  ymax <- max(y_fit, na.rm = TRUE)
  if (!is.finite(ymax) || ymax <= 0) return(NA_real_)

  y_half <- ymax / 2

  # first point where fitted curve reaches half-max
  idx <- which(y_fit >= y_half)[1]
  if (is.na(idx)) return(NA_real_)

  if (idx == 1L) return(d$t[1])

  t1 <- d$t[idx - 1L]
  t2 <- d$t[idx]
  y1 <- y_fit[idx - 1L]
  y2 <- y_fit[idx]

  if (y2 == y1) return(mean(c(t1, t2)))

  t1 + (y_half - y1) * (t2 - t1) / (y2 - y1)
}

t50_dt <- lc_MS_SS_KR_PNAS_dt[oxygen == "21pc"][, .(
  t50 = calc_t50(.SD, time_col = "induction", y_col = "Stoichiometry", inf_value = 100),
  n = .N
), by = .(gene_name, aa_pos)]

t50_dt[, `:=`(
  t50_bin = case_when(
    t50 <= 4 ~ "0-4h",
    t50 <= 8 ~ "4-8h",
    TRUE ~ ">8h"
  ) %>% factor(levels = c("0-4h", "4-8h", ">8h"))
)]

t50_dt

```

# 2.10.6 Kinetics 

```{r kinetics}

kinetics_data <- pc2wt_21pc[!(induction %in% c("0h", "18h"))][WT_stoichiometry > 0]

ggplot(
  data = kinetics_data,
  aes(
    x = WT_stoichiometry,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_point() +
  facet_grid(~ induction2) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 1) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))

kinetics_out <- data.table()

for(i2 in kinetics_data[, unique(induction2)]){
  c1 <- kinetics_data[induction2 == i2] %$%
    cor.test(x = WT_stoichiometry, y = Stoichiometry / WT_stoichiometry, method = "spearman")
  
  kinetics_out <- rbind(kinetics_out, data.table(induction2 = i2, rho = c1$estimate, p_val = c1$p.value, n = nrow(kinetics_data[induction2 == i2])))
}

kinetics_out[, padj := p.adjust(p_val, method = "holm")]

kinetics_out


kinetics_data <- merge(
  kinetics_data,
  t50_dt,
  by = c("gene_name", "aa_pos")
)

ggplot(
  data = kinetics_data[!duplicated(paste(gene_name, aa_pos))],
  aes(
    x = t50_bin,
    y = WT_stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

kinetics_data[!duplicated(paste(gene_name, aa_pos))][, table(gene_name, t50_bin)]

```

# 2.10.7 XIC values against MS_SS, MS_KR_1 and PNAS stoichiometry data. 

```{r XIC_vs_stoic}

#-------------
# XIC vs stoic
#-------------

# Load XIC data
xic_MS_SS <- fread(file.path(data.dir, "xic_MS_SS.csv"))
xic_MS_SS[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  TRUE ~ induction
)]

on_lc_MS_SS_KR_PNAS_dt <- copy(lc_MS_SS_KR_PNAS_dt)[, induction2 := case_when(
  induction == "18h" ~ "18-24h",
  induction == "24h" ~ "18-24h",
  #induction == "Inf" ~ "18-24h",
  TRUE ~ induction
)]

# on_lc_MS_SS_KR_PNAS_dt <- on_lc_MS_SS_KR_PNAS_dt[induction != "24h"]

# Merge XIC data and stoichiometry data
xic_stoic <- merge(on_lc_MS_SS_KR_PNAS_dt, 
                   xic_MS_SS,
                   by = c("induction2", "oxygen", "gene_name", "aa_pos"))

xic_stoic[, XIC := XIC/100]

# Scatter plot - XIC vs stoichiometry
ggplot(
  xic_stoic,
  aes(
    x = XIC,
    y = Stoichiometry
  )
) + 
  geom_point(size = 2) +
  theme_classic_2() +
  theme(aspect.ratio = 1) +
  coord_cartesian(xlim = c(0, 0.7), ylim = c(0, 0.7)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))

xic_stoic %$%
cor.test(
  XIC,
  Stoichiometry
)

m_xic_stoic <- melt(
  xic_stoic,
  id.vars = c("gene_name", "aa_pos", "oxygen", "induction2"),
  measure.vars = c("Stoichiometry", "XIC"),
  variable.name = "type",
  value.name = "stoichiometry"
)

m_xic_stoic[, `:=`(
  oxygen = factor(oxygen, levels = rev(c("1pc", "4pc", "21pc"))),
  induction2 = factor(induction2, levels = c("4h", "8h", "18-24h"))
)]


```


```{r xic_per_pos, fig.width=14, fig.height=4.5}

ggplot(
  m_xic_stoic,
  aes(
    x = induction2,
    y = stoichiometry,
    color = type
  )
) + 
  facet_grid(~ gene_name + aa_pos + oxygen, scale = "free", space = "free") +
  geom_point(size = 4) +
  theme_classic_2() +
  coord_cartesian(ylim = c(0, 0.7)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_manual(values = c("XIC" = "black", "Stoichiometry" = "darkred"))

```

# 2.10.8 Interaction

```{r interaction}

t50_dt[gene_name == "BRD4"][aa_pos > 534 & aa_pos < 560]

calculate_low_hyl_stoic_with_high_hylhiometry <- function(input_per_pos_data, high_stoic_sites, low_stoic_sites, sample_group){
  
  peptide_with_high_sotic_sites <- input_per_pos_data[paste(protein_accession, aa_pos) %in% high_stoic_sites[, paste(protein_accession, aa_pos)]][, unique(peptide_id)]
  peptide_with_low_sotic_sites <- input_per_pos_data[paste(protein_accession, aa_pos) %in% low_stoic_sites[, paste(protein_accession, aa_pos)]][, unique(peptide_id)]
  
  hydroxy_input_per_pos_data <- copy(input_per_pos_data)[peptide_id %in% intersect(peptide_with_high_sotic_sites, peptide_with_low_sotic_sites)]
  
  peptide_with_multiple_hydroxyK_sites <- hydroxy_input_per_pos_data[, .N, by = list(peptide_id)][N > 1, unique(peptide_id)]
  
  sl_input_per_pos_data <- input_per_pos_data[peptide_id %in% peptide_with_multiple_hydroxyK_sites]
  
  interaction_summary <- data.table()
  
  for(high_hyl_site in high_stoic_sites[, paste(protein_accession, aa_pos)]){
    for(low_hyl_site in low_stoic_sites[, paste(protein_accession, aa_pos)]){
      
      peptide_with_both_sites <- sl_input_per_pos_data[(paste(protein_accession, aa_pos) %in% high_hyl_site) | (paste(protein_accession, aa_pos) %in% low_hyl_site)] %>%
        {.[, .N, by = peptide_id][N > 1, unique(peptide_id)]}
      
      if(length(peptide_with_both_sites) > 1){
        stoic_sl_input_per_pos_data <- sl_input_per_pos_data[peptide_id %in% peptide_with_both_sites]
        
        total_raw_stoic <- stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site & ptm %in% c("[Oxidation (K) DI]", "[Oxidised Propionylation (K) DI]"), sum(peak_intensity)] /
          stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site , sum(peak_intensity)]
        
        peptide_with_hyl_in_high <- stoic_sl_input_per_pos_data[
          ptm %in% c("[Oxidation (K) DI]", "[Oxidised Propionylation (K) DI]")
        ][paste(protein_accession, aa_pos) == high_hyl_site, unique(peptide_id)]
        
        highK_stoic_sl_input_per_pos_data <- stoic_sl_input_per_pos_data[peptide_id %in% peptide_with_hyl_in_high]
        
        highHyl_stoic <- highK_stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site & ptm %in% c("[Oxidation (K) DI]", "[Oxidised Propionylation (K) DI]"), sum(peak_intensity)] /
          highK_stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site , sum(peak_intensity)]
        
        no_highK_stoic_sl_input_per_pos_data <- stoic_sl_input_per_pos_data[!(peptide_id %in% peptide_with_hyl_in_high)]
        
        low_hyl_stoic_without_high_hyl <- no_highK_stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site & ptm %in% c("[Oxidation (K) DI]", "[Oxidised Propionylation (K) DI]"), sum(peak_intensity)] /
          no_highK_stoic_sl_input_per_pos_data[paste(protein_accession, aa_pos) == low_hyl_site , sum(peak_intensity)]
        
        interaction_summary <- rbind(
          interaction_summary, 
          data.table(
            protein_accession = str_split_fixed(high_hyl_site, " ", 2)[, 1], 
            high_stoic_aa_pos = str_split_fixed(high_hyl_site, " ", 2)[, 2] %>% as.integer, 
            low_stoic_aa_pos = str_split_fixed(low_hyl_site, " ", 2)[, 2] %>% as.integer, 
            low_hyl_total_stoic = total_raw_stoic, 
            low_hyl_stoic_with_high_hyl = highHyl_stoic,
            low_hyl_stoic_without_high_hyl,
            n = length(peptide_with_both_sites),
            sample_group
          )
        )
      } else {}
    }
  }
  
  interaction_summary <- interaction_summary[low_hyl_total_stoic > 0]
  
  return(interaction_summary)
}

t50_dt[gene_name == "BRD4"][aa_pos > 534 & aa_pos < 560]

high_stoic_sites <- data.table(protein_accession = "O60885", aa_pos = c(535, 538, 548)) # 535 included here for computation purpose
low_stoic_sites <- data.table(protein_accession = "O60885", aa_pos = c(535, 537, 550, 552))


dataA_per_pos <- fread(file.path(results.dir, "p2-analysis-setting/MQ_DI_noH2O/DI_noH2O_data-A_trp_m7_v7_def_all_per_pos_data.csv"))[gene_name %in% c("BRD2", "BRD3", "BRD4")] %>%
  {.[sample_name == "JQ1_HeLaWT_derivatised"]}
dataA_interaction_summary <- calculate_low_hyl_stoic_with_high_hylhiometry(input_per_pos_data = dataA_per_pos, high_stoic_sites, low_stoic_sites, sample_group = "WT_PNAS2022")

MS_KR_1_per_pos <- fread(file.path(results.dir, "p2-analysis-setting/MS_KR_1_noH2O/MS_KR_1_noH2O_all_per_pos_data.csv"))[gene_name %in% c("BRD2", "BRD3", "BRD4")] 
MS_KR_1_WT_per_pos <- MS_KR_1_per_pos[sample_name == "HeLaWT_NA_N_NA"]
MS_KR_1_WT_interaction_summary <- calculate_low_hyl_stoic_with_high_hylhiometry(input_per_pos_data = MS_KR_1_WT_per_pos, high_stoic_sites, low_stoic_sites, sample_group = "WT")

MS_KR_1_24h_per_pos <- MS_KR_1_per_pos[sample_name == "HeLaiJMJD6_Dox_N_NA"]
MS_KR_1_24h_interaction_summary <- calculate_low_hyl_stoic_with_high_hylhiometry(input_per_pos_data = MS_KR_1_24h_per_pos, high_stoic_sites, low_stoic_sites, sample_group = "iJMJD6 24h")

MS_SS_1_per_pos <- fread(file.path(results.dir, "p2-analysis-setting/MS_SS_noH2O/MS_SS_all_per_pos_data.csv")) %>%
  {rbind(.[gene_name %in% c("BRD4")][sample_name == "18h_21pc_BRD4"], .[gene_name %in% c("BRD2", "BRD3")][sample_name == "18h_21pc_BRD23"])}
MS_SS_1_interaction_summary <- calculate_low_hyl_stoic_with_high_hylhiometry(input_per_pos_data = MS_SS_1_per_pos, high_stoic_sites, low_stoic_sites, sample_group = "iJMJD6 18h")

all_interaction_summary <- rbindlist(
  list(
    dataA_interaction_summary, MS_KR_1_WT_interaction_summary, MS_KR_1_24h_interaction_summary, MS_SS_1_interaction_summary 
  )
)

## Stats
interaction_out_data <- data.table()

all_sites_to_analyse <- all_interaction_summary[!duplicated(paste(protein_accession, high_stoic_aa_pos, low_stoic_aa_pos))][, .(protein_accession, high_stoic_aa_pos, low_stoic_aa_pos)]

for(irow in 1:nrow(all_sites_to_analyse)){
  
  sl_all_interaction_summary <- merge(
    all_interaction_summary,
    all_sites_to_analyse[irow],
    by = c("protein_accession", "high_stoic_aa_pos", "low_stoic_aa_pos")
  )
  
  if(nrow(sl_all_interaction_summary) > 2){
    
    t_out <- sl_all_interaction_summary %$%
      t.test(low_hyl_stoic_without_high_hyl, low_hyl_stoic_with_high_hyl, paired = TRUE)
    
    m_all_interaction_summary <- melt(
      sl_all_interaction_summary, 
      id.vars = c("protein_accession", "high_stoic_aa_pos", "low_stoic_aa_pos", "n"), 
      measure.vars = c("low_hyl_stoic_without_high_hyl", "low_hyl_stoic_with_high_hyl"),
      variable.name = "type",
      value.name = "stoichiometry"
    )
    
    interaction_out_data <- rbind(
      interaction_out_data,
      data.table(all_sites_to_analyse[irow], p_val = t_out$p.value, n = nrow(sl_all_interaction_summary))
    )} else {}
}


interaction_out_data

all_interaction_summary[!(high_stoic_aa_pos == 550)] %$%
  t.test(low_hyl_stoic_without_high_hyl, low_hyl_stoic_with_high_hyl, paired = TRUE)

## Plot
m_all_interaction_summary <- melt(
  all_interaction_summary, 
  id.vars = c("protein_accession", "high_stoic_aa_pos", "low_stoic_aa_pos", "n", "sample_group"), 
  measure.vars = c("low_hyl_stoic_with_high_hyl", "low_hyl_stoic_without_high_hyl"),
  variable.name = "type",
  value.name = "stoichiometry"
)

m_all_interaction_summary[, type := gsub("low_hyl_stoic_", "", type)]

ggplot(
  data = m_all_interaction_summary[!(high_stoic_aa_pos == 550)],
  aes(
    x = type,
    y = stoichiometry,
    group = paste(sample_group, high_stoic_aa_pos, low_stoic_aa_pos),
    color = paste(high_stoic_aa_pos, low_stoic_aa_pos)
  )
) +
  geom_line(color = "black") +
  geom_point(size = 4) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_color_bright() +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  theme_classic_2() +
  theme(aspect.ratio = 4)

ggplot(
  data = m_all_interaction_summary[high_stoic_aa_pos == 535],
  aes(
    x = type,
    y = stoichiometry,
    group = paste(sample_group, high_stoic_aa_pos, low_stoic_aa_pos)
  )
) +
  geom_line(color = "black") +
  geom_point(size = 4) +
  scale_color_bright() +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  coord_cartesian(ylim = c(0, 1)) +
  theme_classic_2() +
  theme(aspect.ratio = 4) +
  ggtitle("BRD4 K535-K537 interaction")

```

# 2.10.9 Oxygen sensitivity - Comparing hypoxia stoichiometry under varying dox incubation (+ diagnostic ion)

```{r normoxia_vs_varying_hypoxia_pc_overnightdox}

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
# Data subset to overnight (18/24h) dox induction

on_data <- lc_MS_SS_KR_PNAS_dt[induction2 %in% c("18-24h", "Inf")][
  sample_group != "iJ6_18h_21pc_SS" # As iJ6_24h_21pc_KR exists
]


#---------------------------------------------
# Stoichiometry relative to WT values
#---------------------------------------------

# subset data to WT normoxia (MS_KR_1 sample)
wt_on <- on_data[induction2 == "Inf", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_on, old = "Stoichiometry", "WT_stoichiometry")

# Merge MS_SS normoxia data with MS_KR_1 WT normoxia data
pc2wt_on <- merge(
  on_data[induction2 != "Inf"],
  wt_on,
  by = c("protein_accession", "aa_pos")
)

pc2wt_on <- copy(pc2wt_on)[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
pc2wt_on <- pc2wt_on[data_size_per_site == max(pc2wt_on[, data_size_per_site])]

pc2wt_on <- pc2wt_on[WT_stoichiometry > 0]
pc2wt_on[, `:=`(
  oxygen = factor(oxygen, c("21pc", "4pc", "1pc", "01pc"))
)]

ggplot(
  data = pc2wt_on,
  aes(
    x = oxygen,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  ggtitle("Min WT stoichiometry > 0")

pc2wt_on[, .N, by = oxygen]


# Boxplot - comparing normoxia versus hypoxia
ggplot(
  data = pc2wt_on,
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.5)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

oxygen_t_test <- data.table()

for(i in pc2wt_on[, unique(oxygen)]){
  t.out <- t.test(
    pc2wt_on[oxygen == i][, Stoichiometry/WT_stoichiometry],
    pc2wt_on[oxygen == "21pc"][, Stoichiometry/WT_stoichiometry],
    paired = TRUE
  )
  oxygen_t_test <- rbind(
    oxygen_t_test,
    data.table(induction = i, p_value = t.out$p.value, n = nrow(pc2wt_on[data_size_per_site == max(pc2wt_on[, data_size_per_site])][oxygen == i]))
  )  
}

oxygen_t_test[, padj := p.adjust(p_value, method = "holm")]
oxygen_t_test[, significance := case_when(
  padj < 0.005 ~ "**",
  padj < 0.05 ~ "*",
  TRUE ~ "N.S."
)]

oxygen_t_test


```

# 2.10.10 Relationship of the stoichiometry in WT and changes in stoichiometry by hypoxia

```{r normoxia_vs_hypoxia_scatterplot}

ggplot(
  data = pc2wt_on[oxygen != "21pc"],
  aes(
    x = WT_stoichiometry,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_smooth(method = "lm", se = FALSE) +
  geom_point(size = 2) +
  facet_grid(~ oxygen) +
  theme(
    aspect.ratio = 1,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
  scale_x_continuous(labels = scales::percent_format(accuracy = 1))


oxygen_out <- data.table()

for(o2 in pc2wt_on[oxygen != "21pc"][, unique(oxygen)]){
  c2 <- pc2wt_on[oxygen == o2] %$%
    cor.test(
      WT_stoichiometry, Stoichiometry / WT_stoichiometry, method = "spearman"
    )
  
    oxygen_out <- rbind(oxygen_out, data.table(oxygen = o2, rho = c2$estimate, p_val = c2$p.value, n = nrow(pc2wt_on[oxygen == o2])))
}

oxygen_out[, padj := p.adjust(p_val, method = "holm")]

oxygen_out

```


```{r hypoxic_sensitivity}


# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
on_01pc_MS_KR_1_data <- lc_MS_SS_KR_PNAS_dt[induction2 %in% c("18-24h", "Inf")][
  sample_group == "iJ6_24h_01pc_KR"
]
setnames(on_01pc_MS_KR_1_data, old = "Stoichiometry", "ON01_stioichiomtry")

wt_on <- lc_MS_SS_KR_PNAS_dt[sample_group == "WT_Inf_21pc_KR"]
setnames(wt_on, old = "Stoichiometry", "WT_stoichiometry")

merge.dts <- function(dt_list, by_columns){Reduce(function(...) merge(..., by = by_columns), dt_list)}

sl.columns <- c("protein_accession", "aa_pos")

ref_dts <- merge.dts(
  dt_list = list(
    on_01pc_MS_KR_1_data[, c(sl.columns, "ON01_stioichiomtry"), with = FALSE],
    wt_on[, c(sl.columns, "WT_stoichiometry"), with = FALSE]
  ),
  by = sl.columns
)

non_ref_lc_MS_SS_KR_PNAS_dt <- lc_MS_SS_KR_PNAS_dt[
  !(sample_group %in% c("iJ6_24h_01pc_KR", "WT_Inf_21pc_KR", "iJ6_0h_21pc_SS", "iJ6_0h_21pc_KR", "iJ6_18h_21pc_SS"))
]

induction2vsref_21pc <- merge(
  non_ref_lc_MS_SS_KR_PNAS_dt,
  ref_dts,
  by = sl.columns
)

induction2vsref_21pc <- induction2vsref_21pc[WT_stoichiometry > 0]
## induction2vsref_21pc <- induction2vsref_21pc[sample_group %in% c("iJ6_4h_21pc_SS", "iJ6_8h_21pc_SS", "iJ6_24h_21pc_KR")]
induction2vsref_21pc[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]
induction2vsref_21pc <- induction2vsref_21pc[data_size_per_site == max(induction2vsref_21pc[, data_size_per_site])]

induction2vsref_21pc[, .N, by = list(sample_group, gene_name)][order(sample_group)]

ggplot(
  data = induction2vsref_21pc,
  aes(
    x = ON01_stioichiomtry,
    y = Stoichiometry
  )
) +
  # geom_smooth(method = "lm", se = FALSE) +
  geom_abline(intercept = 0, slope = 1) +
  geom_point() +
  facet_grid(oxygen ~ induction2) +
  theme(aspect.ratio = 1) +
  coord_cartesian(ylim = c(0, 1), xlim = c(0, 1))

rsq_out_dt <- data.table()

for(o2 in induction2vsref_21pc[, unique(oxygen)]){
  for(i2 in induction2vsref_21pc[, unique(induction2)]){
    rsq_dt <- induction2vsref_21pc[
      oxygen == o2 & induction2 == i2
    ]
    rsq_dt[, mean_St := mean(Stoichiometry)]
    rsq <- 1 - sum(rsq_dt[, (Stoichiometry - ON01_stioichiomtry)^2]) / sum(rsq_dt[, (Stoichiometry - mean_St)^2])
    rsq_out_dt <- rbind(
      rsq_out_dt, data.table(oxygen = o2, induction2 = i2, rsq = rsq, n = nrow(rsq_dt))
    )
  }
}

rsq_out_dt

pc2wt_on_kinetics <- merge(
  pc2wt_on,
  t50_dt,
  by = c("gene_name", "aa_pos")
)

library("ggbeeswarm")

ggplot(
  pc2wt_on_kinetics[n == 7][oxygen != "21pc"],
  aes(
    x = t50,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_point(size = 2) +
  facet_grid(~ oxygen) +
  theme(
    aspect.ratio = 1,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  coord_cartesian(xlim = c(0, 24)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

ggplot(
  pc2wt_on_kinetics[n == 7][oxygen != "21pc"],
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  geom_beeswarm() +
  facet_grid(~ t50_bin) +
  theme(
    aspect.ratio = 3,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  coord_cartesian(ylim = c(0, 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# QC
ggplot(
  pc2wt_on_kinetics[n == 7],
  aes(
    x = oxygen,
    y = Stoichiometry / WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ t50_bin) +
  theme(
    aspect.ratio = 3,
    legend.position = NULL
  ) +
  ylab("Stoichiometry [% to WT]") +
  scale_x_discrete(guide = guide_axis(angle = 90)) +
  coord_cartesian(ylim = c(0, 2)) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


pc2wt_on_kinetics[n == 7][, .N, by = list(oxygen, t50_bin)]

```

# Session information

```{r session_Info}

sessioninfo::session_info()

```
