---
title: "2-8. Lysine hydroxylations in varying O2 pc and dox induction time - data visualisation"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines the oxygen sensitivity of lysine hydroxylations in BRD proteins and their respective sites. This script is for visualizing the calculated stoichiometry. 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```


# 2.7.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```


# 2.7.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

# 2.7.3 Calculation of Stoichiometry with diagnostic ion in hypoxia and normoxia

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

```

# 2.7.4 Plotting Stoichiometry values of hypoxia and normoxia data with re-expression of JMJD6 (+ diagnostic ion)

```{r Define_important_functions}

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

#-------------------------------------
# Contrast function - by sample group
#-------------------------------------

contrast_hydroxylation_by_sample_group<- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_group, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, sample_group)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ sample_group, # values in oxygen_stat become into separate columns (Normoxia & Hypoxia)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

#---------------------------------
# Contrast function - by genotype
#---------------------------------

contrast_hydroxylation_by_genotype <- function(all_stoic_dt, min.psm = 2){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    genotype = str_extract(sample_name, "(?<=HeLa)(WT|JMJD6KO)"), # new column - Genotype, extract WT or JMJD6KO from sample_name column
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_name, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than min.psm (def. 2) for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > min.psm
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, genotype)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ genotype, # values in genotype become into separate columns (WT & JMJD6KO)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}
```


```{r load_MS_KR_1_MS_SS_data}

#-----------------------------
# Load MS_KR_1 stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_KR1 data (noH2O loss)
 MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_SS data 
 MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS_noH2O"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]


#-----------------------------
# Load PNAS stoichiometry data
#-----------------------------

pnas2022_stoic_dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022_stoic_dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022_stoic_dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022_stoic_dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

# Adding columns to pnas2022_stoic_dt metadata
pnas2022_stoic_dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos), # combines data from protein_accession and aa_pos
  ptm = fifelse(curated_oxK_site, "[Oxidation (K)]", ""), # add the "[Oxidation (K)]" in ptm column, if curated oxK site is TRUE
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>% #replace sample names as mentioned (from "Hela_" to "HeLa")
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)} 
)]

```


```{r merge_MS_SS_KR_data}

# Filter MS_KR_1 data to retain only BRD2, 3, 4 data
MS_KR1_BRD_dt <- MS_KR1_stoic_dt[
    gene_name %in% c("BRD2", "BRD3", "BRD4") & 
    aa == "K"]

# Filter MS_SS data to retain only BRD2, 3, 4 data 
MS_SS_BRD_dt <-  MS_SS_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")&
    aa == "K"]

# subset the data to curated_oxK_site = TRUE, aa = K and BRD proteins
pnas2022_stoic_dt <-  pnas2022_stoic_dt[
  curated_oxK_site == TRUE &
  ptm == "[Oxidation (K)]" &
  gene_name %in% c("BRD2", "BRD3", "BRD4") &
  aa == "K"
]

# subset rows with diagnostic ion from MS_SS and MS_KR data
DI_sites <- rbind(
  MS_KR1_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  MS_SS_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)]
)

# remove duplicate sites 
DI_sites <- DI_sites[!duplicated(paste(protein_accession, aa_pos))]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_dt <- rbindlist(list(MS_SS_BRD_dt,
                         MS_KR1_BRD_dt),
                         use.names = TRUE)

# Merge MS_SS_KR data to only have unique diagnostic ion sites
MS_SS_KR_dt <- merge(
  DI_sites,
  MS_SS_KR_dt,
  by = c("protein_accession", "aa_pos")
)

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_KR_dt <- MS_SS_KR_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Column "sample group" created to simpilfy sample names
MS_SS_KR_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR"
  )
)]

# Using contrast function on MS_SS_KR raw stoic data
c_MS_SS_KR_dt <- contrast_hydroxylation_by_sample_group(MS_SS_KR_dt)

# Factor the gene names
c_MS_SS_KR_dt[, `:=`(
  gene_name =  
    factor(gene_name)
)]

```


```{r stoichiometry_BRD_proteins}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented 
# in the PNAS2022 paper. 

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_dt[grepl("18h|24h|WT|0h", sample_group) & gene_name == "BRD2"],
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_dt[grepl("18h|24h|WT|0h", sample_group) & gene_name == "BRD3"],
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_dt[grepl("18h|24h|WT|0h", sample_group) & gene_name == "BRD4"],
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

```


```{r overall_MS_SS_KR_comparison}

# Change wide format to long format 
long_c_MS_SS_KR_dt <- melt(
  c_MS_SS_KR_dt,
  id.vars = c("protein_accession", "gene_name", "aa_pos"),
  value.name = "Stoichiometry", 
  variable.name = "sample_group"
  #value.var = grep("_", colnames(c_MS_SS_KR_all_dt), value = TRUE)
)

# Separate the contents of sample_group column into new columns 
long_c_MS_SS_KR_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2] %>%
    factor(levels = c("Inf", "0h", "4h", "8h", "18h", "24h")),
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# subset rows that do not contain NA values 
long_c_MS_SS_KR_dt <- long_c_MS_SS_KR_dt[!is.na(Stoichiometry)]

# counting the data per sample and size 
long_c_MS_SS_KR_dt[, data_size_per_sample := .N, by = list(sample_group)]
long_c_MS_SS_KR_dt[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]

# reorder the oxygen levels
long_c_MS_SS_KR_dt <-  long_c_MS_SS_KR_dt[, oxygen := factor(
  oxygen,
  levels = c("01pc", "1pc", "4pc", "21pc")
)]

```


```{r normoxia_varying_dox_induction}

#---------------------------------------------
# Normoxia samples under varying dox induction
#---------------------------------------------

# subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- long_c_MS_SS_KR_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stioichiomtry")

# Merge MS_SS normoxia data with MS_KR_1 WT normoxia data
pc2wt_21pc <- merge(
  long_c_MS_SS_KR_dt[
    cell == "iJ6" &
    oxygen == "21pc"
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)

# Boxplot - comparing normoxia in different dox incubation timings
ggplot(
  data = pc2wt_21pc[data_size_per_site == max(pc2wt_21pc[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry / WT_stioichiomtry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r Dox_plus_minus_JMJD6_O2_levels}

# Boxplot - Stoichiometry of Dox + and - JMJD6 under O2 levels
ggplot(
  data = long_c_MS_SS_KR_dt[data_size_per_site == max(long_c_MS_SS_KR_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))



long_c_MS_SS_KR_21pc_dt <- long_c_MS_SS_KR_21pc_dt[oxygen == "21pc"]


# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2
ggplot(
  data = long_c_MS_SS_KR_21pc_dt[data_size_per_site == max(long_c_MS_SS_KR_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r normoxia_vs_varying_hypoxia_pc_overnightdox}

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
# Data subset to overnight (18/24h) dox induction

long_MS_SS_KR_O2p_dt <- long_c_MS_SS_KR_dt[grepl("18h|24h|WT", sample_group)]

# reorder the sample_group according to oxygen levels
long_MS_SS_KR_O2p_dt[, sample_group :=
  factor(sample_group,
         levels = c("WT_Inf_21pc_KR","iJ6_24h_01pc_KR","iJ6_18h_1pc_SS","iJ6_18h_4pc_SS","iJ6_18h_21pc_SS","iJ6_24h_21pc_KR"))
]

ggplot(
  data = long_MS_SS_KR_O2p_dt, 
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


# 2.7. Plotting XIC values against MS_SS, MS_KR_1 and PNAS2022 stoichiometry data . 

```{r XIC_vs_stoic}

#-------------
# XIC vs stoic
#-------------

# Load XIC data
xic_MS_SS <- fread(file.path(data.dir, "xic_MS_SS.csv"))

# Merge XIC data and stoichiometry data
xic_stoic <- merge(long_c_MS_SS_KR_dt, 
                   xic_MS_SS,
                   by = c("induction", "oxygen", "gene_name", "aa_pos"))

# Scatter plot - XIC vs stoichiometry
ggplot(
  xic_stoic,
  aes(
    x = XIC,
    y = Stoichiometry
  )
) + 
  geom_point() +
  theme_classic()

```













```{r Normoxia_vs_Hypoxia_in_different_O2%_18h_dox}

#Add new columns to simplify sample names into hypoxia and normoxia category
MS_SS_KR_dt[, `:=`(
  oxygen_levels =
    factor(
      fcase(
        grepl("^18h_21pc", sample_name),
        "HeLaNormoxia_18h",
        grepl("^18h_4pc", sample_name),
        "HeLaHypoxia_18h_4pc",
        grepl("^18h_1pc", sample_name),
        "HeLaHypoxia_18h_1pc",
        grepl("^HeLai", sample_name),
        "HeLaHypoxia_24h_0.1pc",
        default = NA_character_
      ),
      levels = c("HeLaNormoxia_18h", "HeLaHypoxia_18h_4pc", "HeLaHypoxia_18h_1pc", "HeLaHypoxia_24h_0.1pc")
    )
)]

# Use contrast function on 'MS_SS_KR_dt' to set NA to 0, only if the sample contains an #[oxidation K] 
MS_SS_KR_hydroxy_dt <-  contrast_hydroxylation_by_oxygen_stat(MS_SS_KR_dt)

# Generate data table with WT and JMJD6KO with diagnostic peaks
DI_info <- MS_SS_KR_dt[
  , .(protein_accession, aa_pos, diagnostic_peak) # filter these columns and create a separate data table 
][
  order(protein_accession, aa_pos, -diagnostic_peak) # order the data based on the highest diagnostic peak 
][                                              
  !duplicated(paste(protein_accession, aa_pos)) # Keep only the first row for each unique protein and amino acid position,
  # removing duplicates while preserving the row with the highest diagnostic_peak (from previous sorting)
]

# merge the modified data table 'DI_info' with the metadata 'MS_KR_1_hydroxy_dt'
MS_SS_KR_hydroxy_dt <- merge(MS_SS_KR_hydroxy_dt, DI_info, by = c("protein_accession", "aa_pos"))

#Filter data table with presence of diagnostic_peak and WT stoichiometry values
MS_SS_KR_hydroxy_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

# filter data to have hydroxylysine with diagnostic ion 
MS_SS_KR_hydroxy_dt <- MS_SS_KR_hydroxy_dt[is_diagnostic_peak == TRUE]

# Factor the gene names
MS_SS_KR_hydroxy_dt[, `:=`(
  gene_name =  
    factor(gene_name)
)]

# Remove rows with NA values 
MS_SS_KR_hydroxy_dt <- na.omit(MS_SS_KR_hydroxy_dt)

```


```{r plot_stoic_normoxia_diffpc_hypoxia}

# Convert the wide format data into long format for plotting 
MS_hydroxy_dt <- melt(
  MS_SS_KR_hydroxy_dt, 
  id.vars = c("protein_accession", "aa_pos", "gene_name"),
  measure.vars = c("Normoxia","Hypoxia_18h_4pc", "Hypoxia_18h_1pc", "Hypoxia_24h_0.1pc"), 
  variable.name = "sample_name", value.name = "Stoichiometry")

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
ggplot(
  data = MS_hydroxy_dt,
  aes(
    x = sample_name,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r normoxia_vs_hypoxia_scatterplot}

# Scatter plot - Normoxia stoichiometry versus hypoxia at 4% - 18h dox incubation
ggplot(
  data = MS_SS_KR_hydroxy_dt, 
  aes(
    x = Normoxia,
    y = Hypo_18h_4pc_ratio
  )
) +
  geom_point(aes(colour = gene_name)) +
  facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(MS_SS_KR_hydroxy_dt[, Normoxia], MS_SS_KR_hydroxy_dt[, Hypo_18h_4pc_ratio], method = "spearma")

# Scatter plot - Normoxia stoichiometry versus hypoxia at 1% - 18h dox incubation
ggplot(
  data = MS_SS_KR_hydroxy_dt, 
  aes(
    x = Normoxia,
    y = Hypo_18h_1pc_ratio
  )
) +
  geom_point(aes(colour = gene_name)) +
  facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(MS_SS_KR_hydroxy_dt[, Normoxia], MS_SS_KR_hydroxy_dt[, Hypo_18h_1pc_ratio], method = "spearma")

# Scatter plot - Normoxia stoichiometry versus hypoxia at 0.1% - 24h dox incubation
ggplot(
  data = MS_SS_KR_hydroxy_dt, 
  aes(
    x = Normoxia,
    y = Hypo_24h_0.1pc_ratio
  )
) +
  geom_point(aes(colour = gene_name)) +
  facet_grid(~ gene_name, space = "free")+
  theme_classic() +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right")

# correlation test 
cor.test(MS_SS_KR_hydroxy_dt[, Normoxia], MS_SS_KR_hydroxy_dt[, Hypo_24h_0.1pc_ratio], method = "spearma")

```


```{r normoxia_non_subsetting_and_subsetting}

# Calculate the ratio of hypoxic stoichiometry relative to normoxia
MS_SS_KR_hydroxy_dt[, `:=`(
  Hypo_18h_4pc_ratio = Hypoxia_18h_4pc/Normoxia,
  Hypo_18h_1pc_ratio = Hypoxia_18h_1pc/Normoxia,
  Hypo_24h_0.1pc_ratio = Hypoxia_24h_0.1pc/Normoxia
)]

# Convert the wide format data into long format for plotting 
m_MS_SS_KR_hydroxy_dt <- melt(
  MS_SS_KR_hydroxy_dt, 
  id.vars = c("protein_accession", "aa_pos", "gene_name", "Normoxia"),
  measure.vars = c("Hypo_18h_4pc_ratio", "Hypo_18h_1pc_ratio", "Hypo_24h_0.1pc_ratio"), 
  variable.name = "sample_name", value.name = "Stoichiometry_ratio"
)

# Boxplot - hypoxia stoic pc relative to normoxia pc
ggplot(
  data = m_MS_SS_KR_hydroxy_dt,
  aes(
    x = sample_name,
    y = Stoichiometry_ratio
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  #facet_grid(~ gene_name) +
  ylab("Stoichiometry [% to normoxia]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Boxplot - hypoxia stoic pc relative to normoxia pc - normoxia subset 0.1
ggplot(
  data = subset(m_MS_SS_KR_hydroxy_dt, Normoxia >= 0.1),
  aes(
    x = sample_name,
    y = Stoichiometry_ratio
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  ggtitle("N% > 0.1") +
  ylab("Stoichiometry [% to normoxia]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


# 2.7.5 Comparing normoxia samples under varying dox incubation (+ diagnostic ion)

```{r load_contrast_data}

#-----------------------------
# Load PNAS stoichiometry data
#-----------------------------

pnas2022_stoic_dt <- fread(file.path(data.dir, "PNAS2022/long_K_stoichiometry_data.csv"))

# Replace old column names with new ones
setnames(
  pnas2022_stoic_dt, 
  old = c("uniprot_id", "position", "residue", "oxK_ratio", "data_source", "total_n_feature_oxK", "total_n_feature_K"), 
  new = c("protein_accession", "aa_pos", "aa", "stoichiometry", "sample_name", "sum_psm_mapped", "sum_psm_mapped_per_position")
)

# Merge the reference protein data with the PNAS stoichiometry data by protein accession ID
pnas2022_stoic_dt <- merge(
  ref_protein_dt[, .(protein_accession, gene_name)], # Take columns protein_accession and gene_name from the ref data 
  pnas2022_stoic_dt,
  by = "protein_accession" #combine both data.table by protein accession ID
)

# Adding columns to pnas2022_stoic_dt metadata
pnas2022_stoic_dt[, `:=`(
  accession_position = paste0(protein_accession, "_", aa_pos), # combines data from protein_accession and aa_pos
  ptm = fifelse(curated_oxK_site, "[Oxidation (K)]", ""), # add the "[Oxidation (K)]" in ptm column, if curated oxK site is TRUE
  sample_name = gsub("HeLa_", "HeLa", sample_name) %>% #replace sample names as mentioned (from "Hela_" to "HeLa")
    {gsub("HeLaJMJD6FLAG", "HeLaWT_JMJD6FLAG", .)} 
)]

# subset the data to curated_oxK_site = TRUE, aa = K and BRD proteins
pnas2022_stoic_dt <-  pnas2022_stoic_dt[
  curated_oxK_site == TRUE &
  ptm == "[Oxidation (K)]" &
  gene_name %in% c("BRD2", "BRD3", "BRD4") &
  aa == "K"
]

# Contrast non hydroxylated and hydroxylated lysines
# Take only the HeLa samples with JQ1 or J6 peptide pull-down
pnas_hydroxy_dt <- contrast_hydroxylation_by_genotype(
  pnas2022_stoic_dt[
    grepl("HeLa", sample_name) & (grepl("JQ1", sample_name) | grepl("J6pep", sample_name))
  ]
)


#--------------------------------
# Load MS_KR_1 stoichiometry data
#--------------------------------

# Filter MS_KR_1 data to only contain sample name = "HeLaiJMJD6_Dox_01O224h_NA"
MS_KR1_WT_dox_dt <- MS_KR1_stoic_dt[
  is_diagnostic_peak == TRUE &
    sample_name %in% c("HeLaWT_NA_N_NA", "HeLaiJMJD6_Dox_N_NA") &
    gene_name %in% c("BRD2", "BRD3", "BRD4") & 
    aa == "K"
]

# Add new columns to simplify sample names 
MS_KR1_WT_dox_dt[, `:=`(
  oxygen_levels =
    factor(
      fcase(
        grepl("HeLaWT_NA_N_NA", sample_name),
        "WT_JMJD6",
        grepl("HeLaiJMJD6_Dox_N_NA", sample_name),
        "24h_JMJD6",
        default = NA_character_
      ),
      levels = c("WT_JMJD6", "24h_JMJD6")
    )
)]

#--------------------------------
# Load MS_SS stoichiometry data
#--------------------------------

# Filter MS_SS data to retain only BRD2, 3, 4 data at 21pc of O2
MS_SS_21pc_dt <-  MS_SS_stoic_dt[
  is_diagnostic_peak == TRUE &
    grepl("21pc", sample_name) &
    gene_name %in% c("BRD2", "BRD3", "BRD4") &
    aa == "K"
]

# Add new columns to simplify sample names 
MS_SS_21pc_dt[, `:=`(
  oxygen_levels =
    factor(
      fcase(
        grepl("^18h", sample_name),
        "18h_JMJD6",
        grepl("^8h", sample_name),
        "8h_JMJD6",
        grepl("^4h", sample_name),
        "4h_JMJD6",
        default = NA_character_
      ),
      levels = c("18h_JMJD6", "8h_JMJD6","4h_JMJD6")
    )
)]

```


```{r MS_SS_KR_normoxia_vs_dox_incubation}

#-----------------------------------------
## Combine MS_KR_1 and MS_SS normoxia data
#-----------------------------------------
MS_normoxia_dt <- rbindlist(list(MS_SS_21pc_dt,
                         MS_KR1_WT_dox_dt),
                         use.names = TRUE)

# reorder oxygen levels
MS_normoxia_dt[, oxygen_levels := factor(
  oxygen_levels,
  levels = c("WT_JMJD6", "4h_JMJD6", "8h_JMJD6", "18h_JMJD6", "24h_JMJD6")
)]

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could false positives reported by the system 
MS_normoxia_dt <- MS_normoxia_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name)  & gene_name %in% c("BRD2", "BRD3")))
]

# Boxplot - Comparing normoxia stoic in different dox incubation
ggplot(
  data = MS_normoxia_dt,
  aes(
    x = oxygen_levels,
    y = stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  labs(title = "N in diff dox incubation") +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))
#coord_cartesian(ylim = c(0, 1.0))

# Boxplot - Comparing normoxia stoic in different dox incubation
ggplot(
  data = MS_normoxia_dt[stoichiometry >= 0.1],
  aes(
    x = oxygen_levels,
    y = stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  labs(title = "(N >= 0.1) in diff dox incubation") +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right")+
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


```

# 2.7.6 Comparing hypoxia samples under varying dox incubation (+ diagnostic ion)

```{r MS_SS_hypoxia_vs_dox_incubation}

# Filter MS_SS data to retain only BRD2, 3, 4 data 
MS_SS_hypo_dt <-  MS_SS_stoic_dt[is_diagnostic_peak == TRUE 
                                  & gene_name %in% c("BRD2", "BRD3", "BRD4") 
                                  & aa == "K"
]

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could false positives reported by the system 
MS_SS_hypo_dt <- MS_SS_hypo_dt[
  !(
    (grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
    (grepl("_BRD4$",  sample_name) & gene_name %in% c("BRD2", "BRD3"))
  )
]
# Subset the data 
MS_SS_hypo_dt <-  MS_SS_stoic_dt[is_diagnostic_peak == TRUE 
                                  & gene_name %in% c("BRD2", "BRD3", "BRD4") 
                                  & aa == "K"
]

# Categorize the samples based on O2 pc 
MS_SS_hypo_dt[, `:=`(
  O2_pc =
    factor(
      fcase(
       grepl("21pc", sample_name),
        "21pc",
        grepl("4pc", sample_name),
        "4pc",
        grepl("1pc", sample_name),
        "1pc",
        default = NA_character_
      ),
      levels = c("1pc", "4pc", "21pc")
    )
)]

# Categorize the samples based on dox incubation 
MS_SS_hypo_dt[, `:=`(
  dox_incubation =
    factor(
      fcase(
       grepl("^18h", sample_name),
        "18h",
        grepl("^8h", sample_name),
        "8h",
        grepl("^4h", sample_name),
        "4h",
        default = NA_character_
      ),
      levels = c("4h", "8h", "18h")
    )
)]

# Remove rows with NA values 
MS_SS_hypo_dt <- na.omit(MS_SS_hypo_dt)

# box_plot - MS_SS in diff dox incubation
ggplot(
  data = MS_SS_hypo_dt, 
  aes(
    x = dox_incubation,
    y = stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ O2_pc)+
  ggtitle("MS_SS in diff dox incubation") +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))



MS_SS_hypo_Nsub_dt <- MS_SS_hypo_dt[
  O2_pc == "21pc" & stoichiometry >= 0.1
]

MS_SS_hypo_Nsub_dt <- rbindlist(list(MS_SS_hypo_Nsub_dt,
                                     MS_SS_hypo_dt[
                                       !(O2_pc == "21pc" & stoichiometry >= 0.1)
                                     ]),
                                use.names = TRUE)

# box_plot - MS_SS in diff dox incubation N =>0.1
ggplot(
  data = MS_SS_hypo_Nsub_dt, 
  aes(
    x = dox_incubation,
    y = stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ O2_pc)+
  ggtitle("MS_SS in diff dox incubation (N subset)") +
  ylab("Stoichiometry [%]") +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


# Session information

```{r session_Info}

sessioninfo::session_info()

```
