---
title: "2-8. Lysine hydroxylations in varying O2 pc and dox induction time - data visualisation"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines the oxygen sensitivity of lysine hydroxylations in BRD proteins and their respective sites. This script is for visualizing the calculated stoichiometry. 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```

# 2.8.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.8.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p2_MS_SS_KR <- file.path(results.dir, "p2_MS_SS_KR")
# create.dirs(c(results.dir, p2_MS_SS_KR))

```

# 2.8.3 Defining functions and data preprocessing

```{r Define_important_functions}

# Read FASTA data 
all.protein.bs <- Biostrings::readAAStringSet(
  file.path(
    "/fast/AG_Sugimoto/reference/uniprot/human",
    "UP000005640_9606.fasta"
  )
)

# Define function - 'read_stoic_data'
read_stoic_data <- function(prefix, pre_prefix, post_fix = "", dir_path){
  # Read csv file with defined file path,  
  dt <- fread(file.path(dir_path, paste0(pre_prefix, prefix, "PTM_stoichiometry", post_fix, ".csv"))) 
  dt[, condition := gsub("_$", "", prefix)] # Removes '_' at the end of the prefix and creates a new column called condition
  return(dt)
}

#-------------------------------------
# Contrast function - by sample group
#-------------------------------------

contrast_hydroxylation_by_sample_group<- function(all_stoic_dt){
  wt_ko_dt <- all_stoic_dt
  
  # Add new columns into metadata
  wt_ko_dt[, `:=`(
    pos_id = paste0(protein_accession, "_", aa_pos), 
    sample_pos_id = paste0(sample_group, "_", protein_accession, "_", aa_pos)
  )]
  
  # Identify position with at least one oxidation event
  oxidation_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE), # checks for oxidation event
    unique(pos_id) # Returns unique pos_id with oxidation event
  ]
  oxidation_sample_ids <- wt_ko_dt[
    grepl("[Oxidation (K)]", ptm, fixed = TRUE),
    unique(sample_pos_id)
  ]
  
  # Collect non hydroxylated K information for the sites with hydroxylation
  no_hydroxyK_dt <- copy(wt_ko_dt[aa == "K"])
  no_hydroxyK_dt <- no_hydroxyK_dt[pos_id %in% oxidation_ids] %>%
    {.[!sample_pos_id %in% oxidation_sample_ids]} # If hydroxylation data exist, this is not necessary
  
  # Set stoichiometry and PSM count to zero for these positions and update the PTM label
  no_hydroxyK_dt[, `:=`(
    sum_psm_mapped = 0,
    stoichiometry = 0,
    ptm = "[Oxidation (K)]"
  )]
  
  # Combine oxidation data from both original and the newly flagged unmodified K data
  hydroxyK_dt <- rbind(
    wt_ko_dt[grepl("[Oxidation (K)]", ptm, fixed = TRUE)],
    no_hydroxyK_dt
  )
  
  # Filter psm mapped greater than 2 for higher confidence 
  hydroxyK_dt <- hydroxyK_dt[
    sum_psm_mapped_per_position > 2
  ]
  
  # Sort stoichiometry column from highest to lowest
  hydroxyK_dt <- hydroxyK_dt[order(stoichiometry, decreasing = TRUE)][
    !duplicated(paste0(protein_accession, gene_name, aa_pos, sample_group)) # Remove duplicate rows
  ]
  
  # Reshape 'd.hydroxyK_dt' from long to wide based on stoichiometry values  
  d.hydroxyK_dt <- dcast(
    hydroxyK_dt,
    protein_accession + gene_name + aa_pos ~ sample_group, # values in oxygen_stat become into separate columns (Normoxia & Hypoxia)
    value.var = "stoichiometry"
  )
  
  return(d.hydroxyK_dt)
}

```


```{r load_MS_KR_1_MS_SS_data}

#-----------------------------
# Load MS_KR_1 stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_KR1 data (noH2O loss)
 MS_KR1_stoic_dt <- read_stoic_data(
  prefix = "MS_KR_1_noH2O_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_KR_1_noH2O"))


# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_KR1_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#-----------------------------
# Load MS_SS stoichiometry data
#-----------------------------

## Read stoichiometry data for MS_SS data 
 MS_SS_stoic_dt <- read_stoic_data(
  prefix = "MS_SS_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MS_SS_noH2O"))

 #[Oxidation (K)] 

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MS_SS_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant (dataset A)
#--------------------------------------------------------------

 pnas2022_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
pnas2022_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]


#-----------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant (all dataset)
#-----------------------------------------------------------------

# Load sample run info data (MQ DI with no waterloss)
MQ_DI_noH2O_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_with_DI_noH2O" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_noH2O_run_info[, sample_id := 1:.N]

PNAS_all_stoic_dt <- lapply(
  MQ_DI_noH2O_run_info[, prefix],
  read_stoic_data,
  pre_prefix = "DI_noH2O_", 
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O")
) %>% rbindlist

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
PNAS_all_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]
  
```


```{r data_preprocessing}

# Filter MS_KR_1 data to retain only BRD2, 3, 4 data
MS_KR1_BRD_dt <- MS_KR1_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Filter MS_SS data to retain only BRD2, 3, 4 data 
MS_SS_BRD_dt <-  MS_SS_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Filter PNAS2022 data to retain only BRD2, 3, 4 data 
pnas2022_BRD_dt <-  pnas2022_stoic_dt[
  gene_name %in% c("BRD2", "BRD3", "BRD4")
]

# Combine MS_KR1_subset_dt and MS_SS_all_dt data
MS_SS_KR_PNAS_dt <- rbindlist(list(MS_SS_BRD_dt,
                         MS_KR1_BRD_dt,
                         pnas2022_BRD_dt),
                         use.names = TRUE)

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Column "sample group" created to simplify sample names
MS_SS_KR_PNAS_dt[, `:=`(
  sample_group = case_when(
    condition == "MS_SS" & grepl("minusDox", sample_name) ~ paste0("iJ6_0h_21pc_SS"),
    condition == "MS_SS"~ paste0("iJ6_", str_split_fixed(sample_name, "_", 3)[, 1], "_", str_split_fixed(sample_name, "_", 3)[, 2], "_SS"),
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR",
    sample_name == "JQ1_HeLaWT_derivatised" ~ "WT_Inf_21pc_PNAS", 
    sample_name == "JQ1_HeLaJMJD6KO_derivatised" ~ "iJ6_Inf_21pc_PNAS"
  )
)]

```

#

```{r publication_ready_tables}

# Subset the data to contain only Lysine
MS_KR1_stoic_K_dt <- MS_KR1_stoic_dt[aa == "K"]

# removing unsed data columns from data table
MS_KR1_stoic_K_dt <-  MS_KR1_stoic_K_dt[, c("sum_intensity_per_position",
                                            "sum_psm_mapped_per_position",
                                            "sum_the_number_of_peptide",
                                            "localization_prob",
                                            "score_for_localization",
                                            "best_localization_ms_ms_id",
                                            "best_localization_raw_file") := NULL]

# Column "sample group" created to simplify sample names
MS_KR1_stoic_K_dt[, `:=`(
  sample_group = case_when(
    sample_name == "HeLaiJMJD6_noDox_N_NA" ~ "iJ6_0h_21pc_KR",
    sample_name == "HeLaiJMJD6_Dox_N_NA" ~ "iJ6_24h_21pc_KR",
    sample_name == "HeLaWT_NA_N_NA" ~ "WT_Inf_21pc_KR", 
    sample_name == "HeLaiJMJD6_Dox_01O224h_NA" ~ "iJ6_24h_01pc_KR"
  )
)]

# Separate the contents of sample_group column into new columns 
MS_KR1_stoic_K_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2],
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# changing stoichiometry column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "stoichiometry"), "stoichiometry", after = 6)
)

# changing sample_group column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "sample_group"), "sample_group", after = 1)
)
 
# changing sample_group column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MS_KR1_stoic_K_dt, c("sample_name",
                              "sample_group",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "max_score",
                              "diagnostic_peak",
                              "is_diagnostic_peak",
                              "condition",
                              "cell",
                              "induction",
                              "oxygen",
                              "dataset"), 
         c("Sample_Name",
           "Sample_ID",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Max_score",
           "Diagnostic_Peak",
           "Diagnostic_peak_presence",
           "Experimental_condition",
           "Cell_type",
           "JMJD6_induction",
           "Oxygen_levels",
           "Dataset"))

 # install package for writing data table to excel 
install.packages('openxlsx')
 library("openxlsx")
 
 # Write to csv table to identify "K" with DI 
write.xlsx(MS_KR1_stoic_K_dt, file.path(results.dir, "p2_MS_SS_KR", "MS_KR1_stoic_K.xlsx"))

```

# 2.8.4 Plotting raw stoichiometry values of normoxia data with re-expression of JMJD6 

```{r stoichiometry_BRD2_at_21pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented 
# in the PNAS2022 paper. 

# subset data to specific BRD2 protein
MS_SS_KR_PNAS_BRD2_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
                                            gene_name == "BRD2"]


# reorder sample names
MS_SS_KR_PNAS_BRD2_dt <-  MS_SS_KR_PNAS_BRD2_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD2 <- MS_SS_KR_PNAS_BRD2_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD2, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD2.csv"))

```


```{r stoichiometry_BRD3_at_21pc}

# subset data to specific BRD3 protein
MS_SS_KR_PNAS_BRD3_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD23", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD3"]



MS_SS_KR_PNAS_BRD3_dt <-  MS_SS_KR_PNAS_BRD3_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD23",
             "8h_21pc_BRD23",
             "18h_21pc_BRD23",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD3 <- MS_SS_KR_PNAS_BRD3_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD3, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD3.csv"))

```


```{r stoichiometry_BRD4_at_21pc}

# subset data to specific BRD4 protein
MS_SS_KR_PNAS_BRD4_dt <- MS_SS_KR_PNAS_dt[grepl("Inf|0h|4h|8h|18h|24h", sample_group) & 
                                            grepl("21pc", sample_group) & 
                                            !sample_name %in% c("minusDox_BRD4", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised")&
                                            gene_name == "BRD4"]


# reorder sample names
MS_SS_KR_PNAS_BRD4_dt <-  MS_SS_KR_PNAS_BRD4_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_noDox_N_NA",
             "4h_21pc_BRD4",
             "8h_21pc_BRD4",
             "18h_21pc_BRD4",
             "HeLaiJMJD6_Dox_N_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD4 <- MS_SS_KR_PNAS_BRD4_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD4, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD4.csv"))

```

# 2.8.5 Plotting raw stoichiometry values of normoxia and hypoxia data with re-expression of JMJD6 

```{r stoichiometry_BRD2_at_hypoxia_pc}

# Plotting stoichiometry of BRD proteins in different O2% 
# Note: the amino acid region of the proteins is that of the ones represented in the PNAS2022 paper. 

# HeLaWT_NA_N_NA - WT_Inf_21pc_KR
# JQ1_HeLaWT_derivatised - WT_Inf_21pc_PNAS
# JQ1_HeLaJMJD6KO_derivatised - iJ6_Inf_21pc_PNAS
# HeLaiJMJD6_Dox_N_NA - iJ6_24h_21pc_KR
# 18h_4pc_BRD23 - iJ6_18h_4pc_SS
# 18h_1pc_BRD23 - iJ6_18h_1pc_SS
# HeLaiJMJD6_Dox_01O224h_NA - iJ6_24h_01pc_KR
# minusDox_BRD23 - iJ6_0h_21pc_SS
# HeLaiJMJD6_noDox_N_NA - iJ6_0h_21pc_KR


# HeLaWT_NA_N_NA - WT_Inf_21pc_KR
# JQ1_HeLaWT_derivatised - WT_Inf_21pc_PNAS
# JQ1_HeLaJMJD6KO_derivatised - iJ6_Inf_21pc_PNAS
# HeLaiJMJD6_Dox_N_NA - iJ6_24h_21pc_KR
# 18h_4pc_BRD23 - iJ6_18h_4pc_SS
# 18h_1pc_BRD23 - iJ6_18h_1pc_SS
# HeLaiJMJD6_Dox_01O224h_NA - iJ6_24h_01pc_KR
# HeLaiJMJD6_noDox_N_NA - iJ6_0h_21pc_KR

#------
# BRD2
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD2_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD2"]

# reorder sample names
MS_SS_KR_PNAS_BRD2_O2pc_dt <-  MS_SS_KR_PNAS_BRD2_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]

#accession = "P25440", plot_range = c(540, 590) BRD2
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD2_O2pc_dt,
  accession = "P25440",
  plot_range = c(540, 590), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD2_O2pc <- MS_SS_KR_PNAS_BRD2_O2pc_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD2_O2pc, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD2_O2pc.csv"))

```


```{r stoichiometry_BRD3_at_hypoxia_pc}

#------
# BRD3
#------
# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD3_O2pc_dt <- MS_SS_KR_PNAS_dt[
  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD23", "18h_21pc_BRD23","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD3"]


# reorder sample names
MS_SS_KR_PNAS_BRD3_O2pc_dt <-  MS_SS_KR_PNAS_BRD3_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD23",
             "18h_1pc_BRD23",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
]


# accession = "Q15059", plot_range = c(483, 533) BRD3
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD3_O2pc_dt,
  accession = "Q15059",
  plot_range = c(483, 533), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD3_O2pc <- MS_SS_KR_PNAS_BRD3_O2pc_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD3_O2pc, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD3_O2pc.csv"))

```


```{r stoichiometry_BRD4_at_hypoxia_pc}

#------
# BRD4
#------

# subset data to specific BRD protein
MS_SS_KR_PNAS_BRD4_O2pc_dt <- MS_SS_KR_PNAS_dt[  grepl("Inf|18h|24h|0h", sample_group) & 
    !sample_name %in% c("minusDox_BRD4", "18h_21pc_BRD4","HeLaiJMJD6_noDox_N_NA", "JQ1_HeLaWT_derivatised", "JQ1_HeLaJMJD6KO_derivatised") & 
    gene_name == "BRD4"]


#iJ6_0h_21pc_SS - minusDox_BRD23
# HeLaiJMJD6_noDox_N_NA

# reorder sample names
MS_SS_KR_PNAS_BRD4_O2pc_dt <-  MS_SS_KR_PNAS_BRD4_O2pc_dt[, sample_name := factor(
  sample_name, 
  levels = c("HeLaWT_NA_N_NA",
             "HeLaiJMJD6_Dox_N_NA",
             "18h_4pc_BRD4",
             "18h_1pc_BRD4",
             "HeLaiJMJD6_Dox_01O224h_NA")
)
  ]

# accession = "O60885", plot_range = c(531, 581) BRD4 
plot_ptm_stoichiometry(
  MS_SS_KR_PNAS_BRD4_O2pc_dt,
  accession = "O60885",
  plot_range = c(531, 581), 
  all.protein.bs, 
  sample_colors = NA,
)

MS_SS_KR_PNAS_BRD4_O2pc <- MS_SS_KR_PNAS_BRD4_O2pc_dt[is_diagnostic_peak == TRUE & aa == "K"]

# Write to csv table to identify "K" with DI 
fwrite(MS_SS_KR_PNAS_BRD4_O2pc, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_KR_PNAS_BRD4_O2pc.csv"))

```

# 2.8.6 Plotting Stoichiometry values of hypoxia and normoxia data with re-expression of JMJD6 (+ diagnostic ion)

```{r merge_MS_SS_KR_PNAS_data}

# subset rows with diagnostic ion from MS_SS and MS_KR data
DI_sites <- rbind(
  MS_KR1_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  MS_SS_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)],
  pnas2022_BRD_dt[is_diagnostic_peak == TRUE, .(protein_accession, aa_pos)]
)

# remove duplicate sites 
DI_sites <- DI_sites[!duplicated(paste(protein_accession, aa_pos))]

# Merge MS_SS_KR data to only have unique diagnostic ion sites
MS_SS_KR_PNAS_dt <- merge(
  DI_sites,
  MS_SS_KR_PNAS_dt[aa == "K"],
  by = c("protein_accession", "aa_pos")
)

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_KR_PNAS_dt <- MS_SS_KR_PNAS_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# Using contrast function on MS_SS_KR raw stoic data
c_MS_SS_KR_PNAS_dt <- contrast_hydroxylation_by_sample_group(MS_SS_KR_PNAS_dt)

# Factor the gene names
c_MS_SS_KR_PNAS_dt[, `:=`(
  gene_name =  
    factor(gene_name)
)]

# WT values subset to greater than 0 
c_MS_SS_KR_PNAS_dt <- c_MS_SS_KR_PNAS_dt[WT_Inf_21pc_KR > 0]

```


```{r overall_MS_SS_KR_PNAS_comparison}

# Change wide format to long format 
lc_MS_SS_KR_PNAS_dt <- melt(
  c_MS_SS_KR_PNAS_dt,
  id.vars = c("protein_accession", "gene_name", "aa_pos"),
  value.name = "Stoichiometry", 
  variable.name = "sample_group"
  #value.var = grep("_", colnames(c_MS_SS_KR_all_dt), value = TRUE)
)

# Separate the contents of sample_group column into new columns 
lc_MS_SS_KR_PNAS_dt[, `:=`(
  cell = str_split_fixed(sample_group, "_", 4)[, 1],
  induction = str_split_fixed(sample_group, "_", 4)[, 2] %>%
    factor(levels = c("Inf", "0h", "4h", "8h", "18h", "24h")),
  oxygen = str_split_fixed(sample_group, "_", 4)[, 3],
  dataset = str_split_fixed(sample_group, "_", 4)[, 4]
)]

# subset rows that do not contain NA values 
lc_MS_SS_KR_PNAS_dt <- lc_MS_SS_KR_PNAS_dt[!is.na(Stoichiometry)]

# counting the data per sample and size 
lc_MS_SS_KR_PNAS_dt[, data_size_per_sample := .N, by = list(sample_group)]
lc_MS_SS_KR_PNAS_dt[, data_size_per_site := .N, by = list(protein_accession, aa_pos)]

# reorder the oxygen levels
lc_MS_SS_KR_PNAS_dt <-  lc_MS_SS_KR_PNAS_dt[, oxygen := factor(
  oxygen,
  levels = c("01pc", "1pc", "4pc", "21pc")
)]


# subset data to WT normoxia (MS_KR_1 sample)
wt_21pc <- lc_MS_SS_KR_PNAS_dt[sample_group == "WT_Inf_21pc_KR", .(protein_accession, aa_pos, Stoichiometry)]
setnames(wt_21pc, old = "Stoichiometry", "WT_stoichiometry")

wt_21pc <- wt_21pc[WT_stoichiometry > 0]

# Merge WT data with lc_MS_SS_KR_PNAS_dt dtata
lc_MS_SS_KR_PNAS_WT_dt <- merge(
  lc_MS_SS_KR_PNAS_dt[
    cell %in% c("iJ6", "WT")
  ],
  wt_21pc,
  by = c("protein_accession", "aa_pos")
)

```

# 2.8.7 XIC values against MS_SS, MS_KR_1 and PNAS stoichiometry data. 

```{r XIC_vs_stoic}

#-------------
# XIC vs stoic
#-------------

# Load XIC data
xic_MS_SS <- fread(file.path(data.dir, "xic_MS_SS.csv"))

# Merge XIC data and stoichiometry data
xic_stoic <- merge(lc_MS_SS_KR_PNAS_dt, 
                   xic_MS_SS,
                   by = c("induction", "oxygen", "gene_name", "aa_pos"))

# Scatter plot - XIC vs stoichiometry
ggplot(
  xic_stoic,
  aes(
    x = XIC,
    y = Stoichiometry
  )
) + 
  geom_point() +
  geom_smooth(method=lm , color="red", se=TRUE) +
theme_classic_2()

```


```{r Dox_plus_minus_JMJD6_O2_levels}

# Boxplot - Stoichiometry of Dox + and - JMJD6 under O2 levels
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[induction != "Inf" & data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#---------------------------------------------
# Normoxia samples under varying dox induction
#--------------------------------------------- 

# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[oxygen == "21pc" & induction != "Inf" & data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


#------------------------
# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2 (N >= 0.1)
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[WT_stoichiometry >= 0.1 & oxygen == "21pc" & induction != "Inf" & data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  ylab("(N >= 0.1) Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r Dox_plus_minus_JMJD6_O2_levels_relative_WT}

# Boxplot - Stoichiometry of Dox + and - JMJD6 under O2 pc levels
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[induction != "Inf"& data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen) +
  labs(title = "Stoichiometry of Dox +/- JMJD6 at O2pc levels") + 
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#---------------------------------------------
# Normoxia samples under varying dox induction
#---------------------------------------------

# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[oxygen == "21pc" & induction != "Inf"& data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  labs(title = "Stoic of Dox +/- JMJD6 at 21pc O2") + 
  # facet_grid(~ oxygen) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#------------------------
# Boxplot - Stoichiometry of Dox + and - JMJD6 at 21pc O2 (N >= 0.1)
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[WT_stoichiometry >= 0.1 & oxygen == "21pc" & induction != "Inf"& data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site])],
  aes(
    x = induction,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  #geom_hline(yintercept = 1, color = "gray60") +
  geom_boxplot(outlier.shape = NA) +
  # facet_grid(~ oxygen) +
  labs(title = "Stoic of Dox +/- JMJD6 at 21pc O2 (N >= 0.1)") + 
  ylab("(N >= 0.1) Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```

# 2.8.8 Oxygen sensitivity - Comparing hypoxia stoichiometry under varying dox incubation (+ diagnostic ion)

```{r normoxia_vs_varying_hypoxia_pc_overnightdox}

# Boxplot - Comparing the stoic between normoxia and varying pc of hypoxia
# Data subset to overnight (18/24h) dox induction

lc_MS_SS_KR_PNAS_Hxpo_dt <- lc_MS_SS_KR_PNAS_WT_dt[
  grepl("18h|24h|Inf", induction) &
    !sample_group %in% c("iJ6_18h_21pc_SS", "iJ6_Inf_21pc_PNAS", "WT_Inf_21pc_PNAS")
]

#test <-  lc_MS_SS_KR_PNAS_WT_dt[grepl("Inf", induction)]

# reorder the sample_group according to oxygen levels
lc_MS_SS_KR_PNAS_Hxpo_dt[, sample_group :=
                       factor(sample_group,
                              levels = c("WT_Inf_21pc_KR",
                                         "iJ6_24h_21pc_KR",
                                         "iJ6_18h_4pc_SS",
                                         "iJ6_18h_1pc_SS",
                                         "iJ6_24h_01pc_KR"))
]

# Plot
ggplot(
  data = lc_MS_SS_KR_PNAS_Hxpo_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_Hxpo_dt
                                                            [, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Plot - relative to WT_stoichiometry
ggplot(
  data = lc_MS_SS_KR_PNAS_Hxpo_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_Hxpo_dt
                                                            [, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Plot - Subset WT data - N >= 0.1
ggplot(
  data = lc_MS_SS_KR_PNAS_Hxpo_dt[WT_stoichiometry >= 0.1 & data_size_per_site == max(lc_MS_SS_KR_PNAS_Hxpo_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("(N >= 0.1) Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

# Plot - Subset WT data - N >= 0.1, relative to WT_stoichiometry
ggplot(
  data = lc_MS_SS_KR_PNAS_Hxpo_dt[WT_stoichiometry >= 0.1 & data_size_per_site == max(lc_MS_SS_KR_PNAS_Hxpo_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("(N >= 0.1) Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```

# 2.8.9 Oxygen sensitivity - Comparing hypoxia stoichiometry under varying dox incubation - Re-exp J6 (+ diagnostic ion)

```{r reexp_J6_normoxia_vs_varying_hypoxia_pc_overnightdox}

#--------------------------------------
# Only JMJD6 re-expression data plotted
#--------------------------------------

lc_Hxpo_reexp_J6_dt <- lc_MS_SS_KR_PNAS_WT_dt[
  grepl("18h|24h", induction) &
    !sample_group %in% c("iJ6_18h_21pc_SS", "iJ6_Inf_21pc_PNAS")
]

# reorder the sample_group according to oxygen levels
lc_Hxpo_reexp_J6_dt[, sample_group :=
                       factor(sample_group,
                              levels = c("iJ6_24h_21pc_KR",
                                         "iJ6_18h_4pc_SS",
                                         "iJ6_18h_1pc_SS",
                                         "iJ6_24h_01pc_KR"))
]

# Plot
ggplot(
  data = lc_Hxpo_reexp_J6_dt[data_size_per_site == max(lc_Hxpo_reexp_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Plot- relative to WT_stoichiometry
ggplot(
  data = lc_Hxpo_reexp_J6_dt[data_size_per_site == max(lc_Hxpo_reexp_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Subset WT data - N >= 0.1 
# Plot
ggplot(
  data = lc_Hxpo_reexp_J6_dt[WT_stoichiometry >= 0.1 & data_size_per_site == max(lc_Hxpo_reexp_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("(N >= 0.1) Stoichiometry [%]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Plot- N >= 0.1 relative to WT_stoichiometry
ggplot(
  data = lc_Hxpo_reexp_J6_dt[WT_stoichiometry >= 0.1 & data_size_per_site == max(lc_Hxpo_reexp_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("(N >= 0.1) Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```


```{r reexp_J6_varying_hypoxia_pc_overnightdox}

lc_Hxpo_J6_dt <- lc_MS_SS_KR_PNAS_WT_dt[
  grepl("18h|24h", induction) &
    !sample_group %in% c("iJ6_18h_21pc_SS", "iJ6_Inf_21pc_PNAS", "iJ6_24h_21pc_KR")
]

# reorder the sample_group according to oxygen levels
lc_Hxpo_J6_dt[, sample_group :=
                       factor(sample_group,
                              levels = c("iJ6_18h_4pc_SS",
                                         "iJ6_18h_1pc_SS",
                                         "iJ6_24h_01pc_KR"))
]


# Plot- relative to WT_stoichiometry
ggplot(
  data = lc_Hxpo_J6_dt[data_size_per_site == max(lc_Hxpo_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))


# Plot- Subset WT to >= 0.1
ggplot(
  data = lc_Hxpo_J6_dt[WT_stoichiometry >= 0.1 & data_size_per_site == max(lc_Hxpo_J6_dt[, data_size_per_site])],
  aes(
    x = sample_group,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  ylab("(N >= 0.1) Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  coord_cartesian(ylim = c(0, 1.2)) +
  theme(aspect.ratio = 2) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```

# 2.8.10 Oxygen sensitivity - Comparing hypoxia stoichiometry vs normoxia stoichiometry (+ diagnostic ion)

```{r normoxia_vs_hypoxia_scatterplot}

# Scatter plot - Normoxia stoichiometry versus hypoxia at 4% - 18h dox incubation
ggplot(
  data = c_MS_SS_KR_PNAS_dt, 
  aes(
    x = iJ6_24h_21pc_KR,
    y = iJ6_18h_4pc_SS
  )
) +
  geom_point(aes(colour = gene_name)) +
  geom_smooth(method=lm , color="black", size = 0.5, se=TRUE)+
  #facet_grid(~ gene_name, space = "free")+
  theme_classic_2() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR], c_MS_SS_KR_PNAS_dt[, iJ6_18h_4pc_SS], method = "spearma")

wilcox.test(
  c_MS_SS_KR_PNAS_dt[, iJ6_18h_4pc_SS],
  c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR],
  paired = TRUE
)

# Scatter plot - Normoxia stoichiometry versus hypoxia at 1% - 18h dox incubation
ggplot(
  data = c_MS_SS_KR_PNAS_dt, 
  aes(
    x = iJ6_24h_21pc_KR,
    y = iJ6_18h_1pc_SS
  )
) +
  geom_point(aes(colour = gene_name)) +
  geom_smooth(method=lm , color="black", size = 0.5, se=TRUE)+
  #  facet_grid(~ gene_name, space = "free")+
  theme_classic_2() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") 

# correlation test 
cor.test(c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR], c_MS_SS_KR_PNAS_dt[, iJ6_18h_1pc_SS], method = "spearma")

wilcox.test(
  c_MS_SS_KR_PNAS_dt[, iJ6_18h_1pc_SS],
  c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR],
  paired = TRUE
)

# Scatter plot - Normoxia stoichiometry versus hypoxia at 0.1% - 24h dox incubation
ggplot(
  data = c_MS_SS_KR_PNAS_dt, 
  aes(
    x = iJ6_24h_21pc_KR,
    y = iJ6_24h_01pc_KR
  )
) +
  geom_point(aes(colour = gene_name)) +
    geom_smooth(method=lm , color="black", size = 0.5, se=TRUE)+
  #facet_grid(~ gene_name, space = "free")+
  theme_classic_2() +
  coord_cartesian(ylim = c(0, 1)) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right")

# correlation test 
cor.test(c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR], c_MS_SS_KR_PNAS_dt[, iJ6_24h_01pc_KR], method = "spearma")

wilcox.test(
  c_MS_SS_KR_PNAS_dt[, iJ6_24h_01pc_KR],
  c_MS_SS_KR_PNAS_dt[, iJ6_24h_21pc_KR],
  paired = TRUE
)

```

# 2.8.11 Comparing hypoxia stoichiometry under varying dox incubation (+ diagnostic ion)

```{r hypoxia_vs_dox_incubation}

# box_plot - hypoxia stoichiometry in diff dox incubation
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site]) & induction != "Inf" & oxygen != "21pc"], 
  aes(
    x = induction,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen)+
  ggtitle("Hypoxia samples in diff dox incubation") +
  ylab("Stoichiometry [% to WT]") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

#--------------------
# box_plot - Hypoxia samples in diff dox incubation (N =>0.1)
ggplot(
  data = lc_MS_SS_KR_PNAS_WT_dt[data_size_per_site == max(lc_MS_SS_KR_PNAS_WT_dt[, data_size_per_site]) & WT_stoichiometry >= 0.1 & induction != "Inf" & oxygen != "21pc"], 
  aes(
    x = induction,
    y = Stoichiometry/WT_stoichiometry
  )
) +
  geom_boxplot(outlier.shape = NA) +
  facet_grid(~ oxygen)+
  ggtitle("Hypoxia samples in diff dox incubation (N subset)") +
  ylab("Stoichiometry [%] (N =>0.1)") +
  theme_classic_2()+
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  theme(legend.position="right") +
  theme(aspect.ratio = 3) +
  scale_y_continuous(labels = scales::percent_format(accuracy = 1))

```

# Session information

```{r session_Info}

sessioninfo::session_info()

```
