---
title: "2-8. Lysine hydroxylations in varying O2 pc and dox induction time - data visualisation"
author: "Yoichiro Sugimoto and Pallavi Kesavan"
date: "`r format(Sys.time(), '%d %B, %Y')`"
vignette: >
  %\VignetteIndexEntry{Bioconductor style for PDF documents}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
output:
   github_document:
     toc: yes
     toc_depth: 2
---

# Overview

This script examines the oxygen sensitivity of lysine hydroxylations in BRD proteins and their respective sites. This script is for visualizing the calculated stoichiometry. 

# Environment setup

```{r setup_environment}

## Initialize renv (first time only) - re-installed 24.09.2025
# Creates project specific library and renv.lock file. 
# Use 'renv::init(filepath)' to create project library
# renv::init(
#        "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains/R"
#    )

# Define project directory - contains the R scripts, data and result folders
project.dir <-
  file.path(
    "/fast/AG_Sugimoto/home/users/pallavi/projects/20241111_PTMs_in_lysine_rich_domains"
  )

# renv::snapshot(file.path(project.dir, "R"))

#renv::restore completed 24.09.2025
#renv::restore(file.path(project.dir, "R"))

```

# 2.8.1 Install,load essential functions and libraries

```{r load_essential_functions_current_session}

## Load all R scripts from the 'functions' folder into the current session
P2_functions <- sapply(list.files
                       (file.path(project.dir, "R/functions"), 
                         pattern="*.R", 
                         full.names = TRUE), 
                       source)

## Install private package 
# Install ptm.stiochiometry package
# install.packages("/fast/AG_Sugimoto/home/users/pallavi/projects/ptm.stoichiometry", repos = NULL, type = "source")

# Load Libraries
library("readxl")
library("janitor")
library("ptm.stoichiometry")

```

# 2.8.2 Import human protein reference data 

```{r import_data, warning=FALSE}

# Import human protein reference data from specified file path 
ref_protein_dt <- import_reference_fasta(file.path
                                         ("/fast/AG_Sugimoto/reference/uniprot/human", 
                                           "UP000005640_9606.fasta")) 

```

```{r data_directory_MS_SS}

# Define path to data directory
data.dir <- file.path(project.dir, "data")
results.dir <- file.path(project.dir, "results")

#Create p1 results directory
p2_MS_SS_KR <- file.path(results.dir, "p2_MS_SS_KR")
# create.dirs(c(results.dir, p2_MS_SS_KR))

```

# 2.8.12 MS_SS, MS_KR, PNAS Data tables converted to excel 

```{r PNAS_DI_load_data_excel}

# install package for writing data table to excel 
# install.packages('openxlsx')
# library("openxlsx")

#---------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset A)
#---------------------------------------------------------------------

 MQ_DI_dtA_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-A_trp_m7_v7_def_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtA_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#--------------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset B1 and B2)
#--------------------------------------------------------------------------

# Load sample run info data (MQ DI with no waterloss)
MQ_DI_noH2O_run_info <- read_excel(
  file.path(project.dir, "data/analysis_setting/PXD031221_sample_matrix.xlsx"),
  sheet = "MQ_with_DI_noH2O" 
) %>% data.table

# Add a new column 'sample_id' that assigns a unique sequential number to each 
#row (1:N), where N is the total number of rows in the data.table
MQ_DI_noH2O_run_info[, sample_id := 1:.N]

MQ_DI_dtB_stoic_dt <- lapply(
  MQ_DI_noH2O_run_info[data != c("data-A", "data-C"), prefix],
  read_stoic_data,
  pre_prefix = "DI_noH2O_", 
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O")
) %>% rbindlist

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtB_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

#---------------------------------------------------------------------
# Load PNAS stoichiometry data analysed by MaxQuant with DI (dataset C)
#---------------------------------------------------------------------

 MQ_DI_dtC_stoic_dt <- read_stoic_data(
  prefix = "DI_noH2O_data-C_trp_m7_v7_mCC_",
  pre_prefix = "",
  post_fix = "_DI",
  dir_path = file.path(results.dir, "p2-analysis-setting", "MQ_DI_noH2O"))

# Create a logical column for diagnostic peak. For rows where diagnostic peak is 
# present "+", logical vector is "TRUE", or else "FALSE"
MQ_DI_dtC_stoic_dt[, `:=`(
  is_diagnostic_peak = diagnostic_peak == "+" 
)]

```


```{r PNAS_MQ_DI_data_A}

# Subset the data to contain only Lysine
MQ_DI_dtA_stoic_dt <- MQ_DI_dtA_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtA_stoic_dt <-  MQ_DI_dtA_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtA_stoic_dt,
  append(setdiff(names(MQ_DI_dtA_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtA_stoic_dt,
  append(setdiff(names(MQ_DI_dtA_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MQ_DI_dtA_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset A and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtA_stoic_dt <- MQ_DI_dtA_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtA_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtA_stoic_dt.xlsx"))

```


```{r PNAS_MQ_DI_data_B}

# Subset the data to contain only Lysine
MQ_DI_dtB_stoic_dt <- MQ_DI_dtB_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtB_stoic_dt <-  MQ_DI_dtB_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtB_stoic_dt,
  append(setdiff(names(MQ_DI_dtB_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtB_stoic_dt,
  append(setdiff(names(MQ_DI_dtB_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MQ_DI_dtB_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Total_number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset B1 and B2 and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtB_stoic_dt <- MQ_DI_dtB_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtB_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtB_stoic_dt.xlsx"))

```


```{r PNAS_MQ_DI_data_C}

# Subset the data to contain only Lysine
MQ_DI_dtC_stoic_dt <- MQ_DI_dtC_stoic_dt[aa == "K"]

# removing unused data columns from data table
MQ_DI_dtC_stoic_dt <-  MQ_DI_dtC_stoic_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition"
                                          
) := NULL]

# changing stoichiometry column position 
setcolorder(
  MQ_DI_dtC_stoic_dt,
  append(setdiff(names(MQ_DI_dtC_stoic_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MQ_DI_dtC_stoic_dt,
  append(setdiff(names(MQ_DI_dtC_stoic_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

# changing column names - publication ready
setnames(MQ_DI_dtC_stoic_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))

# Merge MQ_DI stoic dataset C and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MQ_DI_dtC_stoic_dt <- MQ_DI_dtC_stoic_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]
 
 # Write to csv table to identify "K" with DI 
write.xlsx(MQ_DI_dtC_stoic_dt, file.path(results.dir, "p2_MS_SS_KR", "MQ_DI_dtC_stoic_dt.xlsx"))

```


```{r MS_KR_excel}

# Subset the data to contain only Lysine
MS_KR1_stoic_K_dt <- MS_KR1_stoic_dt[aa == "K"]

# removing unused data columns from data table
MS_KR1_stoic_K_dt <-  MS_KR1_stoic_K_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition") := NULL]

# changing stoichiometry column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "stoichiometry"), "stoichiometry", after = 6)
)
 
# changing sample_group column position 
setcolorder(
  MS_KR1_stoic_K_dt,
  append(setdiff(names(MS_KR1_stoic_K_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MS_KR1_stoic_K_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))


# Merge MS_KR_1 and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MS_KR1_stoic_K_dt <- MS_KR1_stoic_K_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to Excel table
write.xlsx(MS_KR1_stoic_K_dt, file.path(results.dir, "p2_MS_SS_KR", "MS_KR1_stoic_K.xlsx"))

```


```{r MS_SS_excel}

# Subset the data to contain only Lysine
MS_SS_stoic_K_dt <- MS_SS_stoic_dt[aa == "K"]

# removing unsed data columns from data table
MS_SS_stoic_K_dt <-  MS_SS_stoic_K_dt[, c("sum_intensity_per_position",
                                          "sum_psm_mapped_per_position",
                                          "sum_the_number_of_peptide",
                                          "localization_prob",
                                          "score_for_localization",
                                          "best_localization_ms_ms_id",
                                          "best_localization_raw_file",
                                          "max_score",
                                          "diagnostic_peak",
                                          "condition") := NULL]

# Remove BRD2/3 samples reporting BRD4 stoichiomtery and vice versa
# These samples are removed as it could be false positives reported by the system 
MS_SS_stoic_K_dt <- MS_SS_stoic_K_dt[
  !((grepl("_BRD23$", sample_name) & gene_name == "BRD4") |
       (grepl("_BRD4$", sample_name) & gene_name %in% c("BRD2", "BRD3")))
]

# changing stoichiometry column position 
setcolorder(
  MS_SS_stoic_K_dt,
  append(setdiff(names(MS_SS_stoic_K_dt), "stoichiometry"), "stoichiometry", after = 6)
)

# changing sample_group column position 
setcolorder(
  MS_SS_stoic_K_dt,
  append(setdiff(names(MS_SS_stoic_K_dt), "is_diagnostic_peak"), "is_diagnostic_peak", after = 13)
)

 # changing column names - publication ready
setnames(MS_SS_stoic_K_dt, c("sample_name",
                              "protein_accession",
                              "gene_name",
                              "aa",
                              "aa_pos",
                              "ptm", 
                              "stoichiometry",
                              "sum_peak_intensity",
                              "sum_psm_mapped",
                              "the_number_of_peptide",
                              "is_diagnostic_peak"), 
         c("Sample_Name",
           "Protein_accession",
           "Gene_name",
           "Amino_acid",
           "Amino_acid_pos",
           "PTM_type",
           "Stoichiometry",
           "Total_peak_intensity",
           "Total_PSM_mapped",
           "Number_of_peptide",
           "Diagnostic_peak_presence"))


# Merge MS_SS and pnas2022 manually curated sites by 'Protein_accession' and 'Amino_acid_pos'
 MS_SS_stoic_K_dt <- MS_SS_stoic_K_dt[, `:=`( 
  PNAS2022_curated_oxK_site = 
    paste0(Protein_accession, "_", Amino_acid_pos) %in%
    pnas2022_stoic_dt[PNAS2022_curated_oxK_site == TRUE, paste0(Protein_accession, "_", Amino_acid_pos)] 
)]

 # Write to Excel table
write.xlsx(MS_SS_stoic_K_dt, file.path(results.dir, "p2_MS_SS_KR", "MS_SS_stoic_K.xlsx"))

```


# Session information

```{r session_Info}

sessioninfo::session_info()

```
